---
title: '3-state cSTM in R'
subtitle: "With state-residence time dependency"
author: "The DARTH workgroup"
output:
  pdf_document: default
  html_document: default
---

Developed by the Decision Analysis in R for Technologies in Health (DARTH) workgroup:

- Fernando Alarid-Escudero, PhD (1)
- Eva A. Enns, MS, PhD (2)
- M.G. Myriam Hunink, MD, PhD (3,4)
- Hawre J. Jalal, MD, PhD (5)
- Eline Krijkamp, PhD (6)
- Petros Pechlivanoglou, PhD (7,8)
- Alan Yang, MSc (8)

In collaboration of: 		

1. Stanford University, Stanford, CA, USA
2. University of Minnesota School of Public Health, Minneapolis, MN, USA
3. Erasmus MC, Rotterdam, The Netherlands
4. Harvard T.H. Chan School of Public Health, Boston, USA
5. University of Ottawa, Ottawa, ON, Canada
6. Erasmus University, Rotterdam, The Netherlands
7. University of Toronto, Toronto ON, Canada
8. The Hospital for Sick Children, Toronto ON, Canada


Please cite our publications when using this code:
 
- Jalal H, Pechlivanoglou P, Krijkamp E, Alarid-Escudero F, Enns E, Hunink MG. 
An Overview of R in Health Decision Sciences. Med Decis Making. 2017; 37(3): 735-746. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X16686559
 
- Alarid-Escudero, F., Krijkamp, E.M., Pechlivanoglou, P. et al. A Need for Change! 
 A Coding Framework for Improving Transparency in Decision Modeling. 
 PharmacoEconomics 37, 1329–1339 (2019). https://doi.org/10.1007/s40273-019-00837-x

- Alarid-Escudero F, Krijkamp EM, Enns EA, Yang A, Hunink MGM Pechlivanoglou P,
Jalal H. An Introductory Tutorial on Cohort State-Transition Models in R 
Using a Cost-Effectiveness Analysis Example. Medical Decision Making. 2023;43(1):3-20. doi:10.1177/0272989X221103163

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

Change `eval` to `TRUE` if you want to knit this document.

```{r}
rm(list = ls())      # clear memory (removes all the variables from the workspace)
```

# 01 Load packages

```{r}
if (!require('pacman')) install.packages('pacman'); library(pacman) # use this package to conveniently install other packages
# load (install if required) packages from CRAN
p_load("dplyr", "tidyr", "reshape2", "devtools", "scales", "ellipse", "ggplot2", "lazyeval", "igraph", "truncnorm", "ggraph", "reshape2", "knitr", "stringr", "diagram", "dampack")                                               
# load (install if required) packages from GitHub
# install_github("DARTH-git/darthtools", force = TRUE) #Uncomment if there is a newer version
p_load_gh("DARTH-git/darthtools")
```

# 02 Load functions

```{r}
# all functions are in the darthtools package
```

# 03 Model input

```{r}
## General setup
cycle_length    <- 1                             # cycle length equal to one year (use 1/12 for monthly)
n_cycles        <- 60                            # number of cycles
v_names_cycles  <- paste("cycle", 0:n_cycles)    # cycle names
v_names_states  <- c("H", "S", "D")  # state names
n_states        <- length(v_names_states)        # number of health states 

### Discounting factors 
d_c <- 0.03        # annual discount rate for costs 
d_e <- 0.03        # annual discount rate for QALYs

### Strategies 
v_names_str     <- c("Standard of Care",         # store the strategy names
                     "Treatment A", 
                     "Treatment B")  
n_str           <- length(v_names_str)           # number of strategies

## Within-cycle correction (WCC) using Simpson's 1/3 rule 
v_wcc <- gen_wcc(n_cycles = n_cycles,  method = "Simpson1/3")

### Transition probabilities
p_HS_SoC  <- 0.05  # probability of becoming sick when healthy, conditional on surviving, under standard of care
p_HS_trtA <- 0.04  # probability of becoming sick when healthy, conditional on surviving, under treatment A
p_HS_trtB <- 0.02  # probability of becoming sick when healthy, conditional on surviving, under treatment B
p_SD      <- 0.1   # probability of dying when sick
p_HD_min  <- 0.003 # probability of dying when healthy at t = 0
p_HD_max  <- 0.01  # probability of dying when healthy at t = n_cycles
# probabilities of dying when healthy (age-dependent) - this is now a sequence of numbers
v_p_HD    <- seq(p_HD_min, p_HD_max, length.out = n_cycles)   

### Tunnels
n_tunnel_size           <- n_cycles
# Sick state
v_Sick_tunnel           <- paste("S_", seq(1, n_tunnel_size), "Yr", sep = "")
# Create variables for time-dependent model
v_names_states_tunnels  <- c("H", v_Sick_tunnel, "D")   # state names
n_states_tunnels        <- length(v_names_states_tunnels)        # number of states

# Weibull parameters for state-residence-dependent transition probability
l        <- 0.08
g        <- 1.1
# Weibull function
v_surv_S <- pweibull(0:n_tunnel_size, l, g, lower.tail = F) # survival curve for Sick state
v_p_SD   <- trans_prob(v_surv_S)  # probability of dying when sick (time-in-state dependent)

### State rewards
#### Costs   
c_H       <- 400   # cost of one cycle in healthy state
c_S       <- 1000  # cost of one cycle in sick state
c_D       <- 0     # cost of one cycle in dead state
c_trtA    <- 800   # cost of treatment A (per cycle) in healthy state
c_trtB    <- 1500  # cost of treatment B (per cycle) in healthy state
#### Utilities
u_H       <- 1     # utility when healthy 
u_S       <- 0.5   # utility when sick
u_D       <- 0     # utility when dead
d_e       <- 0.03  # discount rate per cycle equal discount of costs and QALYs by 3%
d_c       <- 0.03  # discount rate per cycle equal discount of costs and QAL  

### Discount weight for costs and effects 
v_dwc     <- 1 / ((1 + (d_e * cycle_length)) ^ (0:n_cycles))
v_dwe     <- 1 / ((1 + (d_c * cycle_length)) ^ (0:n_cycles))
```

# 04 Construct state-transition models

```{r}
m_P_diag <- matrix(0, nrow = n_states, ncol = n_states, dimnames = list(v_names_states, v_names_states))
m_P_diag["H", "H" ] = ""
m_P_diag["H", "S" ] = "" 
m_P_diag["H", "D" ] = ""
m_P_diag["S", "S" ] = ""
m_P_diag["S", "D" ] = ""
m_P_diag["D", "D" ] = ""
layout.fig <- c(2, 1)
plotmat(t(m_P_diag), t(layout.fig), self.cex = 0.5, curve = 0, arr.pos = 0.8,  
        latex = T, arr.type = "curved", relsize = 0.85, box.prop = 0.8, 
        cex = 0.8, box.cex = 0.7, lwd = 1)
```

## 04.1 Initial state vector

```{r}
# All starting healthy
v_m_init_tunnels <- c(1, rep(0, n_tunnel_size), 0) 
```

## 04.2 Initialize cohort traces with tunnels

```{r}
### Initialize cohort trace for state-residece dependent cSTM under SoC 
m_M_tunnels_SoC <- matrix(0, 
                          nrow     = (n_cycles + 1), ncol = n_states_tunnels, 
                          dimnames = list(0:n_cycles, v_names_states_tunnels))
# Store the initial state vector in the first row of the cohort trace
m_M_tunnels_SoC[1, ] <- v_m_init_tunnels

### Initialize cohort trace for strategies A and B
# Structure and initial states are the same as for SoC
m_M_tunnels_trtA <- m_M_tunnels_SoC # Strategy A
m_M_tunnels_trtB <- m_M_tunnels_SoC # Strategy B
```

## 04.3 Create transition probability arrays

```{r}
## Create transition probability arrays for strategy SoC 
### Initialize transition probability array for strategy SoC 
# All transitions to a non-death state are assumed to be conditional on survival
a_P_tunnels_SoC <- array(0,  # Create 3-D array
                       dim = c(n_states_tunnels, n_states_tunnels, n_cycles),               
                       dimnames = list(v_names_states_tunnels, v_names_states_tunnels,
                                       v_names_cycles[-length(v_names_cycles)])) # name the dimensions of the array   

### Fill in array
## from Healthy
a_P_tunnels_SoC["H", "H", ]  <- (1 - v_p_HD) * (1 - p_HS_SoC)
a_P_tunnels_SoC["H", "S_1Yr", ] <- (1 - v_p_HD) *      p_HS_SoC
a_P_tunnels_SoC["H", "D", ]     <-      v_p_HD
## from Sick
for(i in 1:(n_tunnel_size - 1)){ 
  a_P_tunnels_SoC[v_Sick_tunnel[i], v_Sick_tunnel[i + 1], ] <- 1 - v_p_SD[i]
  a_P_tunnels_SoC[v_Sick_tunnel[i], "D", ] <- v_p_SD[i]
}
a_P_tunnels_SoC[v_Sick_tunnel[n_tunnel_size], v_Sick_tunnel[n_tunnel_size], ] <- 1 - v_p_SD[n_tunnel_size]
a_P_tunnels_SoC[v_Sick_tunnel[n_tunnel_size], "D", ] <- v_p_SD[n_tunnel_size]
# from Dead
a_P_tunnels_SoC["D", "D", ] <- 1

## Treatment A
a_P_tunnels_trtA <- a_P_tunnels_SoC
a_P_tunnels_trtA["H", "H", ]     <- (1 - v_p_HD) * (1 - p_HS_trtA)
a_P_tunnels_trtA["H", "S_1Yr", ] <- (1 - v_p_HD) *      p_HS_trtA

## Treatment B
a_P_tunnels_trtB <- a_P_tunnels_SoC
a_P_tunnels_trtB["H", "H", ]     <- (1 - v_p_HD) * (1 - p_HS_trtB)
a_P_tunnels_trtB["H", "S_1Yr", ] <- (1 - v_p_HD) *      p_HS_trtB

# Check if transition array and probabilities are valid
# Check that transition probabilities are in [0, 1]
check_transition_probability(a_P_tunnels_SoC,  verbose = TRUE)
check_transition_probability(a_P_tunnels_trtA, verbose = TRUE)
check_transition_probability(a_P_tunnels_trtB, verbose = TRUE)
# Check that all rows sum to 1
check_sum_of_transition_array(a_P_tunnels_SoC,  n_states = n_states_tunnels, n_cycles = n_cycles, verbose = TRUE)
check_sum_of_transition_array(a_P_tunnels_trtA, n_states = n_states_tunnels, n_cycles = n_cycles, verbose = TRUE)
check_sum_of_transition_array(a_P_tunnels_trtB, n_states = n_states_tunnels, n_cycles = n_cycles, verbose = TRUE)
```

## 04.4 Create transition dynamics arrays

```{r}
# These arrays will capture transitions from each state to another over time 
### Initialize transition dynamics array for strategy SoC 
a_A_tunnels_SoC <- array(0,
                     dim = c(n_states_tunnels, n_states_tunnels, n_cycles + 1),
                     dimnames = list(v_names_states_tunnels, v_names_states_tunnels, 0:n_cycles))
# Set first slice of A with the initial state vector in its diagonal
diag(a_A_tunnels_SoC[, , 1]) <- v_m_init_tunnels
### Initialize transition-dynamics array for strategies A and B
# Structure and initial states are the same as for SoC
a_A_tunnels_trtA <- a_A_tunnels_SoC # Strategy A
a_A_tunnels_trtB <- a_A_tunnels_SoC # Strategy B
```

# 05 Run Markov model

```{r}
# Iterative solution of state-residence dependent cSTM
for (t in 1:n_cycles){  
  ## Fill in cohort trace
  # For SoC
  m_M_tunnels_SoC [t + 1, ] <- m_M_tunnels_SoC [t, ] %*% a_P_tunnels_SoC [, , t] 
  # For Strategy A
  m_M_tunnels_trtA[t + 1, ] <- m_M_tunnels_trtA[t, ] %*% a_P_tunnels_trtA[, , t] 
  # For Strategy B
  m_M_tunnels_trtB[t + 1, ] <- m_M_tunnels_trtB[t, ] %*% a_P_tunnels_trtB[, , t] 
  
  ## Fill in transition-dynamics array
  # For SoC
  a_A_tunnels_SoC[, , t + 1]  <- diag(m_M_tunnels_SoC[t, ])  %*% a_P_tunnels_SoC[, , t]
  # For strategy A
  a_A_tunnels_trtA[, , t + 1] <- diag(m_M_tunnels_trtA[t, ]) %*% a_P_tunnels_trtA[, , t]
  # For strategy B
  a_A_tunnels_trtB[, , t + 1] <- diag(m_M_tunnels_trtB[t, ]) %*% a_P_tunnels_trtB[, , t]
}

# Create aggregated trace
m_M_tunnels_SoC_sum  <- cbind(Healthy = m_M_tunnels_SoC[, "H"], 
                              Sick    = rowSums(m_M_tunnels_SoC[, 2:(n_tunnel_size + 1)]), 
                              Dead    = m_M_tunnels_SoC[, "D"])
m_M_tunnels_trtA_sum <- cbind(Healthy = m_M_tunnels_trtA[, "H"], 
                              Sick    = rowSums(m_M_tunnels_trtA[, 2:(n_tunnel_size + 1)]), 
                              Dead    = m_M_tunnels_trtA[, "D"])
m_M_tunnels_trtB_sum <- cbind(Healthy = m_M_tunnels_trtB[, "H"], 
                              Sick    = rowSums(m_M_tunnels_trtB[, 2:(n_tunnel_size + 1)]), 
                              Dead    = m_M_tunnels_trtB[, "D"])

## Store the cohort traces in a list 
l_m_M <- list(m_M_tunnels_SoC_sum,      
              m_M_tunnels_trtA_sum,
              m_M_tunnels_trtB_sum)
names(l_m_M) <- v_names_str

## Store the transition dynamics array for each strategy in a list 
l_a_A <- list(a_A_tunnels_SoC,
              a_A_tunnels_trtA,
              a_A_tunnels_trtB)
names(l_a_A) <- v_names_str
```

# 06 Plot Outputs

## 06.1 Plot the cohort trace for strategy SoC 

```{r}
## Plot the cohort trace for strategy SoC 
plot_trace(m_M_tunnels_SoC_sum)
```

## 06.2 Overall Survival (OS)

```{r}
v_os <- 1 - m_M_tunnels_SoC_sum[, "D"]    # calculate the overall survival (OS) probability
v_os <- rowSums(m_M_tunnels_SoC_sum[, 1:2])  # alternative way of calculating the OS probability   

# create a simple plot showing the OS
plot(v_os, type = 'l', 
     ylim = c(0, 1),
     ylab = "Survival probability",
     xlab = "Cycle",
     main = "Overall Survival")             
# add grid 
grid(nx = n_cycles, ny = 10, col = "lightgray", lty = "dotted", lwd = par("lwd"), equilogs = TRUE) 
```

## 06.2.1 Life Expectancy (LE)

```{r}
v_le <- sum(v_os) # summing probablity of OS over time  (i.e. life expectancy)
```

## 06.3 Disease prevalence

```{r}
v_prev <- m_M_tunnels_SoC_sum[, "S"]/v_os
plot(v_prev,
     ylim = c(0, 1),
     ylab = "Prevalence",
     xlab = "Cycle",
     main = "Disease prevalence")
```

# 07 State Rewards 

```{r}
## Scale by the cycle length 
# Vector of utilities per cycle for Sick under strategy SoC
v_u_S_SoC        <- rep(u_S, n_tunnel_size)
names(v_u_S_SoC) <- v_Sick_tunnel
# Vector of utilities under strategy SoC
v_u_SoC <- c(H  = u_H, 
             S  = v_u_S_SoC, 
             D  = u_D) * cycle_length

# Vector of costs per cycle for Sick under strategy SoC
v_c_S_SoC        <- rep(c_S, n_tunnel_size)
names(v_c_S_SoC) <- v_Sick_tunnel
# Vector of costs per cycle under strategy SoC
v_c_SoC <- c(H  = c_H,
             S  = v_c_S_SoC, 
             D  = c_D) * cycle_length

# Vector of utilities per cycle for Sick under strategy A
v_u_S_trtA        <- rep(u_S, n_tunnel_size)
names(v_u_S_trtA) <- v_Sick_tunnel
# Vector of  utilities under strategy A
v_u_trtA <- c(H  = u_H, 
              S  = v_u_S_trtA, 
              D  = u_D) * cycle_length
# Vector of costs per cycle for Sick under strategy A
v_c_S_trtA        <- rep(c_S, n_tunnel_size)
names(v_c_S_trtA) <- v_Sick_tunnel
# Vector of costs per cycle under strategy A
v_c_trtA <- c(H  = c_H + c_trtA, 
              S  = v_c_S_trtA, 
              D  = c_D) * cycle_length

# Vector of utilities per cycle for Sick under strategy B
v_u_S_trtB        <- rep(u_S, n_tunnel_size)
names(v_u_S_trtB) <- v_Sick_tunnel
# Vector of  utilities under strategy B
v_u_trtB <- c(H  = u_H, 
              S  = v_u_S_trtB, 
              D  = u_D) * cycle_length
# Vector of costs per cycle for Sick under strategy B
v_c_S_trtB        <- rep(c_S + c_trtB, n_tunnel_size)
names(v_c_S_trtB) <- v_Sick_tunnel
# Vector of costs per cycle under strategy B
v_c_trtB <- c(H  = c_H + c_trtB, 
              S  = v_c_S_trtB, 
              D  = c_D) * cycle_length

## Store state rewards 
# Store the vectors of state utilities for each strategy in a list 
l_u   <- list(SoC = v_u_SoC,
              A   = v_u_trtA,
              B   = v_u_trtB)
# Store the vectors of state cost for each strategy in a list 
l_c   <- list(SoC = v_c_SoC,
              A   = v_c_trtA,
              B   = v_c_trtB)

# assign strategy names to matching items in the lists
names(l_u) <- names(l_c) <- v_names_str
```

# 08 Compute expected outcomes 

```{r}
# Create empty vectors to store total utilities and costs 
v_tot_qaly <- v_tot_cost <- vector(mode = "numeric", length = n_str)
names(v_tot_qaly) <- names(v_tot_cost) <- v_names_str

## Loop through each strategy and calculate total utilities and costs 
for (i in 1:n_str) { # i <- 1
  v_u_str         <- l_u[[i]]   # select the vector of state utilities for the i-th strategy
  v_c_str         <- l_c[[i]]   # select the vector of state costs for the i-th strategy
  a_A_tunnels_str <- l_a_A[[i]] # select the transition array for the i-th strategy
  
  ## Array of state rewards 
  # Create transition matrices of state utilities and state costs for the i-th strategy
  m_u_str   <- matrix(v_u_str, nrow = n_states_tunnels, ncol = n_states_tunnels, byrow = T)
  m_c_str   <- matrix(v_c_str, nrow = n_states_tunnels, ncol = n_states_tunnels, byrow = T)
  # Expand the transition matrix of state utilities across cycles to form a transition array of state utilities
  a_R_u_str <- array(m_u_str, 
                     dim      = c(n_states_tunnels, n_states_tunnels, n_cycles + 1),
                     dimnames = list(v_names_states_tunnels, v_names_states_tunnels, 0:n_cycles))
  # Expand the transition matrix of state costs across cycles to form a transition array of state costs
  a_R_c_str <- array(m_c_str, 
                     dim      = c(n_states_tunnels, n_states_tunnels, n_cycles + 1),
                     dimnames = list(v_names_states_tunnels, v_names_states_tunnels, 0:n_cycles))
  
  ### Expected QALYs and costs for all transitions per cycle
  # QALYs = life years x QoL
  # Note: all parameters are annual in our example. In case your own case example is different make sure you correctly apply.
  a_Y_c_str <- a_A_tunnels_str * a_R_c_str
  a_Y_u_str <- a_A_tunnels_str * a_R_u_str 
  
  ### Expected QALYs and costs per cycle
  ## Vector of QALYs and costs
  v_qaly_str <- apply(a_Y_u_str, 3, sum) # sum the proportion of the cohort across transitions 
  v_cost_str <- apply(a_Y_c_str, 3, sum) # sum the proportion of the cohort across transitions
  
  ## Discounted total expected QALYs and Costs per strategy and apply within-cycle correction if applicable
  # QALYs
  v_tot_qaly[i] <- t(v_qaly_str) %*% (v_dwe * v_wcc)
  # Costs
  v_tot_cost[i] <- t(v_cost_str) %*% (v_dwc * v_wcc)
}
```

# 09 Cost-effectiveness analysis (CEA) 

```{r}
## Incremental cost-effectiveness ratios (ICERs) 
df_cea <- calculate_icers(cost       = v_tot_cost, 
                          effect     = v_tot_qaly,
                          strategies = v_names_str)
df_cea
```

```{r}
## CEA table in proper format 
table_cea <- format_table_cea(df_cea) 
table_cea
```

```{r}
## CEA frontier 
plot_icers(df_cea, label = "all", txtsize = 16) +
  expand_limits(x = max(table_cea$QALYs) + 0.1) +
  theme(legend.position = c(0.8, 0.3))
```
