---
title: 'Calibrating the Sick-Sicker model'
author: "The DARTH workgroup"
output:
  pdf_document: default
  html_document: default
---

Developed by the Decision Analysis in R for Technologies in Health (DARTH) workgroup:

Fernando Alarid-Escudero, PhD (1) 

Eva A. Enns, MS, PhD (2)	

M.G. Myriam Hunink, MD, PhD (3,4)

Hawre J. Jalal, MD, PhD (5) 

Eline M. Krijkamp, MSc (3)	

Petros Pechlivanoglou, PhD (6)

Alan Yang, MSc (7)

In collaboration of: 		

1. Drug Policy Program, Center for Research and Teaching in Economics (CIDE) - CONACyT, 
   Aguascalientes, Mexico
2. University of Minnesota School of Public Health, Minneapolis, MN, USA
3. Erasmus MC, Rotterdam, The Netherlands
4. Harvard T.H. Chan School of Public Health, Boston, USA
5. University of Pittsburgh Graduate School of Public Health, Pittsburgh, PA, USA
6. The Hospital for Sick Children, Toronto and University of Toronto, Toronto ON, Canada
7. The Hospital for Sick Children, Toronto ON, Canada

Please cite our publications when using this code:
 
- Alarid-Escudero F, Maclehose RF, Peralta Y, Kuntz KM, Enns EA. 
Non-identifiability in model calibration and implications for 
medical decision making. Med Decis Making. 2018; 38(7):810-821. 
 
- Jalal H, Pechlivanoglou P, Krijkamp E, Alarid-Escudero F, Enns E, Hunink MG. 
An Overview of R in Health Decision Sciences. Med Decis Making. 2017; 37(3): 735-746. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X16686559
 
A walkthrough of the code could be found in the follwing link:
- https://darth-git.github.io/calibSMDM2018-materials/

Copyright 2017, THE HOSPITAL FOR SICK CHILDREN AND THE COLLABORATING INSTITUTIONS. 
All rights reserved in Canada, the United States and worldwide. Copyright, 
trademarks, trade names and any and all associated intellectual property are 
exclusively owned by THE HOSPITAL FOR Sick CHILDREN and the collaborating 
institutions. These materials may be used, reproduced, modified, distributed 
and adapted with proper attribution.

\newpage

```{r}
rm(list = ls())      # clear memory (removes all the variables from the workspace)
```

# 00  Calibration Specifications 

Model: Sick-Sicker 4-state Markov Model

Inputs to be calibrated: p_S1S2, hr_S1, hr_S2

Targets: Surv, Prev, PropSick

Calibration method: 

Goodness-of-fit measure: Sum of Log-Likelihood

# 01 Load packages

```{r, warning=F, message=F}
# package used to call files from different folders within the project folder
if (!require('here')) install.packages('here'); library(here)
# calibration functionality
if (!require('lhs')) install.packages('lhs'); library(lhs)

# visualization
if (!require('plotrix')) install.packages('plotrix'); library(plotrix)
if (!require('psych')) install.packages('psych'); library(psych)
# now that we have three inputs to estimate, we'll need higher dimension visualization
if (!require('scatterplot3d')) install.packages('scatterplot3d'); library(scatterplot3d)
```

# 02 Load target data

```{r}
# your turn

```

# 03 Load model as a function

```{r}
# - inputs are parameters to be estimated through calibration
# - outputs correspond to the target data

# your turn

```

# 04 Specify calibration parameters

```{r}
# your turn

```

# 05 Calibrate!

```{r}
# your turn

```

# 06 Exploring best-fitting input sets

```{r}
# your turn

```


