---
title: 'Simple 3-state Markov model in R'
subtitle: 'with age dependency AND with time dependency'
author: "The DARTH workgroup"
output:
  pdf_document: default
  html_document: default
---

Developed by the Decision Analysis in R for Technologies in Health (DARTH) workgroup:

Fernando Alarid-Escudero, PhD (1) 

Eva A. Enns, MS, PhD (2)	

M.G. Myriam Hunink, MD, PhD (3,4)

Hawre J. Jalal, MD, PhD (5) 

Eline M. Krijkamp, MSc (3)	

Petros Pechlivanoglou, PhD (6,7)

Alan Yang, MSc (7)

In collaboration of: 		

1. Drug Policy Program, Center for Research and Teaching in Economics (CIDE) - CONACyT, 
   Aguascalientes, Mexico
2. University of Minnesota School of Public Health, Minneapolis, MN, USA
3. Erasmus MC, Rotterdam, The Netherlands
4. Harvard T.H. Chan School of Public Health, Boston, USA
5. University of Pittsburgh Graduate School of Public Health, Pittsburgh, PA, USA
6. University of Toronto, Toronto ON, Canada
7. The Hospital for Sick Children, Toronto ON, Canada

Please cite our publications when using this code:
 
- Jalal H, Pechlivanoglou P, Krijkamp E, Alarid-Escudero F, Enns E, Hunink MG. 
An Overview of R in Health Decision Sciences. Med Decis Making. 2017; 37(3): 735-746. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X16686559
 
- Krijkamp EM, Alarid-Escudero F, Enns EA, Jalal HJ, Hunink MGM, Pechlivanoglou P. 
Microsimulation modeling for health decision sciences using R: A tutorial. 
Med Decis Making. 2018;38(3):400–22. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X18754513
 
- Krijkamp EM, Alarid-Escudero F, Enns E, Pechlivanoglou P, Hunink MM, Jalal H. 
A Multidimensional Array Representation of State-Transition Model Dynamics. 
BioRxiv 670612 2019.https://www.biorxiv.org/content/10.1101/670612v1

Copyright 2017, THE HOSPITAL FOR SICK CHILDREN AND THE COLLABORATING INSTITUTIONS. 
All rights reserved in Canada, the United States and worldwide. Copyright, 
trademarks, trade names and any and all associated intellectual property are 
exclusively owned by THE HOSPITAL FOR Sick CHILDREN and the collaborating 
institutions. These materials may be used, reproduced, modified, distributed 
and adapted with proper attribution.

\newpage

```{r}
rm(list = ls())      # clear memory (removes all the variables from the workspace)
```

# 01 Load packages

```{r}
# no packages required
```

# 02 Load functions

```{r}
# no functions required
```

# 03 Input model parameters

```{r}
# Strategy names
v_names_str <- c("Base Case")  

# Number of strategies
n_str <- length(v_names_str)

# Markov model parameters
v_n  <- c("Healthy", "Sick", "Dead")    # state names
n_s  <- length(v_n)                     # number of states
n_t  <- 60                              # number of cycles

# Tunnels
tunnel_size <- n_t
# Sick state
Sick_tunnel <- paste("Sick_", seq(1, tunnel_size), "Yr", sep = "")
# Create variables for time-dependent model
v_n_td  <- c("Healthy", Sick_tunnel, "Dead")   # state names
n_s_td  <- length(v_n_td)                      # number of states

p_HD <- seq(0.003, 0.01, length.out = n_t)  # probability to die when sick (age-dependent) - this is a sequence - officially v_HD
p_HS <- 0.05                                # probability to become sick when healthy
p_SD <- 0.1                                 # probability to die when sick

# Weibull parameters
l <- 0.08
g <- 1.1
p_SD <- l*g*(1:tunnel_size)^{g-1}       # probability to die when sick (time-dependent)

# Costs and utilities  
c_H  <- 400                             # cost of remaining one cycle healthy
c_S  <- 1000                            # cost of remaining one cycle sick
c_D  <- 0                               # cost of remaining one cycle dead
u_H  <- 0.8                             # utility when healthy 
u_S  <- 0.5                             # utility when sick
u_D  <- 0                               # utility when dead
d_e <- d_c <- 0.03                      # equal discount of costs and QALYs by 3%

# calculate discount weights for costs for each cycle based on discount rate d_c
v_dwc <- 1 / (1 + d_e) ^ (0:n_t) 
# calculate discount weights for effectiveness for each cycle based on discount rate d_e
v_dwe <- 1 / (1 + d_c) ^ (0:n_t) 
```

# 04 Define and initialize matrices and vectors

## 04.1 Cohort trace

```{r}
m_M <- matrix(NA, 
              nrow = n_t + 1,  # create Markov trace (n_t + 1 because R doesn't understand 
                               # Cycle 0)
              ncol = n_s_td,                  
              dimnames = list(0:n_t, v_n_td)) 

# The cohort starts as healthy
# initialize first cycle of Markov trace accounting for the tunnels
m_M[1, ] <- c(1, rep(0, tunnel_size), 0)     
```

## 04.2 Transition probability array

```{r}
# create the transition probability array
a_P <- array(0,                                          # Create 3-D array
             dim = c(n_s_td, n_s_td, n_t),               
             dimnames = list(v_n_td, v_n_td, 0:(n_t-1)))  
```

Fill in the transition probability array:

```{r}
# from Healthy
a_P["Healthy", "Healthy", ]  <- 1 - p_HD - p_HS
a_P["Healthy", "Sick_1Yr", ] <- p_HS
a_P["Healthy", "Dead", ]     <- p_HD

# from Sick
for(i in 1:(tunnel_size - 1)){ 
  a_P[Sick_tunnel[i], Sick_tunnel[i + 1], ] <- 1 - p_SD[i]
  a_P[Sick_tunnel[i], "Dead", ] <- p_SD[i]
}

a_P[Sick_tunnel[tunnel_size], Sick_tunnel[tunnel_size], ] <- 1 - p_SD[tunnel_size]
a_P[Sick_tunnel[tunnel_size], "Dead", ] <- p_SD[tunnel_size]

# from Dead
a_P["Dead", "Dead", ] <- 1
```

## 04.3 Check if transition array and probabilities are valid

```{r}
# Check if transition matrix is valid (i.e., each row should add up to 1)
valid <- apply(a_P, 3, function(x) sum(rowSums(x))==n_s_td)
if (!isTRUE(all.equal(as.numeric(sum(valid)), as.numeric(n_t)))) {
  stop("This is not a valid transition Matrix")
}
```

# 05 Run Markov model

```{r}
for (t in 1:n_t) {                         # loop through the number of cycles
  m_M[t + 1, ] <- m_M[t, ] %*% a_P[, , t]  # estimate the Markov trace for cycle t + 1 
                                           # using the t-th matrix from the 
                                           # probability array 
}
head(m_M, n = 30)
```

Create aggregated trace.

```{r}
m_M_td <- cbind(Healthy = m_M[, "Healthy"], 
                Sick = rowSums(m_M[, 2:(tunnel_size + 1)]), 
                Dead = m_M[, "Dead"])
head(m_M_td) # show the first rows of the aggregated Markov trace
```

# 06 Compute and Plot Epidemiological Outcomes

## 06.1 Cohort trace

```{r}
# create a plot of the data
matplot(m_M_td, type = 'l', 
        ylab = "Probability of state occupancy",
        xlab = "Cycle",
        main = "Cohort Trace", lwd = 2)              
# add a legend to the graph
legend("right", v_n, col = c("black", "red", "green"), lty = 1:3, bty = "n")  
```

## 06.2 Overall Survival (OS)

```{r}
v_os <- 1 - m_M_td[, "Dead"]    # calculate the overall survival (OS) probability
v_os <- rowSums(m_M_td[, 1:2])  # alternative way of calculating the OS probability   

# create a simple plot showing the OS
plot(v_os, type = 'l', 
     ylim = c(0, 1),
     ylab = "Survival probability",
     xlab = "Cycle",
     main = "Overall Survival")             
# add grid 
grid(nx = n_t, ny = 10, col = "lightgray", lty = "dotted", lwd = par("lwd"), equilogs = TRUE) 
```

## 06.2.1 Life Expectancy (LE)

```{r}
v_le <- sum(v_os)           # summing probablity of OS over time  (i.e. life expectancy)
```

## 06.3 Disease prevalence

```{r}
v_prev <- m_M_td[, "Sick"]/v_os
plot(v_prev,
     ylim = c(0, 1),
     ylab = "Prevalence",
     xlab = "Cycle",
     main = "Disease prevalence")
```

# 07 Compute Cost-Effectiveness Outcomes

## 07.1 Mean Costs and QALYs

```{r}
# per cycle
# calculate expected costs by multiplying m_M with the cost vector for the different 
# health states   
v_tc <- m_M_td %*% c(c_H, c_S, c_D)  
# calculate expected QALYs by multiplying m_M with the utilities for the different 
# health states   
v_tu <- m_M_td %*% c(u_H, u_S, u_D)  
```

## 07.2 Discounted Mean Costs and QALYs

```{r}
# Discount costs  by multiplying the cost vector with discount weights (v_dw) 
v_tc_d <-  t(v_tc) %*% v_dwc
# Discount QALYS  by multiplying the QALYs vector with discount weights (v_dw)
v_te_d <-  t(v_tu) %*% v_dwe
```

## 07.3 Results

```{r}
results <- data.frame( "Total Discounted Cost" = v_tc_d, 
                       "Life Expectancy" = v_le, 
                       "Total Discounted QALYs" = v_te_d, 
                       check.names = F)
results
```

