---
title: 'Microsimulation Sick-Sicker model with time dependency with PSA'
subtitle: 'Includes individual characteristics: age, age dependent mortality probabilities, individual treatment effect modifyer, time dependency for the sick (S1) state, increasing change of death in the first 6 year of sickness (tunnel)'
author: "The DARTH workgroup"
output:
  pdf_document: default
  html_document: default
---

Developed by the Decision Analysis in R for Technologies in Health (DARTH) workgroup:

Fernando Alarid-Escudero, PhD (1) 

Eva A. Enns, MS, PhD (2)	

M.G. Myriam Hunink, MD, PhD (3,4)

Hawre J. Jalal, MD, PhD (5) 

Eline M. Krijkamp, MSc (3)	

Petros Pechlivanoglou, PhD (6)

Alan Yang, MSc (7)

In collaboration of: 		

1. Drug Policy Program, Center for Research and Teaching in Economics (CIDE) - CONACyT, 
   Aguascalientes, Mexico
2. University of Minnesota School of Public Health, Minneapolis, MN, USA
3. Erasmus MC, Rotterdam, The Netherlands
4. Harvard T.H. Chan School of Public Health, Boston, USA
5. University of Pittsburgh Graduate School of Public Health, Pittsburgh, PA, USA
6. The Hospital for Sick Children, Toronto and University of Toronto, Toronto ON, Canada
7. The Hospital for Sick Children, Toronto ON, Canada

Please cite our publications when using this code:
 
- Jalal H, Pechlivanoglou P, Krijkamp E, Alarid-Escudero F, Enns E, Hunink MG. 
An Overview of R in Health Decision Sciences. Med Decis Making. 2017; 37(3): 735-746. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X16686559
 
- Krijkamp EM, Alarid-Escudero F, Enns EA, Jalal HJ, Hunink MGM, Pechlivanoglou P. 
Microsimulation modeling for health decision sciences using R: A tutorial. 
Med Decis Making. 2018;38(3):400–22. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X18754513
 
- Krijkamp EM, Alarid-Escudero F, Enns E, Pechlivanoglou P, Hunink MM, Jalal H. 
A Multidimensional Array Representation of State-Transition Model Dynamics. 
BioRxiv 670612 2019.https://www.biorxiv.org/content/10.1101/670612v1

Copyright 2017, THE HOSPITAL FOR SICK CHILDREN AND THE COLLABORATING INSTITUTIONS. 
All rights reserved in Canada, the United States and worldwide. Copyright, 
trademarks, trade names and any and all associated intellectual property are 
exclusively owned by THE HOSPITAL FOR Sick CHILDREN and the collaborating 
institutions. These materials may be used, reproduced, modified, distributed 
and adapted with proper attribution.

\newpage

```{r}
rm(list = ls())      # clear memory (removes all the variables from the workspace)
```

# 01 Load packages

```{r, warning=F, message=F}
if (!require('pacman')) install.packages('pacman'); library(pacman) # use this package to conveniently install other packages
# load (install if required) packages from CRAN
p_load("here", "dplyr", "devtools", "scales", "ellipse", "ggplot2", "lazyeval", "igraph", "truncnorm", "ggraph", 
       "reshape2", "knitr")                                               
# load (install if required) packages from GitHub
# install_github("DARTH-git/dampack", force = TRUE) Uncomment if there is a newer version
p_load_gh("DARTH-git/dampack")  
```

# 02 Load functions

```{r}
source(here("functions", "Functions.R"))
```

# 03 Input model parameters

```{r}
set.seed(1)  # set the seed  

# Model structure 
n_t   <- 30                       # time horizon, 30 cycles
n_i   <- 100000                   # number of simulated individuals
v_n   <- c("H", "S1", "S2", "D")  # the model states names
n_s   <- length(v_n)              # the number of health states
d_r   <- 0.03                     # discount rate of 3% per cycle
v_dwe <- v_dwc <- 1 / ((1 + d_r) ^ (0:n_t))    # discount weight 
v_names_str <- c("no treatment", "treatment")  # strategy names
n_str <- length(v_names_str)      # number of strategies

### Event probabilities (per cycle)
# Annual transition probabilities
p_HS1   <- 0.15                   # probability of becoming sick when healthy
p_S1H   <- 0.5                    # probability of recovering to healthy when sick
p_S1S2  <- 0.105                  # probability of becoming sicker when sick

# Annual probabilities of death
# load age dependent probability
p_mort   <- read.csv(here("data", "mortProb_age.csv"))
# load age distribution
dist_Age <- read.csv(here("data", "MyPopulation-AgeDistribution.csv")) 

# probability to die in S1 by cycle 
p_S1D    <- c(0.0149, 0.018, 0.021, 0.026, 0.031, rep(0.037, n_t - 5)) 
p_S2D    <- 0.048           # probability to die in S2

# Cost inputs
c_H     <- 2000             # cost of one cycle in the healthy state
c_S1    <- 4000             # cost of one cycle in the sick state
c_S2    <- 15000            # cost of one cycle in the sicker state
c_D     <- 0                # cost of one cycle in the dead state
c_Trt   <- 12000            # cost of treatment (per cycle)

# Utility inputs
u_H     <- 1                # utility when healthy 
u_S1    <- 0.75             # utility when sick 
u_S2    <- 0.5              # utility when sicker
u_D     <- 0                # utility when dead
u_Trt   <- 0.95             # utility when sick(er) and being treated
```

# 04 Sample individual level characteristics

## 04.1 Static characteristics

```{r}
v_x     <- runif(n_i, min = 0.95, max = 1.05) # treatment effect modifier at baseline 
# sample from age distribution an initial age for every individual
v_age0  <- sample(x = dist_Age$age, prob = dist_Age$prop, size = n_i, replace = TRUE) 
df_X    <- data.frame(ID = 1:n_i, x = v_x, Age = v_age0)
```

## 04.2 Dynamic characteristics 

```{r}
# Specify the initial health state of the individuals 
# everyone begins in the healthy state (in this example)
# a vector with the initial health state for all individuals
v_M_init  <- rep("H", n_i)       
# a vector with the time of being sick at the start of the model  
v_Ts_init <- rep(0, n_i)         
```

# 05 Define Simulation Functions

## 05.1 Probability function

The function that updates the transition probabilities of every cycle is shown below.

```{r}
Probs <- function(M_t, df_X, v_Ts, t) { 
  # Arguments:
    # M_t: health state occupied by individual i at cycle t (character variable)
    # df_X:     data frame with individual characteristics data 
  # v_Ts: time an individual is sick
  # t:     current cycle 
# Returns: 
  #   transition probabilities for that cycle
  
  # create matrix of state transition probabilities  
  m_p_t           <- matrix(0, nrow = n_s, ncol = n_i) 
  rownames(m_p_t) <-  v_n  # give the state names to the rows
  
  # lookup baseline probability and rate of dying based on individual characteristics
  p_HD_all <- inner_join(df_X, p_mort, by = c("Age"))
  p_HD     <- p_HD_all[M_t == "H","p_HD"]
  
  # update the m_p with the appropriate probabilities   
  # transition probabilities when healthy
  m_p_t[, M_t == "H"]  <- rbind(1 - p_HS1 - p_HD, p_HS1, 0, p_HD)                              
  # transition probabilities when sick
  m_p_t[, M_t == "S1"] <- rbind(p_S1H, 1 - p_S1H - p_S1S2 - p_S1D[v_Ts], p_S1S2, p_S1D[v_Ts])  
  # transition probabilities when sicker
  m_p_t[, M_t == "S2"] <- rbind(0, 0, 1 - p_S2D, p_S2D)                                            
  # transition probabilities when dead   
  m_p_t[, M_t == "D"]  <- rbind(0, 0, 0, 1)                                                        
  
  return(t(m_p_t))
}       
```

## 05.2 Cost function

The `Costs` function estimates the costs at every cycle.

```{r}
Costs <- function (M_t, Trt = FALSE) {
  # M_t: health state occupied by individual i at cycle t (character variable)
  # Trt:  is the individual being treated? (default is FALSE) 
  
  c_t <- 0                                 # by default the cost for everyone is zero 
  c_t[M_t == "H"]  <- c_H                  # update the cost if healthy
  c_t[M_t == "S1"] <- c_S1 + c_Trt * Trt   # update the cost if sick conditional on treatment
  c_t[M_t == "S2"] <- c_S2 + c_Trt * Trt   # update the cost if sicker conditional on treatment
  c_t[M_t == "D"]  <- c_D                  # update the cost if dead
  
  return(c_t)        		                   # return the costs
}
```

## 05.3 Health outcome function

The `Effs` function to update the utilities at every cycle.

```{r}
Effs <- function (M_t, df_X, Trt = FALSE, cl = 1) {
  # M_t: health state occupied by individual i at cycle t (character variable)
  # df_X: data frame with individual characteristics data 
  # Trt:  is the individual treated? (default is FALSE) 
  # cl:   cycle length (default is 1)
  
  u_t <- 0                                 # by default the utility for everyone is zero
  u_t[M_t == "H"]  <- u_H                  # update the utility if healthy
  u_t[M_t == "S1" & Trt == FALSE] <- u_S1  # update the utility if sick
  # update the utility if sick but on treatment (adjust for individual effect modifier) 
  u_t[M_t == "S1" & Trt == TRUE]  <- u_Trt * df_X$x[M_t == "S1"]  
  u_t[M_t == "S2"] <- u_S2                 # update the utility if sicker
  u_t[M_t == "D"]  <- u_D                  # update the utility if dead
  
  QALYs <-  u_t * cl  # calculate the QALYs during cycle t
  return(QALYs)       # return the QALYs
}
```

# 06 Run Microsimulation

```{r}
MicroSim <- function(n_i, df_X , Trt = FALSE, seed = 1) {
# Arguments:  
  # n_i:     number of individuals
    # df_X     data frame with individual characteristics data 
    # Trt:     is this the individual receiving treatment? (default is FALSE)
  # seed:    default is 1
  
  set.seed(seed) # set the seed
  
  n_s <- length(v_n) # the number of health states
  
  # create three matrices called m_M, m_C and m_E
  # number of rows is equal to the n_i, the number of columns is equal to n_t  
  # (the initial state and all the n_t cycles)
  # m_M is used to store the health state information over time for every individual
  # m_C is used to store the costs information over time for every individual
  # m_E is used to store the effects information over time for every individual
  
  m_M <- m_C <- m_E <- m_Ts <-  matrix(nrow = n_i, ncol = n_t + 1, 
                                     dimnames = list(paste("ind"  , 1:n_i, sep = " "), 
                                                     paste("cycle", 0:n_t, sep = " ")))  

    m_M [, 1] <- v_M_init    # initial health state at cycle 0 for individual i
    v_Ts      <- v_Ts_init   # initialize time since illnes onset for individual i
    
    # calculate costs per individual during cycle 0
    m_C[, 1]  <- Costs(m_M[, 1], Trt)     
    # calculate QALYs per individual during cycle 0
    m_E[, 1]  <- Effs (m_M[, 1], df_X, Trt)   
    
    # open a loop for time running cycles 1 to n_t 
    for (t in 1:n_t) {
      # calculate the transition probabilities for the cycle based on  health state t
      m_P <- Probs(m_M[, t], df_X, v_Ts, t)             
      # sample the current health state and store that state in matrix m_M 
      m_M[, t + 1]  <- samplev(m_P, 1)                  
      # calculate costs per individual during cycle t + 1
      m_C[, t + 1]  <- Costs(m_M[, t + 1], Trt)         
      # calculate QALYs per individual during cycle t + 1
      m_E[, t + 1]  <- Effs(m_M[, t + 1], df_X, Trt)    
      
      # update time since illness onset for t + 1 
      v_Ts <- if_else(m_M[, t + 1] == "S1", v_Ts + 1, 0) 
      # update the age of individuals that are alive
      df_X$Age[m_M[, t + 1] != "D"]  <- df_X$Age[m_M[, t + 1] != "D"] + 1
    
      # Display simulation progress
      if(t/(n_t/10) == round(t/(n_t/10), 0)) { # display progress every 10%
        cat('\r', paste(t/n_t * 100, "% done", sep = " "))
      }
      
    } # close the loop for the time points 
  
  # calculate  
  tc <- m_C %*% v_dwc    # total (discounted) cost per individual
  te <- m_E %*% v_dwe    # total (discounted) QALYs per individual 
  tc_hat <- mean(tc)     # average (discounted) cost 
  te_hat <- mean(te)     # average (discounted) QALYs
  
  # store the results from the simulation in a list
  results <- list(m_M = m_M, m_C = m_C, m_E = m_E, tc = tc , te = te, tc_hat = tc_hat, 
                  te_hat = te_hat)   
  
  return(results)  # return the results

} # end of the MicroSim function  

# By specifying all the arguments in the `MicroSim()` the simulation can be started
# In this example the outcomes are of the simulation are stored in the variables `outcomes_no_tr` and `outcomes_trt`.

# Run the simulation for both no treatment and treatment options
outcomes_no_trt  <- MicroSim(n_i, df_X, Trt = FALSE, seed = 1)
outcomes_trt     <- MicroSim(n_i, df_X, Trt = TRUE,  seed = 1)
```

# 07 Visualize results

```{r}
options(scipen = 999)

# No treatment
plot(density(outcomes_no_trt$tc), main = paste("Total cost per person"), xlab = "Cost ($)")
plot(density(outcomes_no_trt$te), main = paste("Total QALYs per person"), xlab = "QALYs")
plot_m_TR(outcomes_no_trt$m_M)  # health state trace

# Treatment
plot(density(outcomes_trt$tc), main = paste("Total cost per person"), xlab = "Cost ($)")
plot(density(outcomes_trt$te), main = paste("Total QALYs per person"), xlab = "QALYs")
plot_m_TR(outcomes_trt$m_M)     # health state trace
```

# 08 Cost Effectiveness Analysis

```{r}
# store the mean costs of each strategy in a new variable C (vector of costs)
v_C <- c(outcomes_no_trt$tc_hat, outcomes_trt$tc_hat)
# store the mean QALYs of each strategy in a new variable E (vector of effects)
v_E <- c(outcomes_no_trt$te_hat, outcomes_trt$te_hat)

# use dampack to calculate the ICER
calculate_icers(cost       = v_C,
                effect     = v_E,
                strategies = v_names_str)
```

# 09 Probabilistic Sensitivity Analysis (PSA)

```{r}
# Function that generates random sample for PSA
gen_psa <- function(n_sim = 1000, seed = 071818){
  set.seed(seed) # set a seed to be able to reproduce the same results
  
  df_psa <- data.frame(
    # Transition probabilities (per cycle)
    p_HS1   = rbeta(n_sim, 30, 170),          # probability to become sick when healthy
    p_S1H   = rbeta(n_sim, 60, 60) ,          # probability to become healthy when sick
    p_S1S2  = rbeta(n_sim, 84, 716),          # probability to become sicker when sick
    p_S2D   = rbeta(n_sim, 22, 434),          # probability to die in S2
    
    # Cost vectors with length n_sim
    # cost of remaining one cycle in state H
    c_H     = rgamma(n_sim, shape = 100, scale = 20)    ,        
    # cost of remaining one cycle in state S1
    c_S1    = rgamma(n_sim, shape = 177.8, scale = 22.5),        
    # cost of remaining one cycle in state S2
    c_S2    = rgamma(n_sim, shape = 225, scale = 66.7)  ,        
    # cost of treatment (per cycle)
    c_trt   = rgamma(n_sim, shape = 73.5, scale = 163.3),        
    # cost of being in the death state
    c_D     = 0                                         ,        
    
    # Utility vectors with length n_sim 
    u_H     = rbeta(n_sim, 9, 0.009),         # utility when healthy
    u_S1    = rbeta(n_sim ,351, 117),         # utility when sick
    u_S2    = rbeta(n_sim, 138, 138),         # utility when sicker
    u_D     = 0,                              # utility when dead
    u_trt   = rbeta(n_sim, 112, 6),           # utility when being treated
    lb_eff  = 0.95,                           # lower bound of effect modifier
    ub_eff  = 1.05                            # upper bound of effect modifier
    )

  return(df_psa)
}
# Try it
gen_psa(10) 

# Decrease number of individuals since PSA takes a lot of time
n_i <- 1000

# update Sample individual level characteristics 

# Dynamic characteristics 
# Specify the initial health state of the individuals 
# everyone begins in the healthy state (in this example)
v_M_init  <- rep("H", n_i)    # a vector with the initial health state for all individuals 
v_Ts_init <- rep(0, n_i)  # a vector with the time of being sick at the start of the model 

# Number of simulations
n_sim <- 500

# Generate PSA input dataset
df_psa_input <- gen_psa(n_sim = n_sim)
# First six observations
head(df_psa_input)

## Histogram of parameters
# Make sure the Plots window is large enough to plot all the histograms
ggplot(melt(df_psa_input, variable.name = "Parameter"), aes(x = value)) +
  facet_wrap(~Parameter, scales = "free") +
  geom_histogram(aes(y = ..density..)) +
  theme_bw(base_size = 16)

# Initialize dataframes with PSA output 
# Dataframe of costs
df_c <- as.data.frame(matrix(0, 
                            nrow = n_sim,
                            ncol = n_str))
colnames(df_c) <- v_names_str
# Dataframe of effectiveness
df_e <- as.data.frame(matrix(0, 
                            nrow = n_sim,
                            ncol = n_str))
colnames(df_e) <- v_names_str
```

## 09.1 Load function of microsimulation model

```{r}
source(here("functions", "Function_Microsim_Sick-Sicker_time.R"))
# Test microsimulation function
calculate_ce_out(df_psa_input[1,])
```

## 09.2 Run microsimulation model on each parameter set of PSA input dataset

```{r}
for(i in 1:n_sim){
  df_out_temp <- calculate_ce_out(df_psa_input[i, ])
  df_c[i, ] <- df_out_temp$Cost
  df_e[i, ] <- df_out_temp$Effect
  # Display simulation progress
  if(i/(n_sim/10) == round(i/(n_sim/10),0)) { # display progress every 10%
    cat('\r', paste('            ', 'Overall progress: ', i/n_sim * 100, "% done", 
                    sep = " "))
  }
}
```

## 09.3 Cost Effectiveness Analysis

Vector with willingness-to-pay (WTP) thresholds

```{r}
v_wtp <- seq(0, 300000, by = 10000)
```

## 09.3.1 ICER

```{r}
# use dampack to calculate the ICER
# store the mean costs of each strategy in a new variable C (vector of costs)
v_C <- c(mean(out_psa$cost$no.treatment), mean(out_psa$cost$treatment))
# store the mean QALYs of each strategy in a new variable E (vector of effects)
v_E <- c(mean(out_psa$effectiveness$no.treatment), mean(out_psa$effectiveness$treatment))
calculate_icers(cost       = v_C,
                effect     = v_E,
                strategies = v_names_str)
```

## 09.3.2 Cost-Effectiveness Acceptability Curves (CEAC) and Frontier (CEAF)

```{r}
out_ceaf <- ceac(v_wtp, out_psa)
plot(out_ceaf)
```

## 09.3.3 Cost-Effectiveness Scatter plot

```{r}
plot(out_psa)
```

## 09.3.4 Expected Value of Perfect Information (EVPI)

```{r}
out_evpi <- calc_evpi(v_wtp, out_psa)
plot(out_evpi) 
```

