---
title: 'Markov Sick-Sicker model in R'
subtitle: "with age-dependency"
author: "The DARTH workgroup"
output:
  html_document: default
  pdf_document: default
---

Developed by the Decision Analysis in R for Technologies in Health (DARTH) workgroup:

Fernando Alarid-Escudero, PhD (1) 

Eva A. Enns, MS, PhD (2)	

M.G. Myriam Hunink, MD, PhD (3,4)

Hawre J. Jalal, MD, PhD (5) 

Eline M. Krijkamp, MSc (3)	

Petros Pechlivanoglou, PhD (6,7)

Alan Yang, MSc (7)

In collaboration of: 		

1. Drug Policy Program, Center for Research and Teaching in Economics (CIDE) - CONACyT, 
   Aguascalientes, Mexico
2. University of Minnesota School of Public Health, Minneapolis, MN, USA
3. Erasmus MC, Rotterdam, The Netherlands
4. Harvard T.H. Chan School of Public Health, Boston, USA
5. University of Pittsburgh Graduate School of Public Health, Pittsburgh, PA, USA
6. University of Toronto, Toronto ON, Canada
7. The Hospital for Sick Children, Toronto ON, Canada

Please cite our publications when using this code:
 
- Jalal H, Pechlivanoglou P, Krijkamp E, Alarid-Escudero F, Enns E, Hunink MG. 
An Overview of R in Health Decision Sciences. Med Decis Making. 2017; 37(3): 735-746. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X16686559
 
- Krijkamp EM, Alarid-Escudero F, Enns EA, Jalal HJ, Hunink MGM, Pechlivanoglou P. 
Microsimulation modeling for health decision sciences using R: A tutorial. 
Med Decis Making. 2018;38(3):400–22. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X18754513
 
- Krijkamp EM, Alarid-Escudero F, Enns E, Pechlivanoglou P, Hunink MM, Jalal H. 
A Multidimensional Array Representation of State-Transition Model Dynamics. 
Med Decis Mak. 2020;40(2):242-248. https://doi.org/10.1177/0272989X19893973

Copyright 2017, THE HOSPITAL FOR SICK CHILDREN AND THE COLLABORATING INSTITUTIONS. 
All rights reserved in Canada, the United States and worldwide. Copyright, 
trademarks, trade names and any and all associated intellectual property are 
exclusively owned by THE HOSPITAL FOR Sick CHILDREN and the collaborating 
institutions. These materials may be used, reproduced, modified, distributed 
and adapted with proper attribution.

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

Change `eval` to `TRUE` if you want to knit this document.

```{r}
rm(list = ls())      # clear memory (removes all the variables from the workspace)
```

# 01 Load packages

```{r, warning = FALSE, message = FALSE}
if (!require('pacman')) install.packages('pacman'); library(pacman) # use this package to conveniently install other packages
# load (install if required) packages from CRAN
p_load("dplyr", "devtools", "scales", "ellipse", "ggplot2", "lazyeval", "igraph", "truncnorm", "ggraph", "reshape2", "knitr", "dampack")                                               
# load (install if required) packages from GitHub
# install_github("DARTH-git/darthtools", force = TRUE) Uncomment if there is a newer version
p_load_gh("DARTH-git/darthtools")
```

# 02 Load functions

```{r}
# all functions are in the darthtools package
```

# 03 Input model parameters

```{r}
# Strategy names
v_names_str <- c("No Treatment", "Treatment") 

# Markov model parameters
age      <- 25                       # age at baseline
max_age  <- 55                       # maximum age of follow up
n_t      <- max_age - age            # time horizon, number of cycles
v_names_states <- c("H", "S1", "S2", "D")  # the 4 states of the model: Healthy (H), Sick (S1), 
                                     # Sicker (S2), Dead (D)
v_init   <- c("H"  = 1, 
              "S1" = 0, 
              "S2" = 0, 
              "D"  = 0)              # initial cohort distribution


# Transition probabilities (per cycle) and hazard ratios
# Read age-specific mortality rates from csv file

# MAKE SURE YOU HAVE A DATA FOLDER! Or when in the same folder remove the ../data/ part 
lt_usa_2005 <- read.csv("../data/HMD_USA_Mx_2015.csv")

v_r_HD <- lt_usa_2005 %>% 
  filter(Age >= age & Age <= (max_age-1)) %>%
  select(Total) %>%
  as.matrix()


# Transition probabilities (per cycle)
p_HD     <- rate_to_prob(r = v_r_HD, t = 1)  # probability to die when healthy
p_HS1    <- 0.15          	         # probability to become sick when healthy, conditional on surviving
p_S1H    <- 0.5           	         # probability to become healthy when sick, conditional on surviving
p_S1S2   <- 0.105         	         # probability to become sicker when sick, conditional on surviving
hr_S1    <- 3             	         # hazard ratio of death in sick vs healthy
hr_S2    <- 10            	         # hazard ratio of death in sicker vs healthy 
r_S1D    <- hr_S1 * v_r_HD  	       # rate of death in sick
r_S2D    <- hr_S2 * v_r_HD  	       # rate of death in sicker
p_S1D    <- rate_to_prob(r = r_S1D, t = 1)    # probability to die in sick
p_S2D    <- rate_to_prob(r = r_S2D, t = 1)    # probability to die in sicker

# Cost and utility inputs 
c_H      <- 2000                     # cost of remaining one cycle in the healthy state
c_S1     <- 4000                     # cost of remaining one cycle in the sick state
c_S2     <- 15000                    # cost of remaining one cycle in the sicker state
c_trt    <- 12000                    # cost of treatment(per cycle)
c_D      <- 0                        # cost of being in the death state
u_H      <- 1                        # utility when healthy
u_S1     <- 0.75                     # utility when sick
u_S2     <- 0.5                      # utility when sicker
u_D      <- 0                        # utility when dead
u_trt    <- 0.95                     # utility when being treated
d_e      <- d_c <- 0.03              # discount rate per cycle equal discount of costs and QALYs by 3%

n_str     <- length(v_names_str)     # number of strategies
n_states  <- length(v_names_states)  # number of states

# Discount weights for costs and effects
v_dwc <- 1 / (1 + d_c) ^ (0:n_t) 
v_dwe <- 1 / (1 + d_e) ^ (0:n_t) 
```

# 04 Define and initialize matrices and vectors

## 04.1 Cohort trace

```{r}
# create the markov trace matrix M capturing the proportion of the cohort in each state 
# at each cycle
m_M_notrt <- m_M_trt <- matrix(NA, 
                               nrow     = n_t + 1, ncol = n_states,
                               dimnames = list(paste("cycle", 0:n_t, sep = " "), v_names_states))

head(m_M_notrt) # show first 6 rows of the matrix 

# The cohort starts as healthy
m_M_notrt[1, ] <- m_M_trt[1, ] <- v_init  # initiate first cycle of cohort trace 
```

## 04.2 Transition probability array

```{r}
# create the transition probability array for NO treatment
a_P_notrt <- array(0,  # Create 3-D array
                   dim = c(n_states, n_states, n_t),
                   dimnames = list(v_names_states, v_names_states, 0:(n_t-1))) # name dimensions
a_P_notrt[, , 1:3]
```

Fill in the transition probability array:

```{r}
# from Healthy

# from Sick

# from Sicker

# from Dead

# create transition probability array for treatment same as no treatment
a_P_trt <- a_P_notrt  

# In the Sick-Sicker example the transitions are similar. Therefore, you can use the same values in the transition probability array. If you are modeling other strategies you have to update the transition probabilities accordingly
```

# 05 Run Markov model

```{r}
# your turn

```

# 06 Compute and Plot Epidemiological Outcomes

## 06.1 Cohort trace

```{r}
# your turn

```

## 06.2 Overall Survival (OS)

```{r}
# your turn

```

## 06.2.1 Life Expectancy (LE)

```{r}
# your turn

```

## 06.3 Disease prevalence

```{r}
# your turn

```

## 06.4 ratio of sick(S1) vs sicker(S2)

```{r}
# your turn

```

# 07 Compute Cost-Effectiveness Outcomes

## 07.1 Mean Costs and QALYs for Treatment and NO Treatment

```{r}
# your turn

```

## 07.2 Discounted Mean Costs and QALYs

```{r}
# your turn

# Store discounted costs and effectiveness for each strategy in a data frame
df_ce       <- data.frame(Strategy = NA, # <-- your value (strategy names)
                          Cost     = NA, # <-- your value (strategy costs)
                          Effect   = NA  # <-- your value (strategy QALYs)
                          )
```

## 07.3 Compute ICERs of the Markov model

```{r}
# Use the function calculate_icers() from the dampack package
df_cea <- calculate_icers(cost       = df_ce$Cost,
                          effect     = df_ce$Effect,
                          strategies = df_ce$Strategy)
df_cea 
```

## 07.4 Plot frontier of the Markov model

```{r}
plot(df_cea, effect_units = "QALYs")
```

