---
title: '3-state Markov model in R'
subtitle: "With simulation-time dependency and sensitivity analysis"
author: "The DARTH workgroup"
output:
  pdf_document: default
  html_document: default
---

Developed by the Decision Analysis in R for Technologies in Health (DARTH) workgroup:

Fernando Alarid-Escudero, PhD (1) 

Eva A. Enns, MS, PhD (2)	

M.G. Myriam Hunink, MD, PhD (3,4)

Hawre J. Jalal, MD, PhD (5) 

Eline M. Krijkamp, MSc (3)	

Petros Pechlivanoglou, PhD (6,7)

Alan Yang, MSc (7)

In collaboration of: 		

1. Division of Public Administration, Center for Research and Teaching in 
   Economics (CIDE), Aguascalientes, Mexico
2. University of Minnesota School of Public Health, Minneapolis, MN, USA
3. Erasmus MC, Rotterdam, The Netherlands
4. Harvard T.H. Chan School of Public Health, Boston, USA
5. University of Pittsburgh Graduate School of Public Health, Pittsburgh, PA, USA
6. University of Toronto, Toronto ON, Canada
7. The Hospital for Sick Children, Toronto ON, Canada

Please cite our publications when using this code:

- Alarid-Escudero F, Krijkamp EM, Enns EA, Yang A, Hunink MGM Pechlivanoglou P,
Jalal H. A Tutorial on Time-Dependent Cohort State-Transition Models in R using 
a Cost-Effectiveness Analysis Example. Medical Decision Making, 2022 (In press):
1-21. https://doi.org/10.1177/0272989X221121747

- Alarid-Escudero F, Krijkamp EM, Enns EA, Yang A, Hunink MGM Pechlivanoglou P,
Jalal H. An Introductory Tutorial on Cohort State-Transition Models in R Using a 
Cost-Effectiveness Analysis Example. Medical Decision Making, 2022 
(Online First):1-18. https://doi.org/10.1177/0272989X221103163

- Alarid-Escudero F, Krijkamp EM, Enns EA, Yang A, Hunink MGM Pechlivanoglou P,
Jalal H. An Introductory Tutorial on Cohort State-Transition Models in R Using a 
Cost-Effectiveness Analysis Example. Medical Decision Making, 2022 
(Online First):1-18. https://doi.org/10.1177/0272989X221103163

- Jalal H, Pechlivanoglou P, Krijkamp E, Alarid-Escudero F, Enns E, Hunink MG. 
An Overview of R in Health Decision Sciences. Med Decis Making. 2017; 37(3): 735-746. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X16686559

- Krijkamp EM, Alarid-Escudero F, Enns EA, Jalal HJ, Hunink MGM, Pechlivanoglou P. 
Microsimulation modeling for health decision sciences using R: A tutorial. 
Med Decis Making. 2018;38(3):400–22. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X18754513
 
- Krijkamp EM, Alarid-Escudero F, Enns E, Pechlivanoglou P, Hunink MM, Jalal H. 
A Multidimensional Array Representation of State-Transition Model Dynamics. 
Med Decis Mak. 2020;40(2):242-248. https://doi.org/10.1177/0272989X19893973

Copyright 2017, THE HOSPITAL FOR SICK CHILDREN AND THE COLLABORATING INSTITUTIONS. 
All rights reserved in Canada, the United States and worldwide. Copyright, 
trademarks, trade names and any and all associated intellectual property are 
exclusively owned by THE HOSPITAL FOR Sick CHILDREN and the collaborating 
institutions. These materials may be used, reproduced, modified, distributed 
and adapted with proper attribution.

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = TRUE)
```

Change `eval` to `TRUE` if you want to knit this document.

```{r}
rm(list = ls())      # clear memory (removes all the variables from the workspace)
```

# 01 Load packages

```{r}
if (!require('pacman')) install.packages('pacman'); library(pacman) # use this package to conveniently install other packages
# load (install if required) packages from CRAN
p_load("dplyr", "tidyr", "reshape2", "devtools", "scales", "ellipse", "ggplot2", "lazyeval", "igraph", "truncnorm", "ggraph", "reshape2", "knitr", "stringr", "diagram", "dampack")                                               
# load (install if required) packages from GitHub
# install_github("DARTH-git/darthtools", force = TRUE) #Uncomment if there is a newer version
p_load_gh("DARTH-git/darthtools")
```

# 02 Load functions

```{r}
# all functions are in the darthtools package
```

# 03 Model input

```{r}
## General setup
cycle_length    <- 1                             # cycle length equal to one year (use 1/12 for monthly)
n_cycles        <- 60                            # number of cycles
v_names_cycles  <- paste("cycle", 0:n_cycles)    # cycle names
v_names_states  <- c("Healthy", "Sick", "Dead")  # state names
n_states        <- length(v_names_states)        # number of health states 

### Discounting factors 
d_c <- 0.03 # annual discount rate for costs 
d_e <- 0.03 # annual discount rate for QALYs

### Strategies 
v_names_str     <- c("Standard of Care",         # store the strategy names
                     "Treatment A", 
                     "Treatment B")  
n_str           <- length(v_names_str)           # number of strategies

## Within-cycle correction (WCC) using Simpson's 1/3 rule 
v_wcc <- gen_wcc(n_cycles = n_cycles,  method = "Simpson1/3")

### Transition probabilities
p_HS_SoC  <- 0.05  # probability of becoming sick when healthy, conditional on surviving, under standard of care
p_HS_trtA <- 0.04  # probability of becoming sick when healthy, conditional on surviving, under treatment A
p_HS_trtB <- 0.02  # probability of becoming sick when healthy, conditional on surviving, under treatment B
p_SD      <- 0.1   # probability of dying when sick
p_HD_min  <- 0.003 # probability of dying when healthy at t = 0
p_HD_max  <- 0.01  # probability of dying when healthy at t = n_cycles
# probabilities of dying when healthy (age-dependent) - this is now a sequence of numbers
v_p_HD    <- seq(p_HD_min, p_HD_max, length.out = n_cycles)   

### State rewards
#### Costs 
c_H       <- 400   # cost of one cycle in healthy state
c_S       <- 1000  # cost of one cycle in sick state
c_D       <- 0     # cost of one cycle in dead state
c_trtA    <- 800   # cost of treatment A (per cycle) in healthy state
c_trtB    <- 1500  # cost of treatment B (per cycle) in healthy state
#### Utilities
u_H       <- 1     # utility when healthy 
u_S       <- 0.5   # utility when sick
u_D       <- 0     # utility when dead

### Discount weight for costs and effects 
v_dwc     <- 1 / ((1 + (d_e * cycle_length)) ^ (0:n_cycles))
v_dwe     <- 1 / ((1 + (d_c * cycle_length)) ^ (0:n_cycles))
```

# 04 Construct state-transition models

```{r}
m_P_diag <- matrix(0, nrow = n_states, ncol = n_states, dimnames = list(v_names_states, v_names_states))
m_P_diag["Healthy", "Sick" ]     = "" 
m_P_diag["Healthy", "Dead" ]     = ""
m_P_diag["Healthy", "Healthy" ]  = ""
m_P_diag["Sick"   , "Dead" ]     = ""
m_P_diag["Sick"   , "Sick" ]     = ""
m_P_diag["Dead"   , "Dead" ]     = ""
layout.fig <- c(2, 1)
plotmat(t(m_P_diag), t(layout.fig), self.cex = 0.5, curve = 0, arr.pos = 0.8,  
        latex = T, arr.type = "curved", relsize = 0.85, box.prop = 0.8, 
        cex = 0.8, box.cex = 0.7, lwd = 1)
```

## 04.1 Initial state vector

```{r}
# All starting healthy
v_m_init <- c("Healthy" = 1, "Sick" = 0, "Dead" = 0)  
v_m_init
```

## 04.2 Initialize cohort traces

```{r}
### Initialize cohort trace for SoC 
m_M_SoC <- matrix(0, 
                  nrow = (n_cycles + 1), ncol = n_states, 
                  dimnames = list(v_names_cycles, v_names_states))
# Store the initial state vector in the first row of the cohort trace
m_M_SoC[1, ] <- v_m_init

## Initialize cohort traces for treatments A and B
# Structure and initial states are the same as for SoC
m_M_trtA <- m_M_trtB <- m_M_SoC
```

## 04.3 Create transition probability arrays

```{r}
## Create transition probability arrays for strategy SoC 
### Initialize transition probability array for strategy SoC 
# All transitions to a non-death state are assumed to be conditional on survival
a_P_SoC <- array(0,  # Create 3-D array
                dim = c(n_states, n_states, n_cycles),
                dimnames = list(v_names_states, v_names_states, 
                                v_names_cycles[-length(v_names_cycles)])) # name the dimensions of the array 

### Fill in array
## Standard of Care
# from Healthy
a_P_SoC["Healthy", "Healthy", ] <- (1 - v_p_HD) * (1 - p_HS_SoC)
a_P_SoC["Healthy", "Sick",    ] <- (1 - v_p_HD) *      p_HS_SoC
a_P_SoC["Healthy", "Dead",    ] <-      v_p_HD

# from Sick
a_P_SoC["Sick", "Sick", ] <- 1 - p_SD
a_P_SoC["Sick", "Dead", ] <-     p_SD

# from Dead
a_P_SoC["Dead", "Dead", ] <- 1

## Treatment A
a_P_trtA <- a_P_SoC
a_P_trtA["Healthy", "Healthy", ] <- (1 - v_p_HD) * (1 - p_HS_trtA)
a_P_trtA["Healthy", "Sick",    ] <- (1 - v_p_HD) *      p_HS_trtA

## Treatment B
a_P_trtB <- a_P_SoC
a_P_trtB["Healthy", "Healthy", ] <- (1 - v_p_HD) * (1 - p_HS_trtB)
a_P_trtB["Healthy", "Sick",    ] <- (1 - v_p_HD) *      p_HS_trtB

## Check if transition array and probabilities are valid
# Check that transition probabilities are in [0, 1]
check_transition_probability(a_P_SoC,  verbose = TRUE)
check_transition_probability(a_P_trtA, verbose = TRUE)
check_transition_probability(a_P_trtB, verbose = TRUE)
# Check that all rows sum to 1
check_sum_of_transition_array(a_P_SoC,  n_states = n_states, n_cycles = n_cycles, verbose = TRUE)
check_sum_of_transition_array(a_P_trtA, n_states = n_states, n_cycles = n_cycles, verbose = TRUE)
check_sum_of_transition_array(a_P_trtB, n_states = n_states, n_cycles = n_cycles, verbose = TRUE)
```

# 05 Run Markov model

```{r}
# Iterative solution of age-dependent cSTM
for(t in 1:n_cycles){
  ## Fill in cohort trace
  # For SoC
  m_M_SoC[t + 1, ]  <- m_M_SoC[t, ]  %*% a_P_SoC[, , t]
  # For strategy A
  m_M_trtA[t + 1, ] <- m_M_trtA[t, ] %*% a_P_trtA[, , t]
  # For strategy B
  m_M_trtB[t + 1, ] <- m_M_trtB[t, ] %*% a_P_trtB[, , t]
}

## Store the cohort traces in a list 
l_m_M <- list(SoC =  m_M_SoC,
              A   =  m_M_trtA,
              B   =  m_M_trtB)
names(l_m_M) <- v_names_str
```

# 06 Plot Outputs

## 06.1 Plot the cohort trace for strategies SoC

```{r, message = FALSE}
plot_trace(m_M_SoC) 
```

## 06.2 Overall Survival (OS)

Print the overall survival for the Standard of Care

```{r}
v_os_SoC <- 1 - m_M_SoC[, "Dead"]    # calculate the overall survival (OS) probability
v_os_SoC <- rowSums(m_M_SoC[, 1:2])  # alternative way of calculating the OS probability   

plot(v_os_SoC, type = 'l', 
     ylim = c(0, 1),
     ylab = "Survival probability",
     xlab = "Cycle",
     main = "Overall Survival")  # create a simple plot showing the OS

# add grid 
grid(nx = n_cycles, ny = 10, col = "lightgray", lty = "dotted", lwd = par("lwd"), 
     equilogs = TRUE) 
```

## 06.2.1 Life Expectancy (LE)

```{r}
le_SoC <- sum(v_os_SoC)  # summing probability of OS over time  (i.e. life expectancy)
```

## 06.2.2 Disease prevalence

```{r}
v_prev <- m_M_SoC[, "Sick"]/v_os_SoC
plot(v_prev,
     ylim = c(0, 1),
     ylab = "Prevalence",
     xlab = "Cycle",
     main = "Disease prevalence")
```

# 07 State Rewards 

```{r}
## Scale by the cycle length 
# Vector of state utilities under strategy SoC
v_u_SoC    <- c(H  = u_H, 
                S  = u_S,
                D  = u_D) * cycle_length
# Vector of state costs under strategy SoC
v_c_SoC    <- c(H  = c_H, 
                S  = c_S,
                D  = c_D) * cycle_length
# Vector of state utilities under treatment A
v_u_trtA   <- c(H  = u_H, 
                S  = u_S, 
                D  = u_D) * cycle_length
# Vector of state costs under treatment A
v_c_trtA   <- c(H  = c_H + c_trtA, 
                S  = c_S, 
                D  = c_D) * cycle_length
# Vector of state utilities under treatment B
v_u_trtB   <- c(H  = u_H, 
                S  = u_S, 
                D  = u_D) * cycle_length
# Vector of state costs under treatment B
v_c_trtB   <- c(H  = c_H + c_trtB, 
                S  = c_S, 
                D  = c_D) * cycle_length

## Store state rewards 
# Store the vectors of state utilities for each strategy in a list 
l_u   <- list(SQ = v_u_SoC,
              A  = v_u_trtA,
              B  = v_u_trtB)
# Store the vectors of state cost for each strategy in a list 
l_c   <- list(SQ = v_c_SoC,
              A  = v_c_trtA,
              B  = v_c_trtB)

# assign strategy names to matching items in the lists
names(l_u) <- names(l_c) <- v_names_str
```

# 08 Compute expected outcomes 

```{r}
# Create empty vectors to store total utilities and costs 
v_tot_qaly <- v_tot_cost <- vector(mode = "numeric", length = n_str)
names(v_tot_qaly) <- names(v_tot_cost) <- v_names_str

## Loop through each strategy and calculate total utilities and costs 
for (i in 1:n_str) {
  v_u_str <- l_u[[i]]   # select the vector of state utilities for the i-th strategy
  v_c_str <- l_c[[i]]   # select the vector of state costs for the i-th strategy
  
  ### Expected QALYs and costs per cycle 
  ## Vector of QALYs and Costs
  # Apply state rewards 
  v_qaly_str <- l_m_M[[i]] %*% v_u_str # sum the utilities of all states for each cycle
  v_cost_str <- l_m_M[[i]] %*% v_c_str # sum the costs of all states for each cycle
  
  ### Discounted total expected QALYs and Costs per strategy and apply within-cycle correction if applicable
  # QALYs
  v_tot_qaly[i] <- t(v_qaly_str) %*% (v_dwe * v_wcc)
  # Costs
  v_tot_cost[i] <- t(v_cost_str) %*% (v_dwc * v_wcc)
}
```

# 09 Cost-effectiveness analysis (CEA) 

```{r}
## Incremental cost-effectiveness ratios (ICERs) 
df_cea <- calculate_icers(cost       = v_tot_cost, 
                          effect     = v_tot_qaly,
                          strategies = v_names_str)
df_cea
```

```{R}
## CEA table in proper format 
table_cea <- format_table_cea(df_cea) 
table_cea
```

```{r}
## CEA frontier 
plot(df_cea, label = "all", txtsize = 14) +
  expand_limits(x = max(table_cea$QALYs) + 0.1) +
  theme(legend.position = c(0.8, 0.3))
```

# 10 Deterministic Sensitivity Analysis (DSA)

```{r}
## Load model, CEA and PSA functions 
source('Functions_markov_3state_time.R')
```

## 10.1 Model input for SA

```{r}
l_params_all <- list(
  # Transition probabilities
  # probability of dying
  p_SD      = 0.1,   # from sick         
  p_HD_min  = 0.003, # from healthy at t = 0
  p_HD_max  = 0.01,  # from healthy at t = n_cycles
  # probability of becoming sick when healthy, conditional on surviving
  p_HS_SoC  = 0.05,  # standard of care
  p_HS_trtA = 0.04,  # treatment A
  p_HS_trtB = 0.02,  # treatment B
  # Costs 
  c_H       = 400,   # cost of one cycle in healthy state
  c_S       = 1000,  # cost of one cycle in sick state
  c_D       = 0,     # cost of one cycle in dead state
  c_trtA    = 800,   # cost of treatment A (per cycle) in healthy state
  c_trtB    = 1500,  # cost of treatment B (per cycle) in healthy state
  # Utilities 
  u_H       = 1,     # utility when healthy 
  u_S       = 0.5,   # utility when sick
  u_D       = 0,     # utility when dead
  # Discount rates
  d_e       = 0.03,  # discount rate per cycle equal discount of costs and QALYs by 3%
  d_c       = 0.03,  # discount rate per cycle equal discount of costs and QALYs by 3%
  # Time horizon
  n_cycles  = 60     # simulation time horizon (number of cycles)
)
```

**Test model functions**

Two functions were defined in the `Functions_markov_3state_time.R` file. The first was the `decision_model()` function which takes in model parameters and outputs the Markov trace for all three strategies. The second is the `calculate_ce_out()` function which calls `decision_model()` and uses the resulting Markov trace to compute the total costs, QALYs, and net monetary benefit (NMB) for each strategy, returning a data frame of costs and QALYs.

```{r}
# Try the decision_model() function
l_trace <- decision_model(l_params_all)
l_trace$m_M_SoC[1:5,]
l_trace$a_P_SoC[,,1:3]

# Try the calculate_ce_out() function
df_ce <- calculate_ce_out(l_params_all)
df_ce

# Get strategies names (will be used to label plots)
v_names_str <- df_ce$Strategy
n_str <- length(v_names_str)
```

## 10.2 One-way sensitivity analysis (OWSA)

```{r}
options(scipen = 999) # disabling scientific notation in R
# dataframe containing all parameters, their base case values, and the min and 
# max values of the parameters of interest 
df_params_owsa <- data.frame(pars = c("c_trtA", "c_trtB", "c_S"),
                             min  = c(300 , 500,   500),  # min parameter values
                             max  = c(1200, 2000, 2000)  # max parameter values
                             )
owsa_nmb  <- run_owsa_det(params_range     = df_params_owsa,    # dataframe with parameters for OWSA
                          params_basecase  = l_params_all,      # list with all parameters
                          nsamp            = 100,               # number of parameter values
                          FUN              = calculate_ce_out,  # function to compute outputs
                          outcomes         = c("NMB"),          # output to do the OWSA on
                          strategies       = v_names_str,       # names of the strategies
                          n_wtp            = 5000)              # extra argument to pass to FUN
```

```{r}
plot(owsa_nmb, txtsize = 10, n_x_ticks = 4, 
     facet_scales = "free") +
     theme(legend.position = "bottom")
```

### 10.2.1 Optimal strategy with OWSA

```{r}
owsa_opt_strat(owsa = owsa_nmb, txtsize = 10)
```

### 10.2.2 Tornado plot

```{r}
owsa_tornado(owsa = owsa_nmb)
```

## 10.3 Two-way sensitivity analysis (TWSA)

```{r}
# dataframe containing all parameters, their basecase values, and the min and 
# max values of the parameters of interest
df_params_twsa <- data.frame(pars = c("c_trtA", "c_trtB"),
                             min  = c(300, 500),  # min parameter values
                             max  = c(1200, 2000) # max parameter values
                             )

twsa_nmb <- run_twsa_det(params_range    = df_params_twsa,    # dataframe with parameters for TWSA
                         params_basecase = l_params_all,      # list with all parameters
                         nsamp           = 40,                # number of parameter values
                         FUN             = calculate_ce_out,  # function to compute outputs
                         outcomes        = "NMB",             # output to do the TWSA on
                         strategies      = v_names_str,       # names of the strategies
                         n_wtp           = 5000)              # extra argument to pass to FUN
```

### 10.3.1 Plot TWSA

```{r}
plot(twsa_nmb)
```

# 11 Probabilistic Sensitivity Analysis (PSA) 

## 11.1 Model input

```{r}
# Store the parameter names into a vector
v_names_params <- names(l_params_all)

## Test functions to generate CE outcomes and PSA dataset 
# Test function to compute CE outcomes
calculate_ce_out(l_params_all) 

# Test function to generate PSA input dataset
generate_psa_params(10) 

## Generate PSA dataset 
# Number of simulations
n_sim <- 1000

# Generate PSA input dataset
df_psa_input <- generate_psa_params(n_sim = n_sim)
# First six observations
head(df_psa_input)

### Histogram of parameters 
ggplot(melt(df_psa_input, variable.name = "Parameter"), aes(x = value)) +
  facet_wrap(~Parameter, scales = "free") +
  geom_histogram(aes(y = ..density..)) +
  ylab("") +
  theme_bw(base_size = 16) + 
  theme(axis.text = element_text(size = 6),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y  = element_blank(),
        axis.ticks.y = element_blank()) 
```

## 11.2 Run PSA

```{r}
# Initialize data.frames with PSA output 
# data.frame of costs
df_c <- as.data.frame(matrix(0, 
                             nrow = n_sim,
                             ncol = n_str))
colnames(df_c) <- v_names_str
# data.frame of effectiveness
df_e <- as.data.frame(matrix(0, 
                             nrow = n_sim,
                             ncol = n_str))
colnames(df_e) <- v_names_str

# Conduct probabilistic sensitivity analysis
# Run Markov model on each parameter set of PSA input dataset
n_time_init_psa_series <- Sys.time()
for (i in 1:n_sim) { # i <- 1
  l_psa_input <- update_param_list(l_params_all, df_psa_input[i,])
  # Outcomes
  l_out_ce_temp  <- calculate_ce_out(l_psa_input)
  df_c[i, ]  <- l_out_ce_temp$Cost  
  df_e[i, ]  <- l_out_ce_temp$Effect
  # Display simulation progress
  if (i/(n_sim/100) == round(i/(n_sim/100), 0)) { # display progress every 5%
    cat('\r', paste(i/n_sim * 100, "% done", sep = " "))
  }
}
n_time_end_psa_series <- Sys.time()
n_time_total_psa_series <- n_time_end_psa_series - n_time_init_psa_series
print(paste0("PSA with ", scales::comma(n_sim), " simulations run in series in ", 
             round(n_time_total_psa_series, 2), " ", 
             units(n_time_total_psa_series)))
```

# 11.3 Visualize PSA results for CEA 

```{r}
### Create PSA object 
l_psa <- make_psa_obj(cost          = df_c, 
                      effectiveness = df_e, 
                      parameters    = df_psa_input, 
                      strategies    = v_names_str)
l_psa$strategies <- v_names_str
colnames(l_psa$effectiveness) <- v_names_str
colnames(l_psa$cost) <- v_names_str

# Vector with willingness-to-pay (WTP) thresholds.
v_wtp <- seq(0, 30000, by = 1000)
```

## 11.3.1 Cost-Effectiveness Scatter plot 

```{r}
### Cost-Effectiveness Scatter plot 
txtsize <- 13
gg_scattter <- plot_psa(l_psa, txtsize = txtsize) +
  ggthemes::scale_color_colorblind() +
  ggthemes::scale_fill_colorblind() +
  scale_y_continuous("Cost (Thousand $)", 
                     breaks = number_ticks(10),
                     labels = function(x) x/1000) +
  xlab("Effectiveness (QALYs)") +
  guides(col = guide_legend(nrow = 2)) +
  theme(legend.position = "bottom")
gg_scattter
```

## 11.3.2 Incremental cost-effectiveness ratios (ICERs) with probabilistic output

```{r}
### Incremental cost-effectiveness ratios (ICERs) with probabilistic output 
# Compute expected costs and effects for each strategy from the PSA
df_out_ce_psa <- summary(l_psa)
df_cea_psa <- calculate_icers(cost       = df_out_ce_psa$meanCost, 
                              effect     = df_out_ce_psa$meanEffect,
                              strategies = df_out_ce_psa$Strategy)
df_cea_psa
```

## 11.3.3 Plot cost-effectiveness frontier with probabilistic output

```{r}
### Plot cost-effectiveness frontier with probabilistic output 
plot_icers(df_cea_psa, label = "all", txtsize = txtsize) +
  expand_limits(x = max(table_cea$QALYs) + 0.1) +
  theme(legend.position = c(0.8, 0.3))
```

## 11.3.4 Cost-effectiveness acceptability curves (CEACs) and frontier (CEAF)

```{r}
### Cost-effectiveness acceptability curves (CEACs) and frontier (CEAF) 
ceac_obj <- ceac(wtp = v_wtp, psa = l_psa)
# Regions of highest probability of cost-effectiveness for each strategy
summary(ceac_obj)
# CEAC & CEAF plot
gg_ceac <- plot_ceac(ceac_obj, txtsize = txtsize, xlim = c(0, NA), n_x_ticks = 14) +
  ggthemes::scale_color_colorblind() +
  ggthemes::scale_fill_colorblind() +
  theme(legend.position = c(0.8, 0.48))
gg_ceac
```

## 11.3.5 Expected Loss Curves (ELCs)

```{r}
### Expected Loss Curves (ELCs) 
elc_obj <- calc_exp_loss(wtp = v_wtp, psa = l_psa)
elc_obj

# ELC plot
gg_elc <- plot_exp_loss(elc_obj, log_y = FALSE, 
               txtsize = txtsize, xlim = c(0, NA), n_x_ticks = 14,
               col = "full") +
  ggthemes::scale_color_colorblind() +
  ggthemes::scale_fill_colorblind() +
  # geom_point(aes(shape = as.name("Strategy"))) +
  scale_y_continuous("Expected Loss (Thousand $)", 
                     breaks = number_ticks(10),
                     labels = function(x) x/1000) +
  theme(legend.position = c(0.4, 0.7),)
gg_elc
```

## 11.3.6 Expected value of perfect information (EVPI) 

```{r}
### Expected value of perfect information (EVPI) 
evpi <- calc_evpi(wtp = v_wtp, psa = l_psa)
# EVPI plot
gg_evpi <- plot_evpi(evpi, effect_units = "QALY", 
                     txtsize = txtsize, xlim = c(0, NA), n_x_ticks = 14) +
  scale_y_continuous("EVPI (Thousand $)", 
                     breaks = number_ticks(10),
                     labels = function(x) x/1000)
gg_evpi
```



