---
title: '3-state Markov model in R'
subtitle: 'with OWSA, TWSA, PSA'
author: "The DARTH workgroup"
output:
  pdf_document: default
  html_document: default
---

Developed by the Decision Analysis in R for Technologies in Health (DARTH) workgroup:

Fernando Alarid-Escudero, PhD (1) 

Eva A. Enns, MS, PhD (2)	

M.G. Myriam Hunink, MD, PhD (3,4)

Hawre J. Jalal, MD, PhD (5) 

Eline M. Krijkamp, MSc (3)	

Petros Pechlivanoglou, PhD (6,7)

Alan Yang, MSc (7)

In collaboration of: 		

1. Division of Public Administration, Center for Research and Teaching in 
   Economics (CIDE), Aguascalientes, Mexico
2. University of Minnesota School of Public Health, Minneapolis, MN, USA
3. Erasmus MC, Rotterdam, The Netherlands
4. Harvard T.H. Chan School of Public Health, Boston, USA
5. University of Pittsburgh Graduate School of Public Health, Pittsburgh, PA, USA
6. University of Toronto, Toronto ON, Canada
7. The Hospital for Sick Children, Toronto ON, Canada

Please cite our publications when using this code:
 
- Jalal H, Pechlivanoglou P, Krijkamp E, Alarid-Escudero F, Enns E, Hunink MG. 
An Overview of R in Health Decision Sciences. Med Decis Making. 2017; 37(3): 735-746. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X16686559

- Alarid-Escudero F, Krijkamp EM, Enns EA, Yang A, Hunink MGM Pechlivanoglou P,
Jalal H. An Introductory Tutorial on Cohort State-Transition Models in R Using a Cost-Effectiveness Analysis Example. arXiv:200107824v3. 2021:1-26. http://arxiv.org/abs/2001.07824

- Krijkamp EM, Alarid-Escudero F, Enns EA, Jalal HJ, Hunink MGM, Pechlivanoglou P. 
Microsimulation modeling for health decision sciences using R: A tutorial. 
Med Decis Making. 2018;38(3):400–22. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X18754513
 
- Krijkamp EM, Alarid-Escudero F, Enns E, Pechlivanoglou P, Hunink MM, Jalal H. 
A Multidimensional Array Representation of State-Transition Model Dynamics. 
Med Decis Making. Online First https://doi.org/10.1177/0272989X19893973

Copyright 2017, THE HOSPITAL FOR SICK CHILDREN AND THE COLLABORATING INSTITUTIONS. 
All rights reserved in Canada, the United States and worldwide. Copyright, 
trademarks, trade names and any and all associated intellectual property are 
exclusively owned by THE HOSPITAL FOR Sick CHILDREN and the collaborating 
institutions. These materials may be used, reproduced, modified, distributed 
and adapted with proper attribution.

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = T)
```

Change `eval` to `TRUE` if you want to knit this document.

```{r}
rm(list = ls())  # clear memory (removes all the variables from the workspace)
```

# 01 Load packages

```{r}
if (!require('pacman')) install.packages('pacman'); library(pacman) # use this package to conveniently install other packages
# load (install if required) packages from CRAN
p_load("diagram", "dampack", "reshape2") 
# install_github("DARTH-git/darthtools", force = TRUE) Uncomment if there is a newer version
p_load_gh("DARTH-git/darthtools")
```

# 02 Load functions

```{r}
# Load decision model
source("Functions_markov_3state_time.R")
```

# 03 Set up model inputs and functions

# 03.1 Set default / base case input parameters

Create list `l_params_all` to store all input probabilities, cost and utilities.

```{r}
l_params_all <- list(
  # Transition probabilities
  # probability of dying
  p_SD      = 0.1,   # from sick         
  p_HD_min  = 0.003, # from healthy at t = 0
  p_HD_max  = 0.01,  # from healthy at t = n_cycles
  # probability of becoming sick when healthy, conditional on surviving
  p_HS_SoC  = 0.05,  # standard of care
  p_HS_trtA = 0.04,  # treatment A
  p_HS_trtB = 0.02,  # treatment B
  # Costs 
  c_H       = 400,   # cost of one cycle in healthy state
  c_S       = 1000,  # cost of one cycle in sick state
  c_D       = 0,     # cost of one cycle in dead state
  c_trtA    = 800,   # cost of treatment A (per cycle) in healthy state
  c_trtB    = 1500,  # cost of treatment B (per cycle) in healthy state
  # Utilities 
  u_H       = 1,     # utility when healthy 
  u_S       = 0.5,   # utility when sick
  u_D       = 0,     # utility when dead
  # Discount rates
  d_e       = 0.03,  # discount rate per cycle equal discount of costs and QALYs by 3%
  d_c       = 0.03,  # discount rate per cycle equal discount of costs and QALYs by 3%
  # Time horizon
  n_cycles  = 60     # simulation time horizon (number of cycles)
)

# store the parameter names into a vector
v_names_params <- names(l_params_all)
```

## 03.2 Test model functions

Two functions were defined in the `Functions_markov_3state_time.R` file. The first was the `decision_model()` function which takes in model parameters and outputs the Markov trace for all three strategies. The second is the `calculate_ce_out()` function which calls `decision_model()` and uses the resulting Markov trace to compute the total costs, QALYs, and net monetary benefit (NMB) for each strategy, returning a data frame of costs and QALYs.

```{r}
# Try the decision_model() function
l_trace <- decision_model(l_params_all)
l_trace$m_M_SoC[1:5,]
l_trace$a_P_SoC[,,1:3]

# Try the calculate_ce_out() function
df_ce <- calculate_ce_out(l_params_all)
df_ce

# Get strategies names (will be used to label plots)
v_names_str <- df_ce$Strategy
n_str <- length(v_names_str)
```


# 04 Deterministic Sensitivity Analysis

## 04.1 One-way sensitivity analysis (OWSA)

```{r}
options(scipen = 999) # disabling scientific notation in R

# dataframe containing all parameters, their base case values, and the min and 
# max values of the parameters of interest 
df_params_owsa <- data.frame(pars = c("c_trtA", "c_trtB", "c_S"),
                             min  = c(300 , 500, 500),  # min parameter values
                             max  = c(1200, 2000, 2000)  # max parameter values
                             )

owsa_nmb  <- run_owsa_det(params_range     = df_params_owsa,    # dataframe with parameters for OWSA
                          params_basecase  = l_params_all,      # list with all parameters
                          nsamp            = 100,               # number of parameter values
                          FUN              = calculate_ce_out,  # function to compute outputs
                          outcomes         = c("NMB"),          # output to do the OWSA on
                          strategies       = v_names_str,       # names of the strategies
                          wtp              = 5000)              # extra argument to pass to FUN
```

## 04.1.1 Plot OWSA

```{r}
plot(owsa_nmb, txtsize = 10, n_x_ticks = 4, 
     facet_scales = "free") +
     theme(legend.position = "bottom")
```

## 04.1.2 Optimal strategy with OWSA

```{r}
owsa_opt_strat(owsa = owsa_nmb, txtsize = 10)
```

## 04.2 Two-way sensitivity analysis (TWSA)

```{r}
# dataframe containing all parameters, their basecase values, and the min and 
# max values of the parameters of interest
df_params_twsa <- data.frame(pars = c("c_trtA", "c_trtB"),
                             min  = c(300, 500),  # min parameter values
                             max  = c(1200, 2000) # max parameter values
                             )

twsa_nmb <- run_twsa_det(params_range    = df_params_twsa,    # dataframe with parameters for TWSA
                         params_basecase = l_params_all,      # list with all parameters
                         nsamp           = 40,                # number of parameter values
                         FUN             = calculate_ce_out,  # function to compute outputs
                         outcomes        = "NMB",             # output to do the TWSA on
                         strategies      = v_names_str,       # names of the strategies
                         wtp             = 5000)              # extra argument to pass to FUN
```

## 04.2.1 Plot TWSA

```{r}
plot(twsa_nmb)
```

# 05 Probabilistic Sensitivity Analysis (PSA)

## 05.1 Sample from parameter distribution reflecting parameter uncertainty

```{r}
# Function to generate PSA input dataset
gen_psa <- function(n_sim = 1000, seed = 071818){
  set.seed(seed) # set a seed to be able to reproduce the same results
  df_psa <- data.frame(
    # Transition probabilities
    # probability of dying
    p_SD      = rbeta(n_sim, shape1 = 22.4, shape2 = 201.6),   # from sick          
    p_HD_min  = 0.003, # from healthy at t = 0
    p_HD_max  = 0.01,  # from healthy at t = n_cycles
    # probability of becoming sick when healthy, conditional on surviving
    p_HS_SoC  = rbeta(n_sim, shape1 = 24, shape2 = 450),  # standard of care
    p_HS_trtA = rbeta(n_sim, shape1 = 15, shape2 = 368),  # treatment A
    p_HS_trtB = rbeta(n_sim, shape1 = 16, shape2 = 767),  # treatment B
    # Costs
    c_H       = rgamma(n_sim, shape = 16, scale = 25),   # cost of one cycle in healthy state
    c_S       = rgamma(n_sim, shape = 100, scale = 10),  # cost of one cycle in sick state
    c_D       = 0,     # cost of one cycle in dead state
    c_trtA    = 800,   # cost of treatment A (per cycle) in healthy state
    c_trtB    = 1500,  # cost of treatment B (per cycle) in healthy state
    # Utilities
    u_H       = rbeta(n_sim, shape1 =  1.5, shape2 = 0.0015), # utility when healthy 
    u_S       = rbeta(n_sim, shape1 = 49.5, shape2 = 49.5),   # utility when sick
    u_D       = 0,     # utility when dead
    # Discount rates
    d_e       = 0.03,  # discount rate per cycle equal discount of costs and QALYs by 3%
    d_c       = 0.03,  # discount rate per cycle equal discount of costs and QALYs by 3%
    # Time horizon
    n_cycles  = 60     # simulation time horizon (number of cycles)
  )
  return(df_psa)
}

# Number of simulations
n_sim <- 1000

# Generate PSA input dataset
df_psa_input <- gen_psa(n_sim = n_sim)
# First six observations
head(df_psa_input)
```

## 0.5.2 Visualize PSA samples
```{r}
# Histogram of parameters
ggplot(melt(df_psa_input, variable.name = "Parameter"), aes(x = value)) +
       facet_wrap(~Parameter, scales = "free") +
       geom_histogram(aes(y = ..density..)) +
       theme_bw(base_size = 16) + 
       theme(axis.text = element_text(size=8))
```

## 05.3 Conduct probabilistic sensitivity analysis
```{r}
# Initialize dataframes with PSA output 
# Dataframe of costs
df_c <- as.data.frame(matrix(0, 
                      nrow = n_sim,
                      ncol = n_str))
colnames(df_c) <- v_names_str
# Dataframe of effectiveness
df_e <- as.data.frame(matrix(0, 
                      nrow = n_sim,
                      ncol = n_str))
colnames(df_e) <- v_names_str

# Run Markov model on each parameter set of PSA input dataset
for(i in 1:n_sim){
  l_out_temp <- calculate_ce_out(df_psa_input[i, ])
  df_c[i, ] <- l_out_temp$Cost
  df_e[i, ] <- l_out_temp$Effect
  # Display simulation progress
  if(i/(n_sim/10) == round(i/(n_sim/10), 0)) { # display progress every 10%
    cat('\r', paste(i/n_sim * 100, "% done", sep = " "))
  }
}
```

## 05.4 Create PSA object for dampack

```{r}
l_psa <- make_psa_obj(cost          = df_c, 
                      effectiveness = df_e, 
                      parameters    = df_psa_input, 
                      strategies    = v_names_str)

# Save PSA output
# save(df_psa_input, df_c, df_e, v_names_str, n_str, l_psa,
#      file = "markov_3state_PSA_dataset.RData")
```

## 05.5 PSA Visualization and Analysis

## 05.5.1 Cost-effectiveness scatter plot

```{r}
plot(l_psa, xlim = c(9.5, 22.5))
```

## 05.5.2 CEA with probabilistic output

```{r}
# Compute expected costs and effects for each strategy from the PSA
df_out_ce_psa <- summary(l_psa)

# Calculate incremental cost-effectiveness ratios (ICERs)
df_cea_psa <- calculate_icers(cost       = df_out_ce_psa$meanCost, 
                              effect     = df_out_ce_psa$meanEffect,
                              strategies = df_out_ce_psa$Strategy)
df_cea_psa

# Save CEA table with ICERs
# # As .RData
# save(df_cea_psa, 
#      file = "markov_3state_probabilistic_CEA_results.RData")
# # As .csv
# write.csv(df_cea_psa, 
#           file = "markov_3state_probabilistic_CEA_results.csv")
```

## 05.5.3 Plot cost-effectiveness frontier

```{r}
plot(df_cea_psa)
```

## 05.5.4 Cost-effectiveness acceptability curves (CEACs) and frontier (CEAF)

```{r}
# vector of willingness-to-pay values 
v_wtp <- seq(from = 0, to = 30000, by = 1000)
# generate cost-effectiveness acceptability curve
ceac_obj <- ceac(wtp = v_wtp, psa = l_psa)
# Regions of highest probability of cost-effectiveness for each strategy
summary(ceac_obj)
# CEAC & CEAF plot
plot(ceac_obj)
```

## 05.5.5 Expected Loss Curves (ELCs)

The expected loss is the the quantification of the foregone benefits when choosing a suboptimal strategy given current evidence.

```{r}
elc_obj <- calc_exp_loss(wtp = v_wtp, psa = l_psa)
# ELC plot
plot(elc_obj, log_y = FALSE)
```

## 05.5.6 Expected value of perfect information (EVPI)

```{r}
evpi <- calc_evpi(wtp = v_wtp, psa = l_psa)
# EVPI plot
plot(evpi, effect_units = "QALY")
```
