<<<<<<< HEAD
---
title: 'Markov Sick-Sicker model in R'
subtitle: "With simulation-time dependency"
author: "The DARTH workgroup"
output:
  pdf_document: default
  html_document: default
keep_tex: yes
self_contained: no
---

This code forms the basis for the state-transition model of the tutorial: 
'A Tutorial on Time-Dependent Cohort State-Transition Models in R using a Cost-Effectiveness Analysis Example' 

Authors: 
- Fernando Alarid-Escudero <fernando.alarid@cide.edu>
- Eline Krijkamp
- Eva A. Enns
- Alan Yang
- M.G. Myriam Hunink
- Petros Pechlivanoglou
- Hawre Jalal
Please cite the article when using this code

To program this tutorial we used:
R version 4.0.5 (2021-03-31)
Platform: 64-bit operating system, x64-based processor
Running under: Mac OS 12.2.1
RStudio: Version 1.4.1717 2009-2021 RStudio, Inc

This code implements a simulation-time-dependent Sick-Sicker cSTM model to  conduct a CEA of two strategies:
- Standard of Care (SoC): best available care for the patients with the disease. This scenario reflects the natural history of the disease  
  progression.
- Strategy AB: This strategy combines treatment A and treatment B. The disease  progression is reduced, and individuals in the Sick state have an  
  improved quality of life.

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

Change `eval` to `TRUE` if you want to knit this document.

```{r}
rm(list = ls())      # clear memory (removes all the variables from the workspace)
```

# 01 Load packages

```{r, warning = F, message = F}
if (!require('pacman')) install.packages('pacman'); library(pacman) # use this package to conveniently install other packages
# load (install if required) packages from CRAN
p_load("dplyr", "tidyr", "reshape2", "devtools", "scales", "ellipse", "ggplot2", "ggrepel", "gridExtra", "lazyeval", "igraph", "truncnorm", "ggraph", "reshape2", "patchwork", "knitr", "stringr", "diagram", "dampack")                                               
# load (install if required) packages from GitHub
# install_github("DARTH-git/darthtools", force = TRUE) #Uncomment if there is a newer version
p_load_gh("DARTH-git/darthtools")
```

# 02 Load functions

```{r}
# all functions are in the darthtools package
```

# 03 Model input

```{r}
## General setup 
cycle_length <- 1   # cycle length equal to one year (use 1/12 for monthly)
n_age_init   <- 25  # age at baseline
n_age_max    <- 100 # maximum age of follow up
n_cycles     <- (n_age_max - n_age_init)/cycle_length # time horizon, number of cycles
# Age labels 
v_age_names  <- paste(rep(n_age_init:(n_age_max-1), each = 1/cycle_length), 
                      1:(1/cycle_length), 
                      sep = ".")
# the 4 health states of the model:
v_names_states <- c("H",  # Healthy (H)
                    "S1", # Sick (S1)
                    "S2", # Sicker (S2)
                    "D")  # Dead (D)
                                           
n_states <- length(v_names_states)   # number of health states 

### Discounting factors 
d_c <- 0.03 # annual discount rate for costs 
d_e <- 0.03 # annual discount rate for QALYs

### Strategies 
v_names_str <- c("Standard of care", # store the strategy names
                 "Strategy AB") 
n_str       <- length(v_names_str)   # number of strategies

## Within-cycle correction (WCC) using Simpson's 1/3 rule 
v_wcc  <- gen_wcc(n_cycles = n_cycles, method = "Simpson1/3")

### Transition rates (annual), and hazard ratios (HRs) 
r_HS1  <- 0.15  # constant annual rate of becoming Sick when Healthy
r_S1H  <- 0.5   # constant annual rate of becoming Healthy when Sick
r_S1S2 <- 0.105 # constant annual rate of becoming Sicker when Sick
hr_S1  <- 3     # hazard ratio of death in Sick vs Healthy 
hr_S2  <- 10    # hazard ratio of death in Sicker vs Healthy 

### Effectiveness of treatment AB 
hr_S1S2_trtAB <- 0.6  # hazard ratio of becoming Sicker when Sick under treatment AB

## Age-dependent mortality rates 
lt_usa_2015 <- read.csv("HMD_USA_Mx_2015.csv")
# Extract age-specific all-cause mortality for ages in model time horizon
v_r_mort_by_age <- lt_usa_2015 %>% 
  dplyr::filter(Age >= n_age_init & Age < n_age_max) %>%
  dplyr::select(Total) %>%
  as.matrix()

### State rewards 
#### Costs 
c_H     <- 2000  # annual cost of being Healthy
c_S1    <- 4000  # annual cost of being Sick
c_S2    <- 15000 # annual cost of being Sicker
c_D     <- 0     # annual cost of being dead
c_trtAB <- 25000 # annual cost of receiving treatment AB
#### Utilities 
u_H     <- 1     # annual utility of being Healthy
u_S1    <- 0.75  # annual utility of being Sick
u_S2    <- 0.5   # annual utility of being Sicker
u_D     <- 0     # annual utility of being dead
u_trtAB <- 0.95  # annual utility when receiving treatment AB

### Transition rewards 
du_HS1  <- 0.01  # disutility when transitioning from Healthy to Sick
ic_HS1  <- 1000  # increase in cost when transitioning from Healthy to Sick
ic_D    <- 2000  # increase in cost when dying

### Discount weight for costs and effects 
v_dwc   <- 1 / ((1 + (d_e * cycle_length)) ^ (0:n_cycles))
v_dwe   <- 1 / ((1 + (d_c * cycle_length)) ^ (0:n_cycles))

# Process model inputs 
## Age-specific transition rates to the Dead state for all cycles 
v_r_HDage  <- rep(v_r_mort_by_age, each = 1/cycle_length)
# Name age-specific mortality vector 
names(v_r_HDage) <- v_age_names

# compute mortality rates
v_r_S1Dage  <- v_r_HDage * hr_S1 # Age-specific mortality rate in the Sick state 
v_r_S2Dage  <- v_r_HDage * hr_S2 # Age-specific mortality rate in the Sicker state 
# transform rates to probabilities adjusting by cycle length
p_HS1       <- rate_to_prob(r = r_HS1,  t = cycle_length) # constant annual probability of becoming Sick when Healthy conditional on surviving 
p_S1H       <- rate_to_prob(r = r_S1H,  t = cycle_length) # constant annual probability of becoming Healthy when Sick conditional on surviving
p_S1S2      <- rate_to_prob(r = r_S1S2, t = cycle_length) # constant annual probability of becoming Sicker when Sick conditional on surviving
v_p_HDage   <- rate_to_prob(v_r_HDage,  t = cycle_length) # Age-specific mortality risk in the Healthy state 
v_p_S1Dage  <- rate_to_prob(v_r_S1Dage, t = cycle_length) # Age-specific mortality risk in the Sick state
v_p_S2Dage  <- rate_to_prob(v_r_S2Dage, t = cycle_length) # Age-specific mortality risk in the Sicker state

## Annual transition probability of becoming Sicker when Sick for treatment AB 
# Apply hazard ratio to rate to obtain transition rate of becoming Sicker when Sick for treatment AB
r_S1S2_trtAB <- r_S1S2 * hr_S1S2_trtAB
# Transform rate to probability to become Sicker when Sick under treatment AB 
# adjusting by cycle length conditional on surviving
p_S1S2_trtAB <- rate_to_prob(r = r_S1S2_trtAB, t = cycle_length)
```

# 04 Construct state-transition models

## 04.1 Initial state vector

```{r}
# All starting healthy
v_m_init <- c(H = 1, S1 = 0, S2 = 0, D = 0) # initial state vector
v_m_init
```

## 04.2 Initialize cohort traces

```{r}
### Initialize cohort trace under SoC 
m_M_SoC <- matrix(NA, 
              nrow = (n_cycles + 1), ncol = n_states, 
              dimnames = list(0:n_cycles, v_names_states))
# Store the initial state vector in the first row of the cohort trace
m_M_SoC[1, ] <- v_m_init

### Initialize cohort trace for strategy AB 
# Structure and initial states are the same as for SoC
m_M_strAB <- m_M_SoC # Strategy AB
```

## 04.3 Create transition probability matrices

```{r}
## Create transition probability arrays for strategy SoC 
### Initialize transition probability array for strategy SoC 
# All transitions to a non-death state are assumed to be conditional on survival
a_P_SoC <- array(0,
                 dim  = c(n_states, n_states, n_cycles),
                 dimnames = list(v_names_states, 
                                 v_names_states, 
                                 0:(n_cycles - 1)))
### Fill in array
## From H
# your turn

## From S1
# your turn

## From S2
# your turn

## From D
# your turn


### Initialize transition probability array for strategy AB 
# your turn

# Update only transition probabilities from S1 involving p_S1S2
# your turn

## Check if transition probability arrays are valid 
### Check that transition probabilities are [0, 1] 
check_transition_probability(a_P_SoC,   verbose = TRUE)
check_transition_probability(a_P_strAB, verbose = TRUE)
### Check that all rows for each slice of the array sum to 1 
check_sum_of_transition_array(a_P_SoC,   n_states = n_states, n_cycles = n_cycles, verbose = TRUE)
check_sum_of_transition_array(a_P_strAB, n_states = n_states, n_cycles = n_cycles, verbose = TRUE)
```

## 04.4 Create transition dynamics arrays

```{r}
# These arrays will capture transitions from each state to another over time 
### Initialize transition dynamics array for strategy SoC 
# your turn

# Set first slice of a_A_SoC with the initial state vector in its diagonal
# your turn

### Initialize transition-dynamics array for strategy AB 
# Structure and initial states are the same as for SoC
# your turn

```

# 05 Run Markov model

```{r}
# Iterative solution of age-dependent cSTM
# your turn

## Store the cohort traces in a list 
# your turn

## Store the transition dynamics array for each strategy in a list 
# your turn
```

# 06 Plot Outputs

## 06.1 Plot the cohort trace for strategy SoC 

```{r}
# your turn
```

# 07 State Rewards 

```{r}
## Scale by the cycle length 
# Vector of state utilities under strategy SoC
# your turn

# Vector of state costs under strategy SoC
# your turn

# Vector of state utilities under strategy AB
# your turn

# Vector of state costs under strategy AB
# your turn

## Store state rewards 
# Store the vectors of state utilities for each strategy in a list 
# your turn

# Store the vectors of state cost for each strategy in a list 
# your turn

# assign strategy names to matching items in the lists
# your turn

```

# 08 Compute expected outcomes 

```{r}
# Create empty vectors to store total utilities and costs 
# your turn

## Loop through each strategy and calculate total utilities and costs 
# your turn

```

# 09 Cost-effectiveness analysis (CEA) 

```{r}
## Incremental cost-effectiveness ratios (ICERs) 
# your turn

```

```{r}
## CEA table in proper format 
# your turn

```

```{r}
## CEA frontier 
# your turn

```

=======
---
title: 'Markov Sick-Sicker model in R'
subtitle: "with age-dependency"
author: "The DARTH workgroup"
output:
  html_document: default
  pdf_document: default
---

Developed by the Decision Analysis in R for Technologies in Health (DARTH) workgroup:

Fernando Alarid-Escudero, PhD (1) 

Eva A. Enns, MS, PhD (2)	

M.G. Myriam Hunink, MD, PhD (3,4)

Hawre J. Jalal, MD, PhD (5) 

Eline M. Krijkamp, MSc (3)	

Petros Pechlivanoglou, PhD (6,7)

Alan Yang, MSc (7)

In collaboration of: 		

1. Division of Public Administration, Center for Research and Teaching in 
Economics (CIDE), Aguascalientes, Mexico
2. University of Minnesota School of Public Health, Minneapolis, MN, USA
3. Erasmus MC, Rotterdam, The Netherlands
4. Harvard T.H. Chan School of Public Health, Boston, USA
5. University of Pittsburgh Graduate School of Public Health, Pittsburgh, PA, USA
6. University of Toronto, Toronto ON, Canada
7. The Hospital for Sick Children, Toronto ON, Canada

Please cite our publications when using this code:

- Alarid-Escudero F, Krijkamp EM, Enns EA, Yang A, Hunink MGM Pechlivanoglou P,
Jalal H. A Tutorial on Time-Dependent Cohort State-Transition Models in R using 
a Cost-Effectiveness Analysis Example. Medical Decision Making, 2022 (In press). https://arxiv.org/abs/2108.13552

- Alarid-Escudero F, Krijkamp EM, Enns EA, Yang A, Hunink MGM Pechlivanoglou P,
Jalal H. An Introductory Tutorial on Cohort State-Transition Models in R Using a 
Cost-Effectiveness Analysis Example. Medical Decision Making, 2022 
(Online First):1-18. https://doi.org/10.1177/0272989X221103163

- Jalal H, Pechlivanoglou P, Krijkamp E, Alarid-Escudero F, Enns E, Hunink MG. 
An Overview of R in Health Decision Sciences. Med Decis Making. 2017; 37(3): 735-746. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X16686559
 
- Krijkamp EM, Alarid-Escudero F, Enns EA, Jalal HJ, Hunink MGM, Pechlivanoglou P. 
Microsimulation modeling for health decision sciences using R: A tutorial. 
Med Decis Making. 2018;38(3):400–22. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X18754513
 
- Krijkamp EM, Alarid-Escudero F, Enns E, Pechlivanoglou P, Hunink MM, Jalal H. 
A Multidimensional Array Representation of State-Transition Model Dynamics. 
Med Decis Mak. 2020;40(2):242-248. https://doi.org/10.1177/0272989X19893973

Copyright 2017, THE HOSPITAL FOR SICK CHILDREN AND THE COLLABORATING INSTITUTIONS. 
All rights reserved in Canada, the United States and worldwide. Copyright, 
trademarks, trade names and any and all associated intellectual property are 
exclusively owned by THE HOSPITAL FOR Sick CHILDREN and the collaborating 
institutions. These materials may be used, reproduced, modified, distributed 
and adapted with proper attribution.

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

Change `eval` to `TRUE` if you want to knit this document.

```{r}
rm(list = ls())      # clear memory (removes all the variables from the workspace)
```

# 01 Load packages

```{r, warning = FALSE, message = FALSE}
if (!require('pacman')) install.packages('pacman'); library(pacman) # use this package to conveniently install other packages
# load (install if required) packages from CRAN
p_load("dplyr", "devtools", "scales", "ellipse", "ggplot2", "lazyeval", "igraph", "truncnorm", "ggraph", "reshape2", "knitr", "dampack")                                               
# load (install if required) packages from GitHub
# install_github("DARTH-git/darthtools", force = TRUE) Uncomment if there is a newer version
p_load_gh("DARTH-git/darthtools")
```

# 02 Load functions

```{r}
# all functions are in the darthtools package
```

# 03 Input model parameters

```{r}
# Strategy names
v_names_str <- c("No Treatment", "Treatment") 

# Markov model parameters
age      <- 25                       # age at baseline
max_age  <- 55                       # maximum age of follow up
n_cycles <- max_age - age            # time horizon, number of cycles
v_names_states <- c("H", "S1", "S2", "D")  # the 4 states of the model: Healthy (H), 
                                     # Sick (S1), Sicker (S2), Dead (D)
v_init   <- c("H"  = 1,
              "S1" = 0,              # initial cohort distribution (everyone
              "S2" = 0,              # allocated to the "healthy" state)
              "D"  = 0)                

# Transition probabilities (per cycle) and hazard ratios
# Read age-specific mortality rates from csv file for the age 25 - 54
lt_usa_2015 <- read.csv("HMD_USA_Mx_2015.csv")
v_r_HD <- lt_usa_2015 %>% 
  filter(Age >= age & Age <= (max_age - 1)) %>%
  select(Total) %>%
  as.matrix()

# Transition probabilities (per cycle)
v_p_HD   <- rate_to_prob(r = v_r_HD, t = 1)  # probability to die when healthy (vector)
p_HS1    <- 0.15          	                 # probability to become sick when healthy, conditional on surviving
p_S1H    <- 0.5           	                 # probability to become healthy when sick, conditional on surviving
p_S1S2   <- 0.105         	                 # probability to become sicker when sick, conditional on surviving
hr_S1    <- 3             	                 # hazard ratio of death in sick vs healthy
hr_S2    <- 10            	                 # hazard ratio of death in sicker vs healthy 
v_r_S1D  <- hr_S1 * v_r_HD  	               # rate of death in sick (vector)
v_r_S2D  <- hr_S2 * v_r_HD  	               # rate of death in sicker (vector)
v_p_S1D  <- rate_to_prob(r = v_r_S1D, t = 1) # probability to die in sick (vector)
v_p_S2D  <- rate_to_prob(r = v_r_S2D, t = 1) # probability to die in sicker (vector)

# Cost and utility inputs 
c_H      <- 2000                             # cost of remaining one cycle in the healthy state
c_S1     <- 4000                             # cost of remaining one cycle in the sick state
c_S2     <- 15000                            # cost of remaining one cycle in the sicker state
c_trt    <- 12000                            # cost of treatment (per cycle)
c_D      <- 0                                # cost of being in the death state
u_H      <- 1                                # utility when healthy
u_S1     <- 0.75                             # utility when sick
u_S2     <- 0.5                              # utility when sicker
u_D      <- 0                                # utility when dead
u_trt    <- 0.95                             # utility when being treated
d_e      <- d_c <- 0.03                      # discount rate per cycle equal discount of costs and QALYs by 3%

n_str    <- length(v_names_str)              # Number of strategies
n_states <- length(v_names_states)           # number of states

# Discount weights for costs and effects
v_dwc    <- 1 / (1 + d_c) ^ (0:n_cycles) 
v_dwe    <- 1 / (1 + d_e) ^ (0:n_cycles) 
```

# 04 Define and initialize matrices and vectors

## 04.1 Cohort trace

```{r}
# create the markov trace matrix M capturing the proportion of the cohort in each state 
# at each cycle
m_M_notrt <- m_M_trt <- matrix(NA, 
                               nrow     = n_cycles + 1, ncol = n_states,
                               dimnames = list(paste("cycle", 0:n_cycles, sep = " "), v_names_states))

head(m_M_notrt) # show first 6 rows of the matrix 

# The cohort starts as healthy
m_M_notrt[1, ] <- m_M_trt[1, ] <- v_init  # initiate first cycle of cohort trace 
```

## 04.2 Transition probability array

```{r}
# create the transition probability array for NO treatment
a_P_notrt <- array(0,  # Create 3-D array
                   dim = c(n_states, n_states, n_cycles),
                   dimnames = list(v_names_states, v_names_states, 0:(n_cycles - 1))) # name dimensions
a_P_notrt[, , 1:3]
```

Fill in the transition probability array:

```{r}
# from Healthy

# from Sick

# from Sicker

# from Dead

# create transition probability array for treatment same as no treatment
a_P_trt <- a_P_notrt  

# In the Sick-Sicker example the transitions are similar. Therefore, you can use the same values in the transition probability array. If you are modeling other strategies you have to update the transition probabilities accordingly
```

# 05 Run Markov model

```{r}
# your turn

```

# 06 Compute and Plot Epidemiological Outcomes

## 06.1 Cohort trace

```{r}
# your turn

```

## 06.2 Overall Survival (OS)

```{r}
# your turn

```

## 06.2.1 Life Expectancy (LE)

```{r}
# your turn

```

## 06.3 Disease prevalence

```{r}
# your turn

```

## 06.4 ratio of sick(S1) vs sicker(S2)

```{r}
# your turn

```

# 07 Compute Cost-Effectiveness Outcomes

## 07.1 Mean Costs and QALYs for Treatment and NO Treatment

```{r}
# your turn

```

## 07.2 Discounted Mean Costs and QALYs

```{r}
# your turn

# Store discounted costs and effectiveness for each strategy in a data frame
df_ce       <- data.frame(Strategy = NA, # <-- your value (strategy names)
                          Cost     = NA, # <-- your value (strategy costs)
                          Effect   = NA  # <-- your value (strategy QALYs)
                          )
```

## 07.3 Compute ICERs of the Markov model

```{r}
# Use the function calculate_icers() from the dampack package
df_cea <- calculate_icers(cost       = df_ce$Cost,
                          effect     = df_ce$Effect,
                          strategies = df_ce$Strategy)
df_cea 
```

## 07.4 Plot frontier of the Markov model

```{r}
plot(df_cea, effect_units = "QALYs")
```

>>>>>>> da36a6af5b5df0f53272afa0e73089f8abde74fa
