---
title: 'PSA: Three-strategy decision tree in R - HVE'
author: "The DARTH workgroup"
output:
  pdf_document: default
  html_document: default
keep_tex: yes
self_contained: no
---

Developed by the Decision Analysis in R for Technologies in Health (DARTH) workgroup:

Fernando Alarid-Escudero, PhD (1) 

Eva A. Enns, MS, PhD (2)	

M.G. Myriam Hunink, MD, PhD (3,4)

Hawre J. Jalal, MD, PhD (5) 

Eline M. Krijkamp, MSc (3)	

Petros Pechlivanoglou, PhD (6)

Alan Yang, MSc (7)

In collaboration of: 		

1. Drug Policy Program, Center for Research and Teaching in Economics (CIDE) - CONACyT, 
   Aguascalientes, Mexico
2. University of Minnesota School of Public Health, Minneapolis, MN, USA
3. Erasmus MC, Rotterdam, The Netherlands
4. Harvard T.H. Chan School of Public Health, Boston, USA
5. University of Pittsburgh Graduate School of Public Health, Pittsburgh, PA, USA
6. The Hospital for Sick Children, Toronto and University of Toronto, Toronto ON, Canada
7. The Hospital for Sick Children, Toronto ON, Canada

Please cite our publications when using this code:
 
- Jalal H, Pechlivanoglou P, Krijkamp E, Alarid-Escudero F, Enns E, Hunink MG. 
An Overview of R in Health Decision Sciences. Med Decis Making. 2017; 37(3): 735-746. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X16686559
 
- Krijkamp EM, Alarid-Escudero F, Enns EA, Jalal HJ, Hunink MGM, Pechlivanoglou P. 
Microsimulation modeling for health decision sciences using R: A tutorial. 
Med Decis Making. 2018;38(3):400–22. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X18754513
 
- Krijkamp EM, Alarid-Escudero F, Enns E, Pechlivanoglou P, Hunink MM, Jalal H. 
A Multidimensional Array Representation of State-Transition Model Dynamics. 
BioRxiv 670612 2019.https://www.biorxiv.org/content/10.1101/670612v1

Copyright 2017, THE HOSPITAL FOR SICK CHILDREN AND THE COLLABORATING INSTITUTIONS. 
All rights reserved in Canada, the United States and worldwide. Copyright, 
trademarks, trade names and any and all associated intellectual property are 
exclusively owned by THE HOSPITAL FOR Sick CHILDREN and the collaborating 
institutions. These materials may be used, reproduced, modified, distributed 
and adapted with proper attribution.

\newpage

```{r}
rm(list = ls())      # clear memory (removes all the variables from the workspace)
```

# 01 Load packages

```{r, warning=F, message=F}
# Uncomment if you don't have all required packages installed
# source("app0_packages_setup.R", echo = TRUE) 
library(dplyr)     # For data manipulation
library(dectree)   # To plot tree
library(lazyeval)
library(igraph)
library(ggplot2)
library(ggraph)
library(stringr)
library(ellipse)
library(reshape2)
library(truncnorm)
library(dampack)   # for CEA
```

# 02 Load functions

```{r}
source(here::here('functions','Functions.R'))
```

# 03 Define parameter input values

```{r}
v_names_str   <- c("No Tx", "Tx All", "Biopsy")    # names of strategies
n_str         <- length(v_names_str)               # number of strategies
wtp           <- 100000                            # willingness to pay threshold

# Probabilities
p_HVE         <- 0.52   # prevalence of HVE
p_HVE_comp    <- 0.71   # complications with untreated HVE
p_OVE_comp    <- 0.01   # complications with untreated OVE
p_HVE_comp_tx <- 0.36   # complications with treated HVE
p_OVE_comp_tx <- 0.20   # complications with treated OVE
p_biopsy_comp <- 0.05   # probability of complications due to biopsy

# Costs
c_VE          <- 1200   # cost of viral encephalitis care without complications
c_VE_comp     <- 9000   # cost of viral encephalitis care with complications
c_tx          <- 9500   # cost of treatment
c_biopsy      <- 25000  # cost of brain biopsy

# QALYs
q_VE          <- 20     # remaining QALYs for those without VE-related complications
q_VE_comp     <- 19     # remaining QALYs for those with VE-related complications
q_loss_biopsy <- -0.01  # one-time QALY loss due to brain biopsy

# store the parameters into a list
l_params_all <- as.list(data.frame(p_HVE, p_HVE_comp, p_OVE_comp, p_HVE_comp_tx, p_OVE_comp_tx, p_biopsy_comp, 
                                   c_VE, c_VE_comp, c_tx, c_biopsy, 
                                   q_VE, q_VE_comp, q_loss_biopsy))
```

# 04 Create and run decision tree model

```{r}
decision_tree_HVE_output <- with(as.list(l_params_all), {
  
  # Create vector of weights for each strategy 
  
  v_w_no_tx  <- c(  p_HVE  *    p_HVE_comp  ,  # HVE, complications
                    p_HVE  * (1-p_HVE_comp) ,  # HVE, no complications
                 (1-p_HVE) *    p_OVE_comp  ,  # OVE, complications
                 (1-p_HVE) * (1-p_OVE_comp))   # OVE, no complications
  
  v_w_tx     <- c(  1                          ,  # On treatment
                    p_HVE  *    p_HVE_comp_tx  ,  # HVE w/tx, complications
                    p_HVE  * (1-p_HVE_comp_tx) ,  # HVE w/tx, no complications
                 (1-p_HVE) *    p_OVE_comp_tx  ,  # OVE w/tx, complications
                 (1-p_HVE) * (1-p_OVE_comp_tx))   # OVE w/tx, no complications
  
  v_w_biopsy <- c(  1                                              , # Undergo biopsy
                    p_biopsy_comp                                  , # biopsy complications
                 # no biopsy comp., HVE w/tx,  complications
                 (1-p_biopsy_comp) *    p_HVE  *    p_HVE_comp_tx  ,  
                 # no biopsy comp., HVE w/tx, no complications
                 (1-p_biopsy_comp) *    p_HVE  * (1-p_HVE_comp_tx) ,  
                 # no biopsy comp., OVE, complications
                 (1-p_biopsy_comp) * (1-p_HVE) *    p_OVE_comp     ,  
                 # no biopsy comp., OVE, no complications
                 (1-p_biopsy_comp) * (1-p_HVE) * (1-p_OVE_comp))      
  
  # Create vector of outcomes (QALYs) for each strategy 
  
  v_qaly_no_tx  <- c(q_VE_comp ,  # HVE, complications
                     q_VE      ,  # HVE, no complications
                     q_VE_comp ,  # OVE, complications
                     q_VE)        # OVE, no complications
  
  v_qaly_tx     <- c(0      ,     # treatment does not directly add any QALYs 
                     q_VE_comp ,  # HVE, complications
                     q_VE      ,  # HVE, no complications
                     q_VE_comp ,  # OVE, complications
                     q_VE)        # OVE, no complications
  
  
  v_qaly_biopsy <- c(q_loss_biopsy  ,  # loss due to biopsy
                     q_VE_comp      ,  # biopsy complications
                     q_VE_comp      ,  # no biopsy comp., HVE w/tx, complications 
                     q_VE           ,  # no biopsy comp., HVE w/tx, no complications
                     q_VE_comp      ,  # no biopsy comp., OVE, complications
                     q_VE)             # no biopsy comp., OVE, no complications
  
  # Create vector of costs for each strategy 
  
  v_cost_no_tx  <- c(c_VE_comp ,  # HVE, complications
                     c_VE      ,  # HVE, no complications
                     c_VE_comp ,  # OVE, complications
                     c_VE)        # OVE, no complications
  
  v_cost_tx     <- c(c_tx      ,  # cost of treatment
                     c_VE_comp ,  # HVE, complications
                     c_VE      ,  # HVE, no complications
                     c_VE_comp ,  # OVE, complications
                     c_VE)        # OVE, no complications
  
  
  v_cost_biopsy <- c(c_biopsy         ,  # cost of biopsy procedure
                     c_VE_comp        ,  # biopsy complications
                     c_VE_comp + c_tx ,  # no biopsy comp., HVE w/tx, complications 
                     c_VE + c_tx      ,  # no biopsy comp., HVE w/tx, no complications
                     c_VE_comp        ,  # no biopsy comp., OVE, complications
                     c_VE)               # no biopsy comp., OVE, no complications
  
  # Calculate total utilities for each strategy 
  total_qaly_no_tx  <- v_w_no_tx  %*%  v_qaly_no_tx      
  total_qaly_tx     <- v_w_tx     %*%  v_qaly_tx
  total_qaly_biopsy <- v_w_biopsy %*%  v_qaly_biopsy
  
  # Calculate total costs for each strategy 
  total_cost_no_tx  <- v_w_no_tx  %*%  v_cost_no_tx    
  total_cost_tx     <- v_w_tx     %*%  v_cost_tx
  total_cost_biopsy <- v_w_biopsy %*%  v_cost_biopsy
  
  # vector of total QALYs
  v_total_qaly <- c(total_qaly_no_tx, total_qaly_tx, total_qaly_biopsy) 
  # vector of total costs
  v_total_cost <- c(total_cost_no_tx, total_cost_tx, total_cost_biopsy) 
  # calculate vector of nmb
  v_nmb        <- v_total_qaly * wtp - v_total_cost                      
  
  # Name outcomes
  names(v_total_qaly) <- v_names_str  # names for the elements of the total QALYs vector
  names(v_total_cost) <- v_names_str  # names for the elements of the total cost vector
  names(v_nmb)        <- v_names_str  # names for the elements of the nmb vector
  
  df_output <- data.frame(Strategy =  v_names_str,
                          Cost     =  v_total_cost,
                          Effect   =  v_total_qaly,
                          NMB      =  v_nmb)
  return(df_output)
})

# model output
decision_tree_HVE_output
```

## 04.1 Plot the decision tree

```{r, out.width='100%'}
branches <- read.csv(here::here('data','decision_tree_HVE_branches.csv'), stringsAsFactors = F, header = T)
tree     <- create_tree(branches)
plot_tree(tree, font.size = 5)
```

## 05 Cost-Effectiveness Analysis

```{r}
# create the transition probability matrix for NO treatment
decision_tree_HVE_cea  <- calculate_icers(cost       = decision_tree_HVE_output$Cost,
                                          effect     = decision_tree_HVE_output$Effect,
                                          strategies = decision_tree_HVE_output$Strategy) #,
                                        # ref_strat  = 'Tx All')
decision_tree_HVE_cea
```

## 05.1 Plot frontier of Decision Tree

```{r}
plot(decision_tree_HVE_cea, effect_units = "QALYs")
```

# 06 Deterministic Sensitivity Analysis

## 06.1 List of input parameters

```{r}
l_params_all 
```

## 06.2 Load decision tree model function

```{r}
source(here::here("functions", "Functions_decision_tree_HVE.R"))
# Test function
calculate_ce_out(l_params_all)
```

## 06.3 One-way sensitivity analysis (OWSA)

```{r}
owsa_nmb <- owsa_det(parms = c("p_HVE", "p_biopsy_comp", "c_tx", "c_biopsy"), 
                     ranges = list("p_HVE"         = c(0.01, 0.99),
                                   "p_biopsy_comp" = c(0.01, 0.50),
                                   "c_tx"          = c(1000, 15000),
                                   "c_biopsy"      = c(10000, 40000)),
                     nsamps = 100, # number of values  
                     FUN = calculate_ce_out, # Function to compute outputs 
                     params_basecase = l_params_all, # List with base-case parameters
                     outcome = "NMB",   #"Effect"   # Output to do the OWSA on
                     strategies = v_names_str, # Names of strategies
                     n_wtp = 450000        # Extra argument to pass to FUN
)
```

## 06.3.1 Plot OWSA

```{r}
plot(owsa_nmb, txtsize = 16, n_x_ticks = 5, 
     facet_scales = "free") +
  theme(legend.position = "bottom")
```

## 06.3.2 Optimal strategy with OWSA

```{r}
owsa_opt_strat(owsa = owsa_nmb)
```

## 06.3.3 Tornado plot

```{r}
owsa_tornado(owsa = owsa_nmb, strategy = c("Biopsy"))
```

## 06.4 Two-way sensitivity analysis (TWSA)

```{r}
twsa_nmb <- twsa_det(parm1  = "p_HVE", # parameter 1 name
                     parm2  = "c_tx",  # parameter 2 name
                     ranges = list("p_HVE" = c(0.01, 0.99),
                                   "c_tx"  = c(10000, 40000)),
                     nsamps = 40, # number of values  
                     FUN = calculate_ce_out, # Function to compute outputs 
                     params_basecase = l_params_all, # Vector with base-case parameters
                     outcome = "NMB",      # Output to do the OWSA on
                     strategies = v_names_str, # Names of strategies
                     n_wtp = 450000        # Extra argument to pass to FUN
)
```

## 06.4.1 Plot TWSA

```{r}
plot(twsa_nmb) + 
  ggtitle(label = "Two-way sensitivity analysis",subtitle = "Net monetary benefit") +
  theme(legend.position = "bottom")
```

# 07 Probabilistic Sensitivity Analysis (PSA)

```{r}
# Function to generate PSA input dataset
generate_psa_params <- function(n_sim = 1000, seed = 071518){
  set.seed(seed)
  # Dataframe of input parameters
  df_psa_params   <- data.frame(
    
    # Transition probabilities (per cycle)
    p_HVE         = rbeta(n_sim, 52, 48),  # prevalence of HVE
    p_HVE_comp    = rbeta(n_sim, 71, 29),  # complications with untreated HVE
    p_OVE_comp    = rbeta(n_sim,  1, 99),  # complications with untreated OVE
    p_HVE_comp_tx = rbeta(n_sim, 36, 64),  # complications with treated HVE
    p_OVE_comp_tx = rbeta(n_sim, 20, 80),  # complications with treated OVE
    p_biopsy_comp = rbeta(n_sim,  1, 19),  # probability of complications due to biopsy
    
    # Costs
    c_VE          = rgamma(n_sim, shape = 36, scale = 33.33),    # cost of remaining one cycle in state H
    c_VE_comp     = rgamma(n_sim, shape = 81, scale = 111.1),    # cost of remaining one cycle in state S1
    c_tx          = rgamma(n_sim, shape = 74.6, scale = 127.4),  # cost of remaining one cycle in state S2
    c_biopsy      = rgamma(n_sim, shape = 25, scale = 1000),     # cost of treatment (per cycle)
    
    # Utilities
    q_VE          = rnorm(n_sim, mean = 20, sd = 1),  # utility when healthy
    q_VE_comp     = rnorm(n_sim, mean = 19, sd = 2),  # utility when sick
    q_loss_biopsy = rtruncnorm(n_sim, mean = -0.01, sd = 0.005, a = -0.1, b = 0)
  )
  return(df_psa_params)
}
# Try it
generate_psa_params(10) 

# Number of simulations
n_sim <- 1000

# Generate PSA input dataset
df_psa_input     <- generate_psa_params(n_sim = n_sim)
# First six observations
head(df_psa_input)

# Histogram of parameters
ggplot(melt(df_psa_input, variable.name = "Parameter"), aes(x = value)) +
  facet_wrap(~Parameter, scales = "free") +
  geom_histogram(aes(y = ..density..)) +
  theme_bw(base_size = 16)

# Initialize dataframes with PSA output 
# Dataframe of costs
df_c <- as.data.frame(matrix(0, 
                      nrow = n_sim,
                      ncol = n_str))
colnames(df_c) <- v_names_str
# Dataframe of effectiveness
df_e <- as.data.frame(matrix(0, 
                      nrow = n_sim,
                      ncol = n_str))
colnames(df_e) <- v_names_str
```

## 07.1 Conduct probabilistic sensitivity analysis

```{r}
# Run decision tree on each parameter set of PSA input dataset
for(i in 1:n_sim){
  df_out_temp <- calculate_ce_out(df_psa_input[i, ])
  df_c[i, ] <- df_out_temp$Cost
  df_e[i, ] <- df_out_temp$Effect
  # Display simulation progress
  if(i/(n_sim/10) == round(i/(n_sim/10),0)) { # display progress every 10%
    cat('\r', paste(i/n_sim * 100, "% done", sep = " "))
  }
}
```

## 07.2 Create PSA object for dampack

```{r}
l_psa <- make_psa_obj(cost          = df_c, 
                      effectiveness = df_e, 
                      parameters    = df_psa_input, 
                      strategies    = v_names_str)
```

## 07.2.1 Save PSA objects

```{r}
save(df_psa_input, df_c, df_e, v_names_str, n_str,
     l_psa,
     file = here::here("output", "decision_tree_HVE_PSA_dataset.RData"))
```

## 07.3 Create probabilistic analysis graphs

```{r}
load(file = here::here("output", "decision_tree_HVE_PSA_dataset.RData"))
```

Vector with willingness-to-pay (WTP) thresholds.

```{r}
v_wtp <- seq(0, 300000, by = 10000)
```

## 07.3.1 Cost-Effectiveness Scatter plot

```{r}
plot(l_psa)
```

## 07.4 Conduct CEA with probabilistic output

```{r}
# Compute expected costs and effects for each strategy from the PSA
df_out_ce_psa <- summary(l_psa)
df_out_ce_psa

# Calculate incremental cost-effectiveness ratios (ICERs)
df_cea_psa <- calculate_icers(cost       = df_out_ce_psa$meanCost, 
                              effect     = df_out_ce_psa$meanEffect,
                              strategies = df_out_ce_psa$Strategy)
df_cea_psa

# Save CEA table with ICERs
# As .RData
save(df_cea_psa, 
     file = here::here("tables", "decision_tree_HVE_probabilistic_CEA_results.RData"))
# As .csv
write.csv(df_cea_psa, 
          file = here::here("tables", "decision_tree_HVE_probabilistic_CEA_results.csv"))
```

## 07.4.1 Plot cost-effectiveness frontier

```{r}
plot(df_cea_psa)
```

## 07.4.2 Cost-effectiveness acceptability curves (CEACs) and frontier (CEAF)

```{r}
ceac_obj <- ceac(wtp = v_wtp, psa = l_psa)
# Regions of highest probability of cost-effectiveness for each strategy
summary(ceac_obj)
# CEAC & CEAF plot
plot(ceac_obj)
```

## 07.4.3 Expected Loss Curves (ELCs)

```{r}
elc_obj <- calc_exp_loss(wtp = v_wtp, psa = l_psa)
elc_obj
# ELC plot
plot(elc_obj, log_y = FALSE)
```

## 07.4.4 Expected value of perfect information (EVPI)

```{r}
evpi <- calc_evpi(wtp = v_wtp, psa = l_psa)
# EVPI plot
plot(evpi, effect_units = "QALY")
```

