---
date: "`r Sys.Date()`"
draft: false
linktitle: Instructors
slug: instructors
menu:
  prework:
    parent: "Welcome"
    weight: 4
title: "Instructors"
type: docs
---

**Eva Enns, PhD** <a href="https://directory.sph.umn.edu/bio/sph-a-z/eva-enns" target="_blank"><i class="fas fa-link"></i></a> <a href="https://github.com/evaenns" target="_blank"><i class="fab fa-github"></i></a> 

```{r, echo=F}
#blogdown::shortcode("vimeo", "472448727")
```

Eva Enns, PhD, is an Associate Professor at the University of Minnesota’s School of Public Health, in the division of Health Policy and Management. She has a bachelor’s degree in electrical engineering from the Massachusetts Institute of Technology, and holds both a Master’s and a PhD in electrical engineering from Stanford University. She develops mathematical models of the spread of infectious diseases that help evaluate strategies aimed at mitigating disease impact through prevention and/or treatment. Her aim is to inform better infectious disease prevention, management, and control policies as well as to develop new methodologies or tools to better address issues surrounding infectious disease outbreaks.

**Fernando Alarid-Escudero, PhD** <a href="https://profiles.stanford.edu/fernando-alarid-escudero" target="_blank"><i class="fas fa-link"></i></a> <a href="https://github.com/feralaes" target="_blank"><i class="fab fa-github"></i></a> 

```{r, echo=F}
# blogdown::shortcode("vimeo", "473565068")
```

Fernando Alarid-Escudero, Ph.D., is an Assistant Professor of Health Policy at Stanford University School of Medicine. He obtained his Ph.D. in Health Decision Sciences from the University of Minnesota School of Public Health. He was an Assistant Professor at the Center for Research and Teaching in Economics (CIDE) Región Centro, Aguascalientes, Mexico, from 2018 to 2022, prior to coming to Stanford. His research focuses on developing statistical and decision-analytic models to identify optimal prevention, control, and treatment policies to address a wide range of public health problems and develops novel methods to quantify the value of future research. Dr. Alarid-Escudero is part of the Cancer Intervention and Surveillance Modeling Network (CISNET), a consortium of NCI-sponsored investigators that includes modeling to improve our understanding of the impact of cancer control interventions (e.g., prevention, screening, and treatment) on population trends in incidence and mortality. Dr. Alarid-Escudero co-founded the Stanford-CIDE Coronavirus Simulation Modeling (SC-COSMO) workgroup (https://www.sc-cosmo.org). He also co-founded the Decision Analysis in R for Technologies in Health (DARTH) workgroup (http://darthworkgroup.com) and the Collaborative Network on Value of Information (ConVOI; https://www.convoi-group.org), international and multi-institutional collaborative efforts where we develop transparent and open-source solutions to implement decision analysis and quantify the value of potential future investigation for health policy analysis. He received a BSc in Biomedical Engineering from the Metropolitan Autonomous University in Iztapalapa (UAM-I), and a Master’s in Economics from CIDE, both in Mexico.

<!-- **Karen M. Kuntz, ScD** -->

<!-- Karen Kuntz, ScD, is a Professor and Director of Graduate Studies in Health Services Research, Policy and Administration MS and Doctoral programs at the University of Minnesota School of Public Health, Division of Health Policy and Management. She is an internationally-recognized expert in the methods and applications of simulation modeling to address clinical and public health interventions. She is past president of the Society for Medical Decision Making (SMDM) – an organization dedicated to the methods and applications of decision modeling. More recently, she served as co-chair of the International Society for Pharmacoeconomics and Outcomes Research (ISPOR)-SMDM Modeling Good Research Practices Task Force (2010-2012) where she also served as co-leader on the state-transition modeling paper. She was also a member of the 2nd Panel on Cost-Effectiveness in Health and Medicine and served as the lead on the chapter on decision modeling. She is a firm believer in the added value that decision models can bring to real-world policy. -->

**Petros Pechlivanoglou, PhD** <a href="https://ihpme.utoronto.ca/faculty/petros-pechlivanoglou/" target="_blank"><i class="fas fa-link"></i></a> <a href="https://github.com/ppehli" target="_blank"><i class="fab fa-github"></i></a>

```{r, echo=F}
#blogdown::shortcode("vimeo", "472699591")
```

Petros Pechlivanoglou, PhD, is a Senior Scientist at The Hospital for Sick Children’s Research Institute and an Associate Professor at the University of Toronto’s Institute of Health Policy, Management and Evaluation. Dr. Pechlivanoglou’s research interests focus on methods and applications of decision analysis in health economics, bridging evidence synthesis, administrative data and decision analysis, and the application and extension of predictive models in health economics. He has received funding from the Canadian Institute for Health Research, the Canadian Respiratory Research Network, and more recently, the Ontario Early Researcher Award to work on projects around the economics of pre-term birth prevention, the economic evaluation of oncology interventions in children and adolescents, the use of decision analysis in paediatric clinical trial design, and the economic impact of COPD in adults.

**Hawre Jalal, MD, PhD** <a href="https://publichealth.pitt.edu/home/directory/hawre-jalal" target="_blank"><i class="fas fa-link"></i></a> <a href="https://github.com/hjalal" target="_blank"><i class="fab fa-github"></i></a>

```{r, echo=F}
#blogdown::shortcode("vimeo", "473576731")
```

Hawre Jalal, MD, PhD is an Assistant Professor at the Department of Health Policy and Management and a researcher at the Public Health Dynamic Laboratory at the University of Pittsburgh. His research focuses on using mathematical modeling, cost-effective analyses and econometrics approaches to address health policy and resource allocation questions. His current research involves modeling the opioid-epidemic and evaluating its mitigation strategies, and research in disability and aging. Hawre’s methodological interests involve value of information analysis, Bayesian calibration and communicating uncertainty to decision makers.

**Eline Krijkamp, PhD** <a href="https://www.researchgate.net/profile/Eline_Krijkamp2" target="_blank"><i class="fas fa-link"></i></a> <a href="https://github.com/krijkamp" target="_blank"><i class="fab fa-github"></i></a>

```{r, echo=F}
#blogdown::shortcode("vimeo", "473041510")
```

Eline Krijkamp, PhD, is a post-doctoral researcher at the Erasmus School of Health Policy and Management at the Erasmus University in Rotterdam, the Netherlands. In her research she uses decision modeling methods for applied health economic research questions with a little touch on developing new methods. She is also part of an external review group for The National institute for health and care excellence (NICE) in the UK and teaches decision modeling to master and graduate students. She holds a bachelor’s degree in health and biomedical Sciences from the VU University in Amsterdam and two master’s degrees: one in Health Economics from the Erasmus University in Rotterdam and one in Epidemiology from the Netherlands Institute of Health Sciences (NIHES), Rotterdam. After she worked in the educational innovation team for NIHES she pursued her career with a PhD in medical decision making focusing on open-source modeling and resource prioritization in healthcare. Her career goal is to solve health(care) issues using data-driven methods. Simplicity and clear communication are also important to her. She believes that many health issues can be prevented or solved by improved communication and sees communicating our complex research in a simple way as an important challenge. Although Eline loves her professional career, she likes to prioritize sports, good food and enough sleep to stay energized. 

<!-- **Alan Yang, MSc** <a href="https://www.linkedin.com/in/alan-yang-287b8080/" target="_blank"><i class="fas fa-link"></i></a> <a href="https://github.com/alanyang0924" target="_blank"><i class="fab fa-github"></i></a> -->

<!-- ```{r, echo=F} -->
<!-- blogdown::shortcode("vimeo", "472839158") -->
<!-- ``` -->

<!-- Alan Yang holds a Master of Science in Biostatistics and a Honours Bachelor of Science in Statistics from the University of Toronto. He is currently working as a Research Analyst in Child Health Evaluative Sciences at the Hospital for Sick Children Research Institute focusing on implementations of decision analysis in the `R` software and developing statistical methods and applications in health technology assessment. In addition, he is working on Bayesian methods for network meta-analysis and evidence synthesis. Outside of work, he enjoys playing basketball, swimming and trying out different cuisines. -->




