---
date: "2020-10-26T00:00:00-08:00"
draft: false
linktitle: Instructors
slug: local
menu:
  prework:
    parent: "Welcome"
    weight: 3
title: "Instructors"
type: docs
---

**Eline Krijkamp, MSc** <a href="https://www.researchgate.net/profile/Eline_Krijkamp2" target="_blank"><i class="fas fa-link"></i></a> <a href="https://github.com/krijkamp" target="_blank"><i class="fab fa-github"></i></a> 

```{r, echo=F}
blogdown::shortcode("vimeo", "473041510")
```

Eline Krijkamp, MSc, has a paradoxical relationship with choice: she doesn’t like having to make choices and yet, she is very enthusiastic about modelling choices in healthcare. As a Master’s student, she joined the DARTH crew and focused on how to build microsimulation models in R. She holds a Master’s in Epidemiology from the Netherlands Institute of Health Sciences and a Master’s in Health Economics from Erasmus University Rotterdam. She is currently a PhD student in epidemiology at the Netherlands Institue of Health Sciences. Right now, she is building a decision model to predict the best choice for her future career path.

**Petros Pechlivanoglou, PhD** <a href="https://ihpme.utoronto.ca/faculty/petros-pechlivanoglou/" target="_blank"><i class="fas fa-link"></i></a> <a href="https://github.com/ppehli" target="_blank"><i class="fab fa-github"></i></a> 

```{r, echo=F}
blogdown::shortcode("vimeo", "472699591")
```

Petros Pechlivanoglou, PhD, is a Scientist at the Hospital for Sick Children’s Research Institute and an Assistant Professor at the University of Toronto’s Institute of Health Policy, Management and Evaluation. After studying economics in his native Greece, he received both a Master’s in econometrics and a PhD in health econometrics from the University of Groningen. His research interests include: the use of health decision analysis in economic evaluation; bridging evidence synthesis, administrative data and decision analysis; and the application and extension of predictive models in health economics. In his spare time, he enjoys playing and performing traditional Greek music and explaining to people how to pronounce his last name.

**Alan Yang, MSc** <a href="https://www.linkedin.com/in/alan-yang-287b8080/" target="_blank"><i class="fas fa-link"></i></a> <a href="https://github.com/alanyang0924" target="_blank"><i class="fab fa-github"></i></a>

```{r, echo=F}
blogdown::shortcode("vimeo", "472839158")
```

Alan Yang holds a Master of Science in Biostatistics and a Honours Bachelor of Science in Statistics from the University of Toronto. He is currently working as a Research Analyst in Child Health Evaluative Sciences at the Hospital for Sick Children Research Institute focusing on implementations of decision analysis in the `R` software and developing statistical methods and applications in health technology assessment. In addition, he is working on Bayesian methods for network meta-analysis and evidence synthesis. Outside of work, he enjoys playing basketball, swimming and trying out different cuisines.
