---
date: "2020-10-26T00:00:00-08:00"
draft: false
linktitle: Instructors
slug: local
menu:
  prework:
    parent: "Welcome"
    weight: 3
title: "Instructors"
type: docs
---

**Eline Krijkamp, MSc** <a href="https://www.researchgate.net/profile/Eline_Krijkamp2" target="_blank"><i class="fas fa-link"></i></a> <a href="https://github.com/krijkamp" target="_blank"><i class="fab fa-github"></i></a> 

```{r, echo=F}
blogdown::shortcode("vimeo", "473041510")
```

Eline Krijkamp, MSc, has a paradoxical relationship with choice: she doesn’t like having to make choices and yet, she is very enthusiastic about modelling choices in healthcare. As a Master’s student, she joined the DARTH crew and focused on how to build microsimulation models in R. She holds a Master’s in Epidemiology from the Netherlands Institute of Health Sciences and a Master’s in Health Economics from Erasmus University Rotterdam. She is currently a PhD student in epidemiology at the Netherlands Institue of Health Sciences. Right now, she is building a decision model to predict the best choice for her future career path.

**Petros Pechlivanoglou, PhD** <a href="https://ihpme.utoronto.ca/faculty/petros-pechlivanoglou/" target="_blank"><i class="fas fa-link"></i></a> <a href="https://github.com/ppehli" target="_blank"><i class="fab fa-github"></i></a> 

```{r, echo=F}
blogdown::shortcode("vimeo", "472699591")
```

Petros Pechlivanoglou, PhD, is a Scientist at the Hospital for Sick Children’s Research Institute and an Assistant Professor at the University of Toronto’s Institute of Health Policy, Management and Evaluation. After studying economics in his native Greece, he received both a Master’s in econometrics and a PhD in health econometrics from the University of Groningen. His research interests include: the use of health decision analysis in economic evaluation; bridging evidence synthesis, administrative data and decision analysis; and the application and extension of predictive models in health economics. In his spare time, he enjoys playing and performing traditional Greek music and explaining to people how to pronounce his last name.

**Alan Yang, MSc** <a href="https://www.linkedin.com/in/alan-yang-287b8080/" target="_blank"><i class="fas fa-link"></i></a> <a href="https://github.com/alanyang0924" target="_blank"><i class="fab fa-github"></i></a>

```{r, echo=F}
blogdown::shortcode("vimeo", "472839158")
```

Alan Yang holds a Master of Science in Biostatistics and a Honours Bachelor of Science in Statistics from the University of Toronto. He is currently working as a Research Analyst in Child Health Evaluative Sciences at the Hospital for Sick Children Research Institute focusing on implementations of decision analysis in the `R` software and developing statistical methods and applications in health technology assessment. In addition, he is working on Bayesian methods for network meta-analysis and evidence synthesis. Outside of work, he enjoys playing basketball, swimming and trying out different cuisines.

**Eva Enns, PhD** <a href="https://directory.sph.umn.edu/bio/sph-a-z/eva-enns" target="_blank"><i class="fas fa-link"></i></a> <a href="https://github.com/evaenns" target="_blank"><i class="fab fa-github"></i></a> 

```{r, echo=F}
blogdown::shortcode("vimeo", "472448727")
```

Eva Enns, PhD, is an Associate Professor at the University of Minnesota’s School of Public Health, in the division of Health Policy and Management. She has a bachelor’s degree in electrical engineering from the Massachusetts Institute of Technology, and holds both a Master’s and a PhD in electrical engineering from Stanford University. She develops mathematical models of the spread of infectious diseases that help evaluate strategies aimed at mitigating disease impact through prevention and/or treatment. Her aim is to inform better infectious disease prevention, management, and control policies as well as to develop new methodologies or tools to better address issues surrounding infectious disease outbreaks.

**Fernando Alarid-Escudero, PhD** <a href="https://www.cide.edu/dap/en/profesores/profesor/?id=277" target="_blank"><i class="fas fa-link"></i></a> <a href="https://github.com/feralaes" target="_blank"><i class="fab fa-github"></i></a> 

```{r, echo=F}
blogdown::shortcode("vimeo", "473565068")
```

Fernando Alarid-Escudero, PhD, is an Assistant Professor in the Division of Public Administration at the Center for Research and Teaching in Economics (CIDE) in Aguascalientes. He has a bachelor’s degree in biomedical engineering from the Autonomous Metropolitan University (UAM) and a master’s degree in Economics from the Center for Research and Teaching in Economics (CIDE), both in Mexico City, and holds a PhD in health decision sciences from the University of Minnesota. He wants to live in an open-source world where academic & scientific developments are transparent, shareable and improvable. He likes to integrate methods from different disciplines such as statistics, engineering, economics and epidemiology to implement decision-analytic models in R. 

**Hawre Jalal, MD, PhD** <a href="https://publichealth.pitt.edu/home/directory/hawre-jalal" target="_blank"><i class="fas fa-link"></i></a> <a href="https://github.com/hjalal" target="_blank"><i class="fab fa-github"></i></a> 

```{r, echo=F}
blogdown::shortcode("vimeo", "473576731")
```

Hawre Jalal, MD, PhD is an Assistant Professor at the Department of Health Policy and Management and a researcher at the Public Health Dynamic Laboratory at the University of Pittsburgh. His research focuses on using mathematical modeling, cost-effective analyses and econometrics approaches to address health policy and resource allocation questions. His current research involves modeling the opioid-epidemic and evaluating its mitigation strategies, and research in disability and aging. Hawre’s methodological interests involve value of information analysis, Bayesian calibration and communicating uncertainty to decision makers.



