---
title: "Introduction to Decision Modeling"
date: "`r Sys.Date()`"
draft: no
output: pdf_document
linktitle: Intro to Decision Modeling
menu:
  day1:
    parent: Pre-workshop Lectures
    weight: 1
type: docs
---
  
## Slides

[Download slides](../../../Course_Modularization/Intro, general, packages/DecisionModelingIntroSlides.pdf)

### Introduction to Decision Analysis lecture videos

Introduction to Decision Analysis 1: Intro

```{r, echo=F}
blogdown::shortcode("vimeo", "588658361")
```

Introduction to Decision Analysis 2: Overview

```{r, echo=F}
blogdown::shortcode("vimeo", "588658378")
```

Introduction to Decision Analysis 3: Outcomes

```{r, echo=F}
blogdown::shortcode("vimeo", "588660950")
```

Introduction to Decision Analysis 4: Strengths and Challenges

```{r, echo=F}
blogdown::shortcode("vimeo", "588655554")
```

Introduction to Decision Analysis 5: Decision Tree Example

```{r, echo=F}
blogdown::shortcode("vimeo", "588655268")
```

Introduction to Decision Analysis 6: STM

```{r, echo=F}
blogdown::shortcode("vimeo", "588654943")
```

Introduction to Decision Analysis 7: Which Model

```{r, echo=F}
blogdown::shortcode("vimeo", "588654750")
```
