---
title: 'PSA: Decision tree model in R - Ondansetron'
author: "The DARTH workgroup"
output:
  pdf_document: default
  html_document: default
keep_tex: yes
self_contained: no
---

Developed by the Decision Analysis in R for Technologies in Health (DARTH) workgroup:

Fernando Alarid-Escudero, PhD (1) 

Eva A. Enns, MS, PhD (2)	

M.G. Myriam Hunink, MD, PhD (3,4)

Hawre J. Jalal, MD, PhD (5) 

Eline M. Krijkamp, MSc (3)	

Petros Pechlivanoglou, PhD (6,7)

Alan Yang, MSc (7)

In collaboration of: 		

1. Drug Policy Program, Center for Research and Teaching in Economics (CIDE) - CONACyT, 
   Aguascalientes, Mexico
2. University of Minnesota School of Public Health, Minneapolis, MN, USA
3. Erasmus MC, Rotterdam, The Netherlands
4. Harvard T.H. Chan School of Public Health, Boston, USA
5. University of Pittsburgh Graduate School of Public Health, Pittsburgh, PA, USA
6. University of Toronto, Toronto ON, Canada
7. The Hospital for Sick Children, Toronto ON, Canada

Please cite our publications when using this code:
 
- Jalal H, Pechlivanoglou P, Krijkamp E, Alarid-Escudero F, Enns E, Hunink MG. 
An Overview of R in Health Decision Sciences. Med Decis Making. 2017; 37(3): 735-746. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X16686559
 
- Krijkamp EM, Alarid-Escudero F, Enns EA, Jalal HJ, Hunink MGM, Pechlivanoglou P. 
Microsimulation modeling for health decision sciences using R: A tutorial. 
Med Decis Making. 2018;38(3):400–22. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X18754513
 
- Krijkamp EM, Alarid-Escudero F, Enns E, Pechlivanoglou P, Hunink MM, Jalal H. 
A Multidimensional Array Representation of State-Transition Model Dynamics. 
Med Decis Making. 2020 Online first. https://doi.org/10.1177/0272989X19893973

Copyright 2017, THE HOSPITAL FOR SICK CHILDREN AND THE COLLABORATING INSTITUTIONS. 
All rights reserved in Canada, the United States and worldwide. Copyright, 
trademarks, trade names and any and all associated intellectual property are 
exclusively owned by THE HOSPITAL FOR Sick CHILDREN and the collaborating 
institutions. These materials may be used, reproduced, modified, distributed 
and adapted with proper attribution.

\newpage

```{r}
rm(list = ls())    # clear memory (removes all the variables from the workspace)
```

# 01 Load packages

```{r, warning = FALSE, message = FALSE}
if (!require('pacman')) install.packages('pacman'); library(pacman) # use this package to conveniently install other packages
# load (install if required) packages from CRAN
p_load("here", "dplyr", "devtools", "scales", "ellipse", "ggplot2", "lazyeval", "igraph", "truncnorm", "ggraph", "reshape2", "knitr")                                               
# load (install if required) packages from GitHub
# install_github("DARTH-git/dampack", force = TRUE) Uncomment if there is a newer version
# install_github("DARTH-git/dectree", force = TRUE) Uncomment if there is a newer version
p_load_gh("DARTH-git/dampack", "DARTH-git/dectree")
```

# 02 Load functions

```{r}
source(here::here('functions','Functions.R'))
```

# 03 Define parameter input values

```{r}
v_names_str <- c('Ondasentron', 'Metoclopramide') # names of strategies
n_str       <- length(v_names_str)                # number of strategies
wtp         <- 500                                # willingness to pay threshold

# input values - Ondansetron  
p_sem_on                    <- 0.25   # Prob. of significant emesis
p_sem_ade_on                <- 0.26   # Prob. of ADE after emesis
p_sem_ade_trt_on            <- 0.60   # Prob. of treatment after ADE with emesis
p_sem_ade_trt_res_on        <- 0.66   # Prob. of resolution of treated ADE with emesis
p_sem_ade_no_trt_res_on     <- 1      # Prob. of resolution after ADE that did not 
                                      # require treament

p_no_sem_ade_on             <- 0.11   # Prob. of ADE without emesis
p_no_sem_ade_trt_on         <- 0.17   # Prob. of treatment after ADE without emesis
p_no_sem_ade_trt_res_on     <- 1      # Prob. of resolution of treated ADE without emesis
p_no_sem_ade_no_trt_res_on  <- 1      # Prob. of resolution of non treated ADE without 
                                      # emesis

c_trt_on                    <- 10     # cost of treatment
c_sem_on                    <- 30     # cost of episode of emesis
c_ade_on                    <- 20     # cost of ADE
c_ade_trt_on                <-  5     # cost of treating an ADE

# input values - metoclopramide  
# your turn

# store the parameters into a list
l_params_all <- as.list(data.frame(p_sem_on, p_sem_ade_on, p_sem_ade_trt_on, 
                                   p_sem_ade_trt_res_on, p_sem_ade_no_trt_res_on, 
                                   p_no_sem_ade_on, p_no_sem_ade_trt_on, 
                                   p_no_sem_ade_trt_res_on, p_no_sem_ade_no_trt_res_on,
                                   c_trt_on, c_sem_on, c_ade_on, c_ade_trt_on,
                                   
                                   p_sem_met, p_sem_ade_met, p_sem_ade_trt_met, 
                                   p_sem_ade_trt_res_met, p_sem_ade_no_trt_res_met,
                                   p_no_sem_ade_met, p_no_sem_ade_trt_met, 
                                   p_no_sem_ade_trt_res_met, p_no_sem_ade_no_trt_res_met,
                                   c_trt_met, c_sem_met, c_ade_met, c_ade_trt_met))

# store the names of the parameters into a vector
v_names_params <- c(               'p_sem_on', 'p_sem_ade_on', 'p_sem_ade_trt_on', 
                                   'p_sem_ade_trt_res_on', 'p_sem_ade_no_trt_res_on', 
                                   'p_no_sem_ade_on', 'p_no_sem_ade_trt_on', 
                                   'p_no_sem_ade_trt_res_on', 'p_no_sem_ade_no_trt_res_on',
                                   'c_trt_on', 'c_sem_on', 'c_ade_on', 'c_ade_trt_on',
                                   
                                   'p_sem_met', 'p_sem_ade_met', 'p_sem_ade_trt_met', 
                                   'p_sem_ade_trt_res_met', 'p_sem_ade_no_trt_res_met',
                                   'p_no_sem_ade_met', 'p_no_sem_ade_trt_met', 
                                   'p_no_sem_ade_trt_res_met', 'p_no_sem_ade_no_trt_res_met',
                                   'c_trt_met', 'c_sem_met', 'c_ade_met', 'c_ade_trt_met')
```

# 04 Create and run decision tree model

```{r}
decision_tree_ondansetron_output <- with(as.list(l_params_all), {
  
  # Create vector of weights for each strategy 
  # weight per path for the ondasentron arm
  
  v_w_on <-  c(
   # emesis with no ADEs
   p_sem_on * (1 - p_sem_ade_on),      
   # ADE after emesis that was treated and resolved
   p_sem_on * p_sem_ade_on * p_sem_ade_trt_on * p_sem_ade_trt_res_on,  
   # ADE after emesis that was treated and did not resolve
   p_sem_on * p_sem_ade_on * p_sem_ade_trt_on * (1 - p_sem_ade_trt_res_on),  
   # ADE after emesis that was not treated and resolved
   p_sem_on * p_sem_ade_on * (1 - p_sem_ade_trt_on) * p_sem_ade_no_trt_res_on,
   # ADE after emesis that was not treated and did not resolve
   p_sem_on * p_sem_ade_on * (1 - p_sem_ade_trt_on) * (1 - p_sem_ade_no_trt_res_on),
   # no emesis with no ADEs
   (1 - p_sem_on) * (1 - p_no_sem_ade_on),
   # ADE after no emesis that was treated and resolved
   (1 - p_sem_on) * p_no_sem_ade_on * p_no_sem_ade_trt_on * p_no_sem_ade_trt_res_on, 
   # ADE after no emesis that was treated and did not resolve
   (1 - p_sem_on) * p_no_sem_ade_on * p_no_sem_ade_trt_on * (1 - p_no_sem_ade_trt_res_on),
   # ADE after no emesis that was not treated and resolved
   (1 - p_sem_on) * p_no_sem_ade_on * (1 - p_no_sem_ade_trt_on) * p_no_sem_ade_no_trt_res_on, 
   # ADE after no emesis that was not treated and did not resolve
   (1 - p_sem_on) * p_no_sem_ade_on * (1 - p_no_sem_ade_trt_on) * 
   (1 - p_no_sem_ade_no_trt_res_on))   
  
  # weight per path for the metoclopramide arm
  
  v_w_met <-  c(
   # emesis with no ADEs
   p_sem_met * (1 - p_sem_ade_met),      
   # ADE after emesis that was treated and resolved
   p_sem_met * p_sem_ade_met * p_sem_ade_trt_met * p_sem_ade_trt_res_met,  
   # ADE after emesis that was treated and did not resolve
   p_sem_met * p_sem_ade_met * p_sem_ade_trt_met * (1 - p_sem_ade_trt_res_met),  
   # ADE after emesis that was not treated and resolved
   p_sem_met * p_sem_ade_met * (1 - p_sem_ade_trt_met) * p_sem_ade_no_trt_res_met,
   # ADE after emesis that was not treated and did not resolve
   p_sem_met * p_sem_ade_met * (1 - p_sem_ade_trt_met) * (1 - p_sem_ade_no_trt_res_met),
   # no emesis with no ADEs
   (1 - p_sem_met) * (1 - p_no_sem_ade_met),
   # ADE after no emesis that was treated and resolved
   (1 - p_sem_met) * p_no_sem_ade_met * p_no_sem_ade_trt_met * p_no_sem_ade_trt_res_met, 
   # ADE after no emesis that was treated and did not resolve
   (1 - p_sem_met) * p_no_sem_ade_met * p_no_sem_ade_trt_met * (1 - p_no_sem_ade_trt_res_met),
   # ADE after no emesis that was not treated and resolved
   (1 - p_sem_met) * p_no_sem_ade_met * (1 - p_no_sem_ade_trt_met) * p_no_sem_ade_no_trt_res_met, 
   # ADE after no emesis that was not treated and did not resolve
   (1 - p_sem_met) * p_no_sem_ade_met * (1 - p_no_sem_ade_trt_met) * 
   (1 - p_no_sem_ade_no_trt_res_met))   
  
  # Create vector of costs for each strategy 
 
  # Estimating cost per path for the ondasentron arm
  v_c_on <-                 
    c(c_trt_on + c_sem_on,
      c_trt_on + c_sem_on + c_ade_on + c_ade_trt_on,
      c_trt_on + c_sem_on + c_ade_on + c_ade_trt_on, 
      c_trt_on + c_sem_on + c_ade_on,
      c_trt_on + c_sem_on + c_ade_on,
      c_trt_on,
      c_trt_on + c_ade_on + c_ade_trt_on,
      c_trt_on + c_ade_on + c_ade_trt_on, 
      c_trt_on + c_ade_on,
      c_trt_on + c_ade_on)  
  
  # Estimating cost per path for the metoclopramide arm
  v_c_met <-              
    c(c_trt_met + c_sem_met,
      c_trt_met + c_sem_met + c_ade_met + c_ade_trt_met,
      c_trt_met + c_sem_met + c_ade_met + c_ade_trt_met, 
      c_trt_met + c_sem_met + c_ade_met,
      c_trt_met + c_sem_met + c_ade_met,
      c_trt_met,
      c_trt_met + c_ade_met + c_ade_trt_met,
      c_trt_met + c_ade_met + c_ade_trt_met, 
      c_trt_met + c_ade_met,
      c_trt_met + c_ade_met) 
  
  # Create vector of utilities for each strategy 
  # vector of health outcomes for both therapies for each path
  v_e_on  <- v_e_met <- c(0, 
                          0,
                          0,
                          0, 
                          0, 
                          1,
                          0,
                          0, 
                          0,
                          0) 
  
  # Calculate total costs for each strategy 
  tc_on   <- v_w_on   %*%  v_c_on      
  tc_met  <- v_w_met  %*%  v_c_met
  
  # Calculate total utilities for each strategy 
  te_on   <- v_w_on   %*%  v_e_on      
  te_met  <- v_w_met  %*%  v_e_met
  
  v_tc    <- c(tc_on, tc_met)   # vector of total costs
  v_te    <- c(te_on, te_met)   # vector of total life years
  v_nmb   <- v_te * wtp - v_tc  # calculate vector of nmb
  
  # Name outcomes
  names(v_tc)  <- v_names_str   # names for the elements of the  tc  vector
  names(v_te)  <- v_names_str   # names for the elements of the  te vector
  names(v_nmb) <- v_names_str   # names for the elements of the nmb vector
  
  df_output <- data.frame(Strategy        =  v_names_str,
                          Cost            =  v_tc,
                          Effectiveness   =  v_te,
                          NMB             =  v_nmb)
  return(df_output) 
})

# model output
decision_tree_ondansetron_output
```

## 04.1 Plot the decision tree

```{r, out.width='100%'}
# your turn

```

## 05 Cost-Effectiveness Analysis

```{r}
# your turn

```

## 05.1 Plot frontier of Decision Tree

```{r}
# your turn

```

# 06 Deterministic Sensitivity Analysis

## 06.1 List of input parameters

```{r}
# your turn

```

## 06.2 Load decision tree model function

```{r}
# your turn

```

## 06.3 One-way sensitivity analysis (OWSA)

```{r}
# your turn

```

## 06.3.1 Plot OWSA

```{r}
# your turn

```

## 06.3.2 Optimal strategy with OWSA

```{r}
# your turn

```

## 06.3.3 Tornado plot

```{r}
# your turn

```

## 06.4 Two-way sensitivity analysis (TWSA)

```{r}
# your turn

```

## 06.4.1 Plot TWSA

```{r}
# your turn

```

# 07 Probabilistic Sensitivity Analysis (PSA)

```{r}
# your turn

```

## 07.1 Conduct probabilistic sensitivity analysis

```{r}
# your turn

```

## 07.2 Create PSA object for dampack

```{r}
# your turn

```

## 07.2.1 Save PSA objects

```{r}
# your turn

```

## 07.3 Create probabilistic analysis graphs

```{r}
# your turn

```

Vector with willingness-to-pay (WTP) thresholds.

```{r}
# your turn

```

## 07.3.1 Cost-Effectiveness Scatter plot

```{r}
# your turn

```

## 07.4 Conduct CEA with probabilistic output

```{r}
# your turn

```

## 07.4.1 Plot cost-effectiveness frontier

```{r}
# your turn

```

## 07.4.2 Cost-effectiveness acceptability curves (CEACs) and frontier (CEAF)

```{r}
# your turn

```

## 07.4.3 Expected Loss Curves (ELCs)

```{r}
# your turn

```

## 07.4.4 Expected value of perfect information (EVPI)

```{r}
# your turn

```

