---
title: "Three-strategy decision tree in R - HVE"
author: "The DARTH workgroup"
output:
  word_document: default
  pdf_document: default
  html_document: default
self_contained: no
keep_tex: yes
---

Developed by the Decision Analysis in R for Technologies in Health (DARTH) workgroup:

Fernando Alarid-Escudero, PhD (1) 

Eva A. Enns, MS, PhD (2)	

M.G. Myriam Hunink, MD, PhD (3,4)

Hawre J. Jalal, MD, PhD (5) 

Eline M. Krijkamp, MSc (3)	

Petros Pechlivanoglou, PhD (6, 7)

Alan Yang, MSc (7)

In collaboration of: 		

1. Drug Policy Program, Center for Research and Teaching in Economics (CIDE) - CONACyT, 
   Aguascalientes, Mexico
2. University of Minnesota School of Public Health, Minneapolis, MN, USA
3. Erasmus MC, Rotterdam, The Netherlands
4. Harvard T.H. Chan School of Public Health, Boston, USA
5. University of Pittsburgh Graduate School of Public Health, Pittsburgh, PA, USA
6. University of Toronto, Toronto ON, Canada
7. The Hospital for Sick Children, Toronto ON, Canada

Please cite our publications when using this code:
 
- Jalal H, Pechlivanoglou P, Krijkamp E, Alarid-Escudero F, Enns E, Hunink MG. 
An Overview of R in Health Decision Sciences. Med Decis Making. 2017; 37(3): 735-746. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X16686559
 
- Krijkamp EM, Alarid-Escudero F, Enns EA, Jalal HJ, Hunink MGM, Pechlivanoglou P. 
Microsimulation modeling for health decision sciences using R: A tutorial. 
Med Decis Making. 2018;38(3):400–22. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X18754513
 
- Krijkamp EM, Alarid-Escudero F, Enns E, Pechlivanoglou P, Hunink MM, Jalal H. 
A Multidimensional Array Representation of State-Transition Model Dynamics. 
Med Decis Making. 2020 Online first. https://doi.org/10.1177/0272989X19893973

Copyright 2017, THE HOSPITAL FOR SICK CHILDREN AND THE COLLABORATING INSTITUTIONS. 
All rights reserved in Canada, the United States and worldwide. Copyright, 
trademarks, trade names and any and all associated intellectual property are 
exclusively owned by THE HOSPITAL FOR Sick CHILDREN and the collaborating 
institutions. These materials may be used, reproduced, modified, distributed 
and adapted with proper attribution.

\newpage

```{r}
rm(list = ls())      # clear memory (removes all the variables from the workspace)
```

# 01 Load packages

```{r, warning = FALSE, message = FALSE}
# all packages were loaded at the start of the course
```

# 02 Load functions

```{r}
source(here('functions','Functions.R'))
```

# 03 Define parameter input values

```{r}
v_names_str   <- c("No Tx", "Tx All", "Biopsy")    # names of strategies
n_str         <- length(v_names_str)               # number of strategies
wtp           <- 100000                            # willingness to pay threshold

# Probabilities
p_HVE         <- 0.52   # prevalence of HVE
p_HVE_comp    <- 0.71   # complications with untreated HVE
p_OVE_comp    <- 0.01   # complications with untreated OVE
p_HVE_comp_tx <- 0.36   # complications with treated HVE
p_OVE_comp_tx <- 0.20   # complications with treated OVE
p_biopsy_death <- 0.005   # probability of death due to biopsy

# Costs
c_VE          <- 1200   # cost of viral encephalitis care without complications
c_VE_comp     <- 9000   # cost of viral encephalitis care with complications
c_tx          <- 9500   # cost of treatment
c_biopsy      <- 25000  # cost of brain biopsy

# QALYs
q_VE           <- 20     # remaining QALYs for those without VE-related complications
q_VE_comp      <- 19     # remaining QALYs for those with VE-related complications
q_loss_biopsy  <- 0.01   # one-time QALY loss due to brain biopsy
q_death_biopsy <- 0      # remaining QALYs for those who died during biopsy
```

# 04 Create and run decision tree model

```{r}

  v_w_no_tx  <- c(  p_HVE  *    p_HVE_comp  ,  # HVE, complications
                    p_HVE  * (1-p_HVE_comp) ,  # HVE, no complications
                 (1-p_HVE) *    p_OVE_comp  ,  # OVE, complications
                 (1-p_HVE) * (1-p_OVE_comp))   # OVE, no complications
  
  v_w_tx     # <-  your turn
    
  v_w_biopsy # <-  your turn
    
  # Create vector of outcomes (QALYs) for each strategy 
  
  v_qaly_no_tx  <- c(q_VE_comp ,  # HVE, complications
                     q_VE      ,  # HVE, no complications
                     q_VE_comp ,  # OVE, complications
                     q_VE)        # OVE, no complications
  
  v_qaly_tx     # <-  your turn
    
  v_qaly_biopsy # <-  your turn
  
  
  # Create vector of costs for each strategy 

  v_cost_no_tx  <- c(c_VE_comp ,  # HVE, complications
                     c_VE      ,  # HVE, no complications
                     c_VE_comp ,  # OVE, complications
                     c_VE)        # OVE, no complications
  
  v_costs_tx     # <-  your turn
    
  v_costs_biopsy # <-  your turn 
  
  # Calculate total utilities for each strategy ####
  total_qaly_no_tx  <- v_w_no_tx  %*%  v_qaly_no_tx      
  total_qaly_tx     <- v_w_tx     %*%  v_qaly_tx
  total_qaly_biopsy <- v_w_biopsy %*%  v_qaly_biopsy
  
  # Calculate total costs for each strategy ####
  total_cost_no_tx  <- v_w_no_tx  %*%  v_cost_no_tx    
  total_cost_tx     <- v_w_tx     %*%  v_cost_tx
  total_cost_biopsy <- v_w_biopsy %*%  v_cost_biopsy
  
  v_total_qaly <- c(total_qaly_no_tx, total_qaly_tx, total_qaly_biopsy)  # vector of total QALYs
  v_total_cost <- c(total_cost_no_tx, total_cost_tx, total_cost_biopsy)  # vector of total costs
  v_nmb        <- v_total_qaly * wtp - v_total_cost                      # calculate vector of nmb
  
  # Name outcomes
  names(v_total_qaly) <- v_names_str  # names for the elements of the total QALYs vector
  names(v_total_cost) <- v_names_str  # names for the elements of the total cost vector
  names(v_nmb)        <- v_names_str  # names for the elements of the nmb vector

# model output
  
  df_output <- data.frame(Strategy =  v_names_str,
                          Cost     =  v_total_cost,
                          Effect   =  v_total_qaly,
                          NMB      =  v_nmb)
  
```


## 05 Cost-Effectiveness Analysis

```{r, eval = TRUE, echo = FALSE}
# your turn


```

## 05.1 Plot frontier of Decision Tree

```{r}
# your turn

```

