---
title: 'PSA: Decision tree model in R - Ondansetron'
author: "The DARTH workgroup"
output:
  pdf_document: default
  html_document: default
keep_tex: yes
self_contained: no
---

Developed by the Decision Analysis in R for Technologies in Health (DARTH) workgroup:

Fernando Alarid-Escudero, PhD (1) 

Eva A. Enns, MS, PhD (2)	

M.G. Myriam Hunink, MD, PhD (3,4)

Hawre J. Jalal, MD, PhD (5) 

Eline M. Krijkamp, MSc (3)	

Petros Pechlivanoglou, PhD (6, 7)

Alan Yang, MSc (7)

In collaboration of: 		

1. Drug Policy Program, Center for Research and Teaching in Economics (CIDE) - CONACyT, 
   Aguascalientes, Mexico
2. University of Minnesota School of Public Health, Minneapolis, MN, USA
3. Erasmus MC, Rotterdam, The Netherlands
4. Harvard T.H. Chan School of Public Health, Boston, USA
5. University of Pittsburgh Graduate School of Public Health, Pittsburgh, PA, USA
6. University of Toronto, Toronto ON, Canada
7. The Hospital for Sick Children, Toronto ON, Canada

Please cite our publications when using this code:
 
- Jalal H, Pechlivanoglou P, Krijkamp E, Alarid-Escudero F, Enns E, Hunink MG. 
An Overview of R in Health Decision Sciences. Med Decis Making. 2017; 37(3): 735-746. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X16686559
 
- Krijkamp EM, Alarid-Escudero F, Enns EA, Jalal HJ, Hunink MGM, Pechlivanoglou P. 
Microsimulation modeling for health decision sciences using R: A tutorial. 
Med Decis Making. 2018;38(3):400–22. 
https://journals.sagepub.com/doi/abs/10.1177/0272989X18754513
 
- Krijkamp EM, Alarid-Escudero F, Enns E, Pechlivanoglou P, Hunink MM, Jalal H. 
A Multidimensional Array Representation of State-Transition Model Dynamics. 
BioRxiv 670612 2019.https://www.biorxiv.org/content/10.1101/670612v1

Copyright 2017, THE HOSPITAL FOR SICK CHILDREN AND THE COLLABORATING INSTITUTIONS. 
All rights reserved in Canada, the United States and worldwide. Copyright, 
trademarks, trade names and any and all associated intellectual property are 
exclusively owned by THE HOSPITAL FOR Sick CHILDREN and the collaborating 
institutions. These materials may be used, reproduced, modified, distributed 
and adapted with proper attribution.

\newpage

```{r}
rm(list = ls())    # clear memory (removes all the variables from the workspace)
```

# 01 Load packages

```{r, warning = FALSE, message = FALSE}
if (!require('pacman')) install.packages('pacman'); library(pacman) # use this package to conveniently install other packages
# load (install if required) packages from CRAN
p_load("here", "dplyr", "devtools", "scales", "ellipse", "ggplot2", "lazyeval", "igraph", "truncnorm", "ggraph", 
       "reshape2", "knitr")                                               
# load (install if required) packages from GitHub
# install_github("DARTH-git/dampack", force = TRUE) Uncomment if there is a newer version
# install_github("DARTH-git/dectree", force = TRUE) Uncomment if there is a newer version
p_load_gh("DARTH-git/dampack", "DARTH-git/dectree")
```

# 02 Load functions

```{r}
source(here('functions','Functions.R'))
```

# 03 Define parameter input values

```{r}
v_names_str <- c('Ondasentron', 'Metoclopramide') # names of strategies
n_str       <- length(v_names_str)                # number of strategies
wtp         <- 500                                # willingness to pay threshold

# input values - Ondansetron  
p_sem_on                    <- 0.25   # Prob. of significant emesis
p_sem_ade_on                <- 0.26   # Prob. of ADE after emesis
p_sem_ade_trt_on            <- 0.60   # Prob. of treatment after ADE with emesis
p_sem_ade_trt_res_on        <- 0.66   # Prob. of resolution of treated ADE with emesis
p_sem_ade_no_trt_res_on     <- 1      # Prob. of resolution after ADE that did not 
                                      # require treament

p_no_sem_ade_on             <- 0.11   # Prob. of ADE without emesis
p_no_sem_ade_trt_on         <- 0.17   # Prob. of treatment after ADE without emesis
p_no_sem_ade_trt_res_on     <- 1      # Prob. of resolution of treated ADE without emesis
p_no_sem_ade_no_trt_res_on  <- 1      # Prob. of resolution of non treated ADE without 
                                      # emesis

c_trt_on                    <- 10     # cost of treatment
c_sem_on                    <- 30     # cost of episode of emesis
c_ade_on                    <- 20     # cost of ADE
c_ade_trt_on                <-  5     # cost of treating an ADE

# input values - metoclopramide  
p_sem_met                   <- 0.58   # Prob. of significant emesis
p_sem_ade_met               <- 0.34   # Prob. of ADE after emesis
p_sem_ade_trt_met           <- 0.60   # Prob. of treatment after ADE with emesis
p_sem_ade_trt_res_met       <- 0.78   # Prob. of resolution of treated ADE with emesis
p_sem_ade_no_trt_res_met    <- 1      # Prob. of resolution after ADE that did not 
                                      # require treament

p_no_sem_ade_met            <- 0.12   # Prob. of ADE without emesis
p_no_sem_ade_trt_met        <- 0.50   # Prob. of treatment after ADE without emesis
p_no_sem_ade_trt_res_met    <- 1      # Prob. of resolution of treated ADE without emesis
p_no_sem_ade_no_trt_res_met <- 1      # Prob. of resolution of non treated ADE without 
                                      # emesis 

c_trt_met                   <- 10     # cost of treatment
c_sem_met                   <- 30     # cost of episode of emesis
c_ade_met                   <- 20     # cost of ADE
c_ade_trt_met               <-  5     # cost of treating an ADE

# store the parameters into a list
l_params_all <- as.list(data.frame(p_sem_on, p_sem_ade_on, p_sem_ade_trt_on, 
                                   p_sem_ade_trt_res_on, p_sem_ade_no_trt_res_on, 
                                   p_no_sem_ade_on, p_no_sem_ade_trt_on, 
                                   p_no_sem_ade_trt_res_on, p_no_sem_ade_no_trt_res_on,
                                   c_trt_on, c_sem_on, c_ade_on, c_ade_trt_on,
                                   
                                   p_sem_met, p_sem_ade_met, p_sem_ade_trt_met, 
                                   p_sem_ade_trt_res_met, p_sem_ade_no_trt_res_met,
                                   p_no_sem_ade_met, p_no_sem_ade_trt_met, 
                                   p_no_sem_ade_trt_res_met, p_no_sem_ade_no_trt_res_met,
                                   c_trt_met, c_sem_met, c_ade_met, c_ade_trt_met))
# store the names of the parameters into a vector
v_names_params <- c(               'p_sem_on', 'p_sem_ade_on', 'p_sem_ade_trt_on', 
                                   'p_sem_ade_trt_res_on', 'p_sem_ade_no_trt_res_on', 
                                   'p_no_sem_ade_on', 'p_no_sem_ade_trt_on', 
                                   'p_no_sem_ade_trt_res_on', 'p_no_sem_ade_no_trt_res_on',
                                   'c_trt_on', 'c_sem_on', 'c_ade_on', 'c_ade_trt_on',
                                   
                                   'p_sem_met', 'p_sem_ade_met', 'p_sem_ade_trt_met', 
                                   'p_sem_ade_trt_res_met', 'p_sem_ade_no_trt_res_met',
                                   'p_no_sem_ade_met', 'p_no_sem_ade_trt_met', 
                                   'p_no_sem_ade_trt_res_met', 'p_no_sem_ade_no_trt_res_met',
                                   'c_trt_met', 'c_sem_met', 'c_ade_met', 'c_ade_trt_met')
```

# 04 Create and run decision tree model

```{r}
decision_tree_ondansetron_output <- with(as.list(l_params_all), {
  
  # Create vector of weights for each strategy 
  # weight per path for the ondasentron arm
  
  v_w_on <-  c(
   # emesis with no ADEs
   p_sem_on * (1 - p_sem_ade_on),      
   # ADE after emesis that was treated and resolved
   p_sem_on * p_sem_ade_on * p_sem_ade_trt_on * p_sem_ade_trt_res_on,  
   # ADE after emesis that was treated and did not resolve
   p_sem_on * p_sem_ade_on * p_sem_ade_trt_on * (1 - p_sem_ade_trt_res_on),  
   # ADE after emesis that was not treated and resolved
   p_sem_on * p_sem_ade_on * (1 - p_sem_ade_trt_on) * p_sem_ade_no_trt_res_on,
   # ADE after emesis that was not treated and did not resolve
   p_sem_on * p_sem_ade_on * (1 - p_sem_ade_trt_on) * (1 - p_sem_ade_no_trt_res_on),
   # no emesis with no ADEs
   (1 - p_sem_on) * (1 - p_no_sem_ade_on),
   # ADE after no emesis that was treated and resolved
   (1 - p_sem_on) * p_no_sem_ade_on * p_no_sem_ade_trt_on * p_no_sem_ade_trt_res_on, 
   # ADE after no emesis that was treated and did not resolve
   (1 - p_sem_on) * p_no_sem_ade_on * p_no_sem_ade_trt_on * (1 - p_no_sem_ade_trt_res_on),
   # ADE after no emesis that was not treated and resolved
   (1 - p_sem_on) * p_no_sem_ade_on * (1 - p_no_sem_ade_trt_on) * p_no_sem_ade_no_trt_res_on, 
   # ADE after no emesis that was not treated and did not resolve
   (1 - p_sem_on) * p_no_sem_ade_on * (1 - p_no_sem_ade_trt_on) * 
   (1 - p_no_sem_ade_no_trt_res_on))   
  
  # weight per path for the metoclopramide arm
  
  v_w_met <-  c(
   # emesis with no ADEs
   p_sem_met * (1 - p_sem_ade_met),      
   # ADE after emesis that was treated and resolved
   p_sem_met * p_sem_ade_met * p_sem_ade_trt_met * p_sem_ade_trt_res_met,  
   # ADE after emesis that was treated and did not resolve
   p_sem_met * p_sem_ade_met * p_sem_ade_trt_met * (1 - p_sem_ade_trt_res_met),  
   # ADE after emesis that was not treated and resolved
   p_sem_met * p_sem_ade_met * (1 - p_sem_ade_trt_met) * p_sem_ade_no_trt_res_met,
   # ADE after emesis that was not treated and did not resolve
   p_sem_met * p_sem_ade_met * (1 - p_sem_ade_trt_met) * (1 - p_sem_ade_no_trt_res_met),
   # no emesis with no ADEs
   (1 - p_sem_met) * (1 - p_no_sem_ade_met),
   # ADE after no emesis that was treated and resolved
   (1 - p_sem_met) * p_no_sem_ade_met * p_no_sem_ade_trt_met * p_no_sem_ade_trt_res_met, 
   # ADE after no emesis that was treated and did not resolve
   (1 - p_sem_met) * p_no_sem_ade_met * p_no_sem_ade_trt_met * (1 - p_no_sem_ade_trt_res_met),
   # ADE after no emesis that was not treated and resolved
   (1 - p_sem_met) * p_no_sem_ade_met * (1 - p_no_sem_ade_trt_met) * p_no_sem_ade_no_trt_res_met, 
   # ADE after no emesis that was not treated and did not resolve
   (1 - p_sem_met) * p_no_sem_ade_met * (1 - p_no_sem_ade_trt_met) * 
   (1 - p_no_sem_ade_no_trt_res_met))   
  
  # Create vector of costs for each strategy 
 
  # Estimating cost per path for the ondasentron arm
  v_c_on <-                 
    c(c_trt_on + c_sem_on,
      c_trt_on + c_sem_on + c_ade_on + c_ade_trt_on,
      c_trt_on + c_sem_on + c_ade_on + c_ade_trt_on, 
      c_trt_on + c_sem_on + c_ade_on,
      c_trt_on + c_sem_on + c_ade_on,
      c_trt_on,
      c_trt_on + c_ade_on + c_ade_trt_on,
      c_trt_on + c_ade_on + c_ade_trt_on, 
      c_trt_on + c_ade_on,
      c_trt_on + c_ade_on)  
  
  # Estimating cost per path for the metoclopramide arm
  v_c_met <-              
    c(c_trt_met + c_sem_met,
      c_trt_met + c_sem_met + c_ade_met + c_ade_trt_met,
      c_trt_met + c_sem_met + c_ade_met + c_ade_trt_met, 
      c_trt_met + c_sem_met + c_ade_met,
      c_trt_met + c_sem_met + c_ade_met,
      c_trt_met,
      c_trt_met + c_ade_met + c_ade_trt_met,
      c_trt_met + c_ade_met + c_ade_trt_met, 
      c_trt_met + c_ade_met,
      c_trt_met + c_ade_met) 
  
  # Create vector of utilities for each strategy 
  # vector of health outcomes for both therapies for each path
  v_e_on  <- v_e_met <- c(0, 
                          0,
                          0,
                          0, 
                          0, 
                          1,
                          0,
                          0, 
                          0,
                          0) 
  
  # Calculate total costs for each strategy 
  tc_on   <- v_w_on   %*%  v_c_on      
  tc_met  <- v_w_met  %*%  v_c_met
  
  # Calculate total utilities for each strategy 
  te_on   <- v_w_on   %*%  v_e_on      
  te_met  <- v_w_met  %*%  v_e_met
  
  v_tc    <- c(tc_on, tc_met)   # vector of total costs
  v_te    <- c(te_on, te_met)   # vector of total life years
  v_nmb   <- v_te * wtp - v_tc  # calculate vector of nmb
  
  # Name outcomes
  names(v_tc)  <- v_names_str   # names for the elements of the  tc  vector
  names(v_te)  <- v_names_str   # names for the elements of the  te vector
  names(v_nmb) <- v_names_str   # names for the elements of the nmb vector
  
  df_output <- data.frame(Strategy        =  v_names_str,
                          Cost            =  v_tc,
                          Effectiveness   =  v_te,
                          NMB             =  v_nmb)
  return(df_output) 
})

# model output
decision_tree_ondansetron_output
```

## 04.1 Plot the decision tree

```{r, out.width='100%'}
branches <- read.csv(here('data','decision_tree_ondansetron_branches.csv'), stringsAsFactors = F, header = T)
tree     <- create_tree(branches)
# plot_tree(tree, font.size = 5)
```

## 05 Cost-Effectiveness Analysis

```{r}
# create the transition probability matrix for NO treatment
decision_tree_ondansetron_cea  <- calculate_icers(cost       = decision_tree_ondansetron_output$Cost,
                                                  effect     = decision_tree_ondansetron_output$Effect,
                                                  strategies = decision_tree_ondansetron_output$Strategy)
decision_tree_ondansetron_cea
```

## 05.1 Plot frontier of Decision Tree

```{r}
plot(decision_tree_ondansetron_cea, effect_units = "QALYs")
```

# 06 Deterministic Sensitivity Analysis

## 06.1 List of input parameters

```{r}
l_params_all 
```

## 06.2 Load decision tree model function

```{r}
source(here("functions", "Functions_decision_tree_ondansetron.R"))
# Test function
calculate_ce_out(l_params_all)
```

## 06.3 One-way sensitivity analysis (OWSA)

```{r, warning=F}
params_owsa      <- c("p_sem_on", "p_no_sem_ade_on", "p_sem_met", "p_no_sem_ade_met") # parameters of interest for OWSA
# dataframe containing all parameters, their basecase values, and the min and max values of the parameters of interest 
params_all_owsa  <- data.frame(pars = v_names_params, basecase = as.numeric(l_params_all), min = rep(NA, length(v_names_params)), 
                               max  = rep(NA, length(v_names_params))) 
params_all_owsa$min[params_all_owsa$pars %in% params_owsa] <- c(0, 0, 0, 0)  # min parameter values
params_all_owsa$max[params_all_owsa$pars %in% params_owsa] <- c(1, 1, 1, 1)  # max parameter values
owsa_nmb         <- run_owsa_det(params     = params_owsa,      # parameters of interest
                                 params_all = params_all_owsa,  # dataframe containing relevant parameter values
                                 nsamp      = 100,              # number of parameter values
                                 FUN        = calculate_ce_out, # function to compute outputs
                                 outcomes   = c("NMB"),         # output to do the OWSA on
                                 strategies = v_names_str)      # names of the strategies
```

## 06.3.1 Plot OWSA

```{r}
plot(owsa_nmb, txtsize = 16, n_x_ticks = 5, 
     facet_scales = "free") +
     theme(legend.position = "bottom")
```

## 06.3.2 Optimal strategy with OWSA

```{r}
owsa_opt_strat(owsa = owsa_nmb)
```

## 06.3.3 Tornado plot

```{r}
owsa_tornado(owsa = owsa_nmb)
```

## 06.4 Two-way sensitivity analysis (TWSA)

```{r, warning=F}
params_twsa     <- c("p_sem_on", "p_no_sem_ade_met") # parameters of interest for TWSA
# dataframe containing all parameters, their basecase values, and the min and max values of the parameters of interest
params_all_twsa <- data.frame(pars = v_names_params, basecase = as.numeric(l_params_all), min = rep(NA, length(v_names_params)), 
                              max  = rep(NA, length(v_names_params))) 
params_all_twsa$min[params_all_twsa$pars %in% params_twsa] <- c(0, 0)  # min parameter values
params_all_twsa$max[params_all_twsa$pars %in% params_twsa] <- c(1, 1)  # max parameter values
twsa_nmb        <- run_twsa_det(param1     = params_twsa[1],   # first parameters of interest
                                param2     = params_twsa[2],   # second parameters of interest 
                                params_all = params_all_twsa,  # dataframe containing relevant parameter values
                                nsamp      = 40,               # number of parameter values
                                FUN        = calculate_ce_out, # function to compute outputs
                                outcomes   = c("NMB"),         # output to do the twsa on
                                strategies = v_names_str)      # names of the strategies
```

## 06.4.1 Plot TWSA

```{r}
plot(twsa_nmb) + 
  ggtitle(label = "Two-way sensitivity analysis", subtitle = "Net monetary benefit") +
  theme(legend.position = "bottom")
```

# 07 Probabilistic Sensitivity Analysis (PSA)

```{r, warning=F}
# Function to generate PSA input dataset
generate_psa_params <- function(n_sim = 1000, seed = 071518){
  set.seed(seed)
  ### Dataframe of input parameters
  df_psa_params   <- data.frame(
    
    # Transition probabilities (per cycle)
    p_sem_on                      = rbeta(n_sim, 19, 56),  
    p_sem_ade_on                  = rbeta(n_sim, 19, 57),  
    p_sem_ade_trt_on              = rbeta(n_sim, 14, 9),  
    p_sem_ade_trt_res_on          = rbeta(n_sim, 14, 7),  
    p_sem_ade_no_trt_res_on       = rbeta(n_sim, 0.99, 0.03),  
    p_no_sem_ade_on               = rbeta(n_sim, 4, 34),  
    p_no_sem_ade_trt_on           = rbeta(n_sim, 26, 130),  
    p_no_sem_ade_trt_res_on       = 1,  
    p_no_sem_ade_no_trt_res_on    = 0,  
    p_sem_met                     = rbeta(n_sim, 6, 4),  
    p_sem_ade_met                 = rbeta(n_sim, 7, 14),  
    p_sem_ade_trt_met             = rbeta(n_sim, 6, 4),  
    p_sem_ade_trt_res_met         = rbeta(n_sim, 3, 1),  
    p_sem_ade_no_trt_res_met      = rbeta(n_sim, 0.99, 0.03),  
    p_no_sem_ade_met              = rbeta(n_sim, 4, 25),  
    p_no_sem_ade_trt_met          = rbeta(n_sim, 12, 12),  
    p_no_sem_ade_trt_res_met      = 1,  
    p_no_sem_ade_no_trt_res_met   = 0,  

    ### Costs
    c_trt_on                      = rgamma(n_sim, shape=25, scale=0.4),  
    c_sem_on                      = rgamma(n_sim, shape=56, scale=0.5),  
    c_ade_on                      = rgamma(n_sim, shape=45, scale=0.45),  
    c_ade_trt_on                  = rgamma(n_sim, shape=25, scale=0.2),  
    c_trt_met                     = rgamma(n_sim, shape=25, scale=0.4),  
    c_sem_met                     = rgamma(n_sim, shape=56, scale=0.5),  
    c_ade_met                     = rgamma(n_sim, shape=45, scale=0.45),  
    c_ade_trt_met                 = rgamma(n_sim, shape=25, scale=0.2) 
  )
  return(df_psa_params)
}
# Try it
generate_psa_params(10) 

# Number of simulations
n_sim <- 1000

# Generate PSA input dataset
df_psa_input     <- generate_psa_params(n_sim = n_sim)
# First six observations
head(df_psa_input)

# Histogram of parameters
ggplot(melt(df_psa_input, variable.name = "Parameter"), aes(x = value)) +
  facet_wrap(~Parameter, scales = "free") +
  geom_histogram(aes(y = ..density..)) +
  theme_bw(base_size = 16)

# Initialize dataframes with PSA output 
# Dataframe of costs
df_c <- as.data.frame(matrix(0, 
                      nrow = n_sim,
                      ncol = n_str))
colnames(df_c) <- v_names_str
# Dataframe of effectiveness
df_e <- as.data.frame(matrix(0, 
                      nrow = n_sim,
                      ncol = n_str))
colnames(df_e) <- v_names_str
```

## 07.1 Conduct probabilistic sensitivity analysis

```{r}
# Run decision tree on each parameter set of PSA input dataset
for(i in 1:n_sim){
  df_out_temp <- calculate_ce_out(df_psa_input[i, ])
  df_c[i, ] <- df_out_temp$Cost
  df_e[i, ] <- df_out_temp$Effect
  # Display simulation progress
  if(i/(n_sim/10) == round(i/(n_sim/10),0)) { # display progress every 10%
    cat('\r', paste(i/n_sim * 100, "% done", sep = " "))
  }
}
```

## 07.2 Create PSA object for dampack

```{r}
l_psa <- make_psa_obj(cost          = df_c, 
                      effectiveness = df_e, 
                      parameters    = df_psa_input, 
                      strategies    = v_names_str)
```

## 07.2.1 Save PSA objects

```{r}
save(df_psa_input, df_c, df_e, v_names_str, n_str,
     l_psa,
     file = here("output", "decision_tree_ondansetron_PSA_dataset.RData"))
```

## 07.3 Create probabilistic analysis graphs

```{r}
load(file = here("output", "decision_tree_ondansetron_PSA_dataset.RData"))
```

Vector with willingness-to-pay (WTP) thresholds.

```{r}
v_wtp <- seq(0, 300000, by = 10000)
```

## 07.3.1 Cost-Effectiveness Scatter plot

```{r}
plot(l_psa)
```

## 07.4 Conduct CEA with probabilistic output

```{r}
# Compute expected costs and effects for each strategy from the PSA
df_out_ce_psa <- summary(l_psa)
df_out_ce_psa

# Calculate incremental cost-effectiveness ratios (ICERs)
df_cea_psa <- calculate_icers(cost       = df_out_ce_psa$meanCost, 
                              effect     = df_out_ce_psa$meanEffect,
                              strategies = df_out_ce_psa$Strategy)
df_cea_psa

# Save CEA table with ICERs
# As .RData
save(df_cea_psa, 
     file = here("tables", 
                 "decision_tree_ondansetron_probabilistic_CEA_results.RData"))
# As .csv
write.csv(df_cea_psa, 
          file = here("tables", 
                      "decision_tree_ondansetron_probabilistic_CEA_results.csv"))
```

## 07.4.1 Plot cost-effectiveness frontier

```{r}
plot(df_cea_psa)
```

## 07.4.2 Cost-effectiveness acceptability curves (CEACs) and frontier (CEAF)

```{r}
ceac_obj <- ceac(wtp = v_wtp, psa = l_psa)
# Regions of highest probability of cost-effectiveness for each strategy
summary(ceac_obj)
# CEAC & CEAF plot
plot(ceac_obj)
```

## 07.4.3 Expected Loss Curves (ELCs)

```{r}
elc_obj <- calc_exp_loss(wtp = v_wtp, psa = l_psa)
elc_obj
# ELC plot
plot(elc_obj, log_y = FALSE)
```

## 07.4.4 Expected value of perfect information (EVPI)

```{r}
evpi <- calc_evpi(wtp = v_wtp, psa = l_psa)
# EVPI plot
plot(evpi, effect_units = "QALY")
```

