---
title: "PSA template - DRAFT"
author: "DARTH"
date: "5/04/2021"
output: html_document
---

Warning: This is a draft idea of how a csv or excel data set can be used to import parameter values information for many parameters at once and how this information in turn can be used to generate a PSA data set. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 01 Load packages

```{r}
if (!require('pacman')) install.packages('pacman'); library(pacman) # use this package to conveniently install other packages
# load (install if required) packages from CRAN
p_load( "dplyr", "devtools", "scales", "ellipse", "ggplot2", "lazyeval", "igraph", "ggraph", "reshape2", "knitr", "stringr", "diagram", "triangle")   
# load (install if required) packages from GitHub
p_load_gh("DARTH-git/dampack", "DARTH-git/darthtools")
```

# Load the functions
We need to load a function that is able to make a PSA dataset for us based on the data input in the csv or xlsx file.
```{r}
source("make_psa_df_function.R")
```

## Load the data and remove empty lines
The data can be loaded as a csv or a xlsx file. 
```{r cars}
# Read from the xlsx
#df_param  <- data.frame(readxl::read_xlsx("param.xlsx", sheet = "param"))

# Read from the csv
df_param <- read.csv(file = "param.csv", header = TRUE, sep = ";")

# document which parameters do not have a unit and will be deleted
# < 0 is used to find rows without information
param_NA <- df_param[which((df_param$unit < 0)), ]

# Remove parameters without unit
df_param <- df_param[df_param$unit>0, ]

# Make parameters values numeric, for those not missing
numvar <- c("par1", "par2", "par3")
for (i in numvar){
  v_row_NA <- is.na(df_param[, i])
  df_param[, i][!is.na(df_param[, i])] <- as.numeric(df_param[, i][!is.na(df_param[, i])])
}

```

# Run the functions
We use the `make_psa_df` function to generate as PSA dataset with parameter values for each PSA iteration. the parameter values are sampled from their corresponding distribution as is specified in the input data. 
```{r}
m_psa_input <- make_psa_df(df_param = df_param, n_iter = 5) # test run with 5 iterations 

# Note the output of make_psa_df is a matrix 
```

# Visually inspect the sampled results 
```{r}

m_psa_input <- make_psa_df(df_param = df_param, n_iter = 10000) # test run with 5 iterations 
# Adjust the numbers of which parameters you like to see here

df_psa_input <- as.data.frame(m_psa_input)

# Note the 

ggplot(melt(df_psa_input, variable.name =  "Parameter"), aes(x = value)) +
       facet_wrap(~Parameter, scales = "free") +
       geom_histogram(aes(y = ..density..)) +
       theme_bw(base_size = 16) + 
       theme(axis.text = element_text(size = 8))

```
