<<<<<<< HEAD
---
title: 'Markov Sick-Sicker model in R'
subtitle: "With simulation-time dependency"
author: "The DARTH workgroup"
output:
  pdf_document: default
  html_document: default
keep_tex: yes
self_contained: no
---

This code forms the basis for the state-transition model of the tutorial: 
'A Tutorial on Time-Dependent Cohort State-Transition Models in R using a Cost-Effectiveness Analysis Example' 

Authors: 
- Fernando Alarid-Escudero <fernando.alarid@cide.edu>
- Eline Krijkamp
- Eva A. Enns
- Alan Yang
- M.G. Myriam Hunink
- Petros Pechlivanoglou
- Hawre Jalal
Please cite the article when using this code

To program this tutorial we used:
R version 4.0.5 (2021-03-31)
Platform: 64-bit operating system, x64-based processor
Running under: Mac OS 12.2.1
RStudio: Version 1.4.1717 2009-2021 RStudio, Inc

This code implements a simulation-time-dependent Sick-Sicker cSTM model to  conduct a CEA of two strategies:
- Standard of Care (SoC): best available care for the patients with the disease. This scenario reflects the natural history of the disease  
  progression.
- Strategy AB: This strategy combines treatment A and treatment B. The disease  progression is reduced, and individuals in the Sick state have an  
  improved quality of life.

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

Change `eval` to `TRUE` if you want to knit this document.

```{r}
rm(list = ls())      # clear memory (removes all the variables from the workspace)
```

# 01 Load packages

```{r, warning = F, message = F}
if (!require('pacman')) install.packages('pacman'); library(pacman) # use this package to conveniently install other packages
# load (install if required) packages from CRAN
p_load("dplyr", "tidyr", "reshape2", "devtools", "scales", "ellipse", "ggplot2", "ggrepel", "gridExtra", "lazyeval", "igraph", "truncnorm", "ggraph", "reshape2", "patchwork", "knitr", "stringr", "diagram", "dampack")                                               
# load (install if required) packages from GitHub
# install_github("DARTH-git/darthtools", force = TRUE) #Uncomment if there is a newer version
p_load_gh("DARTH-git/darthtools")
```

# 02 Load functions

```{r}
# all functions are in the darthtools package
```

# 03 Model input

```{r}
## General setup 
cycle_length <- 1   # cycle length equal to one year (use 1/12 for monthly)
n_age_init   <- 25  # age at baseline
n_age_max    <- 100 # maximum age of follow up
n_cycles     <- (n_age_max - n_age_init)/cycle_length # time horizon, number of cycles
# Age labels 
v_age_names  <- paste(rep(n_age_init:(n_age_max-1), each = 1/cycle_length), 
                      1:(1/cycle_length), 
                      sep = ".")
# the 4 health states of the model:
v_names_states <- c("H",  # Healthy (H)
                    "S1", # Sick (S1)
                    "S2", # Sicker (S2)
                    "D")  # Dead (D)
                                           
n_states <- length(v_names_states)   # number of health states 

### Discounting factors 
d_c <- 0.03 # annual discount rate for costs 
d_e <- 0.03 # annual discount rate for QALYs

### Strategies 
v_names_str <- c("Standard of care", # store the strategy names
                 "Strategy AB") 
n_str       <- length(v_names_str)   # number of strategies

## Within-cycle correction (WCC) using Simpson's 1/3 rule 
v_wcc  <- gen_wcc(n_cycles = n_cycles, method = "Simpson1/3")

### Transition rates (annual), and hazard ratios (HRs) 
r_HS1  <- 0.15  # constant annual rate of becoming Sick when Healthy
r_S1H  <- 0.5   # constant annual rate of becoming Healthy when Sick
r_S1S2 <- 0.105 # constant annual rate of becoming Sicker when Sick
hr_S1  <- 3     # hazard ratio of death in Sick vs Healthy 
hr_S2  <- 10    # hazard ratio of death in Sicker vs Healthy 

### Effectiveness of treatment AB 
hr_S1S2_trtAB <- 0.6  # hazard ratio of becoming Sicker when Sick under treatment AB

## Age-dependent mortality rates 
lt_usa_2015 <- read.csv("HMD_USA_Mx_2015.csv")
# Extract age-specific all-cause mortality for ages in model time horizon
v_r_mort_by_age <- lt_usa_2015 %>% 
  dplyr::filter(Age >= n_age_init & Age < n_age_max) %>%
  dplyr::select(Total) %>%
  as.matrix()

### State rewards 
#### Costs 
c_H     <- 2000  # annual cost of being Healthy
c_S1    <- 4000  # annual cost of being Sick
c_S2    <- 15000 # annual cost of being Sicker
c_D     <- 0     # annual cost of being dead
c_trtAB <- 25000 # annual cost of receiving treatment AB
#### Utilities 
u_H     <- 1     # annual utility of being Healthy
u_S1    <- 0.75  # annual utility of being Sick
u_S2    <- 0.5   # annual utility of being Sicker
u_D     <- 0     # annual utility of being dead
u_trtAB <- 0.95  # annual utility when receiving treatment AB

### Transition rewards 
du_HS1  <- 0.01  # disutility when transitioning from Healthy to Sick
ic_HS1  <- 1000  # increase in cost when transitioning from Healthy to Sick
ic_D    <- 2000  # increase in cost when dying

### Discount weight for costs and effects 
v_dwc   <- 1 / ((1 + (d_e * cycle_length)) ^ (0:n_cycles))
v_dwe   <- 1 / ((1 + (d_c * cycle_length)) ^ (0:n_cycles))

# Process model inputs 
## Age-specific transition rates to the Dead state for all cycles 
v_r_HDage  <- rep(v_r_mort_by_age, each = 1/cycle_length)
# Name age-specific mortality vector 
names(v_r_HDage) <- v_age_names

# compute mortality rates
v_r_S1Dage  <- v_r_HDage * hr_S1 # Age-specific mortality rate in the Sick state 
v_r_S2Dage  <- v_r_HDage * hr_S2 # Age-specific mortality rate in the Sicker state 
# transform rates to probabilities adjusting by cycle length
p_HS1       <- rate_to_prob(r = r_HS1,  t = cycle_length) # constant annual probability of becoming Sick when Healthy conditional on surviving 
p_S1H       <- rate_to_prob(r = r_S1H,  t = cycle_length) # constant annual probability of becoming Healthy when Sick conditional on surviving
p_S1S2      <- rate_to_prob(r = r_S1S2, t = cycle_length) # constant annual probability of becoming Sicker when Sick conditional on surviving
v_p_HDage   <- rate_to_prob(v_r_HDage,  t = cycle_length) # Age-specific mortality risk in the Healthy state 
v_p_S1Dage  <- rate_to_prob(v_r_S1Dage, t = cycle_length) # Age-specific mortality risk in the Sick state
v_p_S2Dage  <- rate_to_prob(v_r_S2Dage, t = cycle_length) # Age-specific mortality risk in the Sicker state

## Annual transition probability of becoming Sicker when Sick for treatment AB 
# Apply hazard ratio to rate to obtain transition rate of becoming Sicker when Sick for treatment AB
r_S1S2_trtAB <- r_S1S2 * hr_S1S2_trtAB
# Transform rate to probability to become Sicker when Sick under treatment AB 
# adjusting by cycle length conditional on surviving
p_S1S2_trtAB <- rate_to_prob(r = r_S1S2_trtAB, t = cycle_length)
```

# 04 Construct state-transition models

## 04.1 Initial state vector

```{r}
# All starting healthy
v_m_init <- c(H = 1, S1 = 0, S2 = 0, D = 0) # initial state vector
v_m_init
```

## 04.2 Initialize cohort traces

```{r}
### Initialize cohort trace under SoC 
m_M_SoC <- matrix(NA, 
              nrow = (n_cycles + 1), ncol = n_states, 
              dimnames = list(0:n_cycles, v_names_states))
# Store the initial state vector in the first row of the cohort trace
m_M_SoC[1, ] <- v_m_init

### Initialize cohort trace for strategy AB 
# Structure and initial states are the same as for SoC
m_M_strAB <- m_M_SoC # Strategy AB
```

## 04.3 Create transition probability matrices

```{r}
## Create transition probability arrays for strategy SoC 
### Initialize transition probability array for strategy SoC 
# All transitions to a non-death state are assumed to be conditional on survival
a_P_SoC <- array(0,
                 dim  = c(n_states, n_states, n_cycles),
                 dimnames = list(v_names_states, 
                                 v_names_states, 
                                 0:(n_cycles - 1)))
### Fill in array
## From H
a_P_SoC["H", "H", ]   <- (1 - v_p_HDage) * (1 - p_HS1)
a_P_SoC["H", "S1", ]  <- (1 - v_p_HDage) *      p_HS1
a_P_SoC["H", "D", ]   <-      v_p_HDage
## From S1
a_P_SoC["S1", "H", ]  <- (1 - v_p_S1Dage) *       p_S1H
a_P_SoC["S1", "S1", ] <- (1 - v_p_S1Dage) * (1 - (p_S1H + p_S1S2))
a_P_SoC["S1", "S2", ] <- (1 - v_p_S1Dage) *               p_S1S2
a_P_SoC["S1", "D", ]  <-      v_p_S1Dage
## From S2
a_P_SoC["S2", "S2", ] <- 1 - v_p_S2Dage
a_P_SoC["S2", "D", ]  <-     v_p_S2Dage
## From D
a_P_SoC["D", "D", ]   <- 1

### Initialize transition probability array for strategy AB 
a_P_strAB <- a_P_SoC
# Update only transition probabilities from S1 involving p_S1S2
a_P_strAB["S1", "S1", ] <- (1 - v_p_S1Dage) * (1 - (p_S1H + p_S1S2_trtAB))
a_P_strAB["S1", "S2", ] <- (1 - v_p_S1Dage) *               p_S1S2_trtAB

## Check if transition probability arrays are valid 
### Check that transition probabilities are [0, 1] 
check_transition_probability(a_P_SoC,   verbose = TRUE)
check_transition_probability(a_P_strAB, verbose = TRUE)
### Check that all rows for each slice of the array sum to 1 
check_sum_of_transition_array(a_P_SoC,   n_states = n_states, n_cycles = n_cycles, verbose = TRUE)
check_sum_of_transition_array(a_P_strAB, n_states = n_states, n_cycles = n_cycles, verbose = TRUE)
```

## 04.4 Create transition dynamics arrays

```{r}
# These arrays will capture transitions from each state to another over time 
### Initialize transition dynamics array for strategy SoC 
a_A_SoC <- array(0,
             dim      = c(n_states, n_states, n_cycles + 1),
             dimnames = list(v_names_states, v_names_states, 0:n_cycles))
# Set first slice of a_A_SoC with the initial state vector in its diagonal
diag(a_A_SoC[, , 1]) <- v_m_init
### Initialize transition-dynamics array for strategy AB 
# Structure and initial states are the same as for SoC
a_A_strAB <- a_A_SoC
```

# 05 Run Markov model

```{r}
# Iterative solution of age-dependent cSTM
for(t in 1:n_cycles){
  ## Fill in cohort trace
  # For SoC
  m_M_SoC[t + 1, ]   <- m_M_SoC[t, ]   %*% a_P_SoC[, , t]
  # For strategy AB 
  m_M_strAB[t + 1, ] <- m_M_strAB[t, ] %*% a_P_strAB[, , t]
  
  ## Fill in transition-dynamics array
  # For SoC
  a_A_SoC[, , t + 1]   <- diag(m_M_SoC[t, ])   %*% a_P_SoC[, , t]
  # For strategy AB
  a_A_strAB[, , t + 1] <- diag(m_M_strAB[t, ]) %*% a_P_strAB[, , t]
}

## Store the cohort traces in a list 
l_m_M <- list(SoC =  m_M_SoC,
              AB  =  m_M_strAB)
names(l_m_M) <- v_names_str

## Store the transition dynamics array for each strategy in a list 
l_a_A <- list(SoC =  a_A_SoC,
              AB  =  a_A_strAB)
names(l_a_A) <- v_names_str
```

# 06 Plot Outputs

## 06.1 Plot the cohort trace for strategy SoC 

```{r}
plot_trace(m_M_SoC)
```

# 07 State Rewards 

```{r}
## Scale by the cycle length 
# Vector of state utilities under strategy SoC
v_u_SoC    <- c(H  = u_H, 
                S1 = u_S1, 
                S2 = u_S2, 
                D  = u_D) * cycle_length
# Vector of state costs under strategy SoC
v_c_SoC    <- c(H  = c_H, 
                S1 = c_S1,
                S2 = c_S2, 
                D  = c_D) * cycle_length
# Vector of state utilities under strategy AB
v_u_strAB  <- c(H  = u_H, 
                S1 = u_trtAB, 
                S2 = u_S2, 
                D  = u_D) * cycle_length
# Vector of state costs under strategy AB
v_c_strAB  <- c(H  = c_H, 
                S1 = c_S1 + c_trtAB, 
                S2 = c_S2 + c_trtAB, 
                D  = c_D) * cycle_length

## Store state rewards 
# Store the vectors of state utilities for each strategy in a list 
l_u <- list(SoC = v_u_SoC,
            AB  = v_u_strAB)
# Store the vectors of state cost for each strategy in a list 
l_c <- list(SoC =  v_c_SoC,
            AB  =  v_c_strAB)

# assign strategy names to matching items in the lists
names(l_u) <- names(l_c) <- v_names_str
```

# 08 Compute expected outcomes 

```{r}
# Create empty vectors to store total utilities and costs 
v_tot_qaly <- v_tot_cost <- vector(mode = "numeric", length = n_str)
names(v_tot_qaly) <- names(v_tot_cost) <- v_names_str

## Loop through each strategy and calculate total utilities and costs 
for (i in 1:n_str) { # i <- 1
  v_u_str <- l_u[[i]]   # select the vector of state utilities for the i-th strategy
  v_c_str <- l_c[[i]]   # select the vector of state costs for the i-th strategy
  a_A_str <- l_a_A[[i]] # select the transition array for the i-th strategy
  
  ## Array of state rewards 
  # Create transition matrices of state utilities and state costs for the i-th strategy 
  m_u_str   <- matrix(v_u_str, nrow = n_states, ncol = n_states, byrow = T)
  m_c_str   <- matrix(v_c_str, nrow = n_states, ncol = n_states, byrow = T)
  # Expand the transition matrix of state utilities across cycles to form a transition array of state utilities
  a_R_u_str <- array(m_u_str, 
                     dim      = c(n_states, n_states, n_cycles + 1),
                     dimnames = list(v_names_states, v_names_states, 0:n_cycles))
  # Expand the transition matrix of state costs across cycles to form a transition array of state costs
  a_R_c_str <- array(m_c_str, 
                     dim      = c(n_states, n_states, n_cycles + 1),
                     dimnames = list(v_names_states, v_names_states, 0:n_cycles))
  
  ## Apply transition rewards
  # Apply disutility due to transition from H to S1
  a_R_u_str["H", "S1", ]      <- a_R_u_str["H", "S1", ]       - du_HS1
  # Add transition cost per cycle due to transition from H to S1
  a_R_c_str["H", "S1", ]      <- a_R_c_str["H", "S1", ]       + ic_HS1
  # Add transition cost  per cycle of dying from all non-dead states
  a_R_c_str[-n_states, "D", ] <- a_R_c_str[-n_states, "D", ] + ic_D
  
  ### Expected QALYs and costs for all transitions per cycle
  # QALYs = life years x QoL
  # Note: all parameters are annual in our example. In case your own case example is different make sure you correctly apply.
  a_Y_c_str <- a_A_str * a_R_c_str
  a_Y_u_str <- a_A_str * a_R_u_str 
  
  ### Expected QALYs and costs per cycle
  ## Vector of QALYs and costs
  v_qaly_str <- apply(a_Y_u_str, 3, sum) # sum the proportion of the cohort across transitions 
  v_cost_str <- apply(a_Y_c_str, 3, sum) # sum the proportion of the cohort across transitions
  
  ## Discounted total expected QALYs and Costs per strategy and apply within-cycle correction if applicable
  # QALYs
  v_tot_qaly[i] <- t(v_qaly_str) %*% (v_dwe * v_wcc)
  # Costs
  v_tot_cost[i] <- t(v_cost_str) %*% (v_dwc * v_wcc)
}
```

# 09 Cost-effectiveness analysis (CEA) 

```{r}
## Incremental cost-effectiveness ratios (ICERs) 
df_cea <- calculate_icers(cost       = v_tot_cost, 
                          effect     = v_tot_qaly,
                          strategies = v_names_str)
df_cea
```

```{r}
## CEA table in proper format 
table_cea <- format_table_cea(df_cea)
table_cea
```

```{r}
## CEA frontier 
plot(df_cea, label = "all", txtsize = 16) +
  expand_limits(x = max(table_cea$QALYs) + 0.1) +
  theme(legend.position = c(0.82, 0.3))
```

=======
---
title: 'Markov Sick-Sicker model in R'
subtitle: "with simulation-time dependency"
author: "The DARTH workgroup"
output:
  pdf_document: default
  html_document: default
keep_tex: yes
self_contained: no
---

Please cite our publications when using this code:

- Alarid-Escudero F, Krijkamp EM, Enns EA, Yang A, Hunink MGM Pechlivanoglou P,
Jalal H. A Tutorial on Time-Dependent Cohort State-Transition Models in R using 
a Cost-Effectiveness Analysis Example. Medical Decision Making, 2022 (In press). https://arxiv.org/abs/2108.13552

Authors: 
- Fernando Alarid-Escudero <fernando.alarid@cide.edu>
- Eline Krijkamp
- Eva A. Enns
- Alan Yang
- M.G. Myriam Hunink
- Petros Pechlivanoglou
- Hawre Jalal
Please cite the article when using this code

To program this tutorial we used:
R version 4.0.5 (2021-03-31)
Platform: 64-bit operating system, x64-based processor
Running under: Mac OS 12.2.1
RStudio: Version 1.4.1717 2009-2021 RStudio, Inc

This code implements a simulation-time-dependent Sick-Sicker cSTM model to  conduct a CEA of four strategies:
- Standard of Care (SoC): best available care for the patients with the disease. This scenario reflects the natural history of the disease  
  progression.
- Strategy A: treatment A is given to patients in the Sick and Sicker states, but only improves the quality of life of those in the Sick state.
- Strategy B: treatment B is given to all sick patients and reduces disease progression from the Sick to Sicker state.
- Strategy AB: This strategy combines treatment A and treatment B. The disease  progression is reduced, and individuals in the Sick state have an  
  improved quality of life.

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE)
```

Change `eval` to `TRUE` if you want to knit this document.

```{r}
rm(list = ls())      # clear memory (removes all the variables from the workspace)
```

# 01 Load packages

```{r, warning = F, message = F}
if (!require('pacman')) install.packages('pacman'); library(pacman) # use this package to conveniently install other packages
# load (install if required) packages from CRAN
p_load("dplyr", "tidyr", "reshape2", "devtools", "scales", "ellipse", "ggplot2", "ggrepel", "gridExtra", "lazyeval", "igraph", "truncnorm", "ggraph", "reshape2", "patchwork", "knitr", "stringr", "diagram", "dampack")                                               
# load (install if required) packages from GitHub
# install_github("DARTH-git/darthtools", force = TRUE) #Uncomment if there is a newer version
p_load_gh("DARTH-git/darthtools")
```

# 02 Load functions

```{r}
# all functions are in the darthtools package
```

# 03 Model input

```{r}
## General setup 
cycle_length <- 1   # cycle length equal to one year (use 1/12 for monthly)
n_age_init   <- 25  # age at baseline
n_age_max    <- 100 # maximum age of follow up
n_cycles     <- (n_age_max - n_age_init)/cycle_length # time horizon, number of cycles
# Age labels 
v_age_names  <- paste(rep(n_age_init:(n_age_max-1), each = 1/cycle_length), 
                      1:(1/cycle_length), 
                      sep = ".")
# the 4 health states of the model:
v_names_states <- c("H",  # Healthy (H)
                    "S1", # Sick (S1)
                    "S2", # Sicker (S2)
                    "D")  # Dead (D)
                                           
n_states <- length(v_names_states)   # number of health states 

### Discounting factors 
d_c <- 0.03 # annual discount rate for costs 
d_e <- 0.03 # annual discount rate for QALYs

### Strategies 
v_names_str <- c("Standard of care", # store the strategy names
                 "Strategy A", 
                 "Strategy B",
                 "Strategy AB") 
n_str       <- length(v_names_str)   # number of strategies

## Within-cycle correction (WCC) using Simpson's 1/3 rule 
v_wcc  <- gen_wcc(n_cycles = n_cycles, method = "Simpson1/3")

### Transition rates (annual), and hazard ratios (HRs) 
r_HS1  <- 0.15  # constant annual rate of becoming Sick when Healthy
r_S1H  <- 0.5   # constant annual rate of becoming Healthy when Sick
r_S1S2 <- 0.105 # constant annual rate of becoming Sicker when Sick
hr_S1  <- 3     # hazard ratio of death in Sick vs Healthy 
hr_S2  <- 10    # hazard ratio of death in Sicker vs Healthy 

### Effectiveness of treatment B 
hr_S1S2_trtB <- 0.6  # hazard ratio of becoming Sicker when Sick under treatment B

## Age-dependent mortality rates 
lt_usa_2015 <- read.csv("LifeTable_USA_Mx_2015.csv")
# Extract age-specific all-cause mortality for ages in model time horizon
v_r_mort_by_age <- lt_usa_2015 %>% 
  dplyr::filter(Age >= n_age_init & Age < n_age_max) %>%
  dplyr::select(Total) %>%
  as.matrix()

### State rewards 
#### Costs 
c_H    <- 2000  # annual cost of being Healthy
c_S1   <- 4000  # annual cost of being Sick
c_S2   <- 15000 # annual cost of being Sicker
c_D    <- 0     # annual cost of being dead
c_trtA <- 12000 # annual cost of receiving treatment A
c_trtB <- 13000 # annual cost of receiving treatment B
#### Utilities 
u_H    <- 1     # annual utility of being Healthy
u_S1   <- 0.75  # annual utility of being Sick
u_S2   <- 0.5   # annual utility of being Sicker
u_D    <- 0     # annual utility of being dead
u_trtA <- 0.95  # annual utility when receiving treatment A

### Transition rewards 
du_HS1 <- 0.01  # disutility when transitioning from Healthy to Sick
ic_HS1 <- 1000  # increase in cost when transitioning from Healthy to Sick
ic_D   <- 2000  # increase in cost when dying

### Discount weight for costs and effects 
v_dwc  <- 1 / ((1 + (d_e * cycle_length)) ^ (0:n_cycles))
v_dwe  <- 1 / ((1 + (d_c * cycle_length)) ^ (0:n_cycles))

# Process model inputs 
## Age-specific transition rates to the Dead state for all cycles 
v_r_HDage  <- rep(v_r_mort_by_age, each = 1/cycle_length)
# Name age-specific mortality vector 
names(v_r_HDage) <- v_age_names

# compute mortality rates
v_r_S1Dage  <- v_r_HDage * hr_S1 # Age-specific mortality rate in the Sick state 
v_r_S2Dage  <- v_r_HDage * hr_S2 # Age-specific mortality rate in the Sicker state 
# transform rates to probabilities adjusting by cycle length
p_HS1       <- rate_to_prob(r = r_HS1, t = cycle_length)  # constant annual probability of becoming Sick when Healthy conditional on surviving 
p_S1H       <- rate_to_prob(r = r_S1H, t = cycle_length)  # constant annual probability of becoming Healthy when Sick conditional on surviving
p_S1S2      <- rate_to_prob(r = r_S1S2, t = cycle_length) # constant annual probability of becoming Sicker when Sick conditional on surviving
v_p_HDage   <- rate_to_prob(v_r_HDage, t = cycle_length)  # Age-specific mortality risk in the Healthy state 
v_p_S1Dage  <- rate_to_prob(v_r_S1Dage, t = cycle_length) # Age-specific mortality risk in the Sick state
v_p_S2Dage  <- rate_to_prob(v_r_S2Dage, t = cycle_length) # Age-specific mortality risk in the Sicker state

## Annual transition probability of becoming Sicker when Sick for treatment B 
# Apply hazard ratio to rate to obtain transition rate of becoming Sicker when Sick for treatment B
r_S1S2_trtB <- r_S1S2 * hr_S1S2_trtB
# Transform rate to probability to become Sicker when Sick under treatment B 
# adjusting by cycle length conditional on surviving
p_S1S2_trtB <- rate_to_prob(r = r_S1S2_trtB, t = cycle_length)
```

# 04 Construct state-transition models

## 04.1 Initial state vector

```{r}
# All starting healthy
v_m_init <- c(H = 1, S1 = 0, S2 = 0, D = 0) # initial state vector
v_m_init
```

## 04.2 Initialize cohort traces

```{r}
### Initialize cohort trace under SoC 
m_M_SoC <- matrix(NA, 
              nrow = (n_cycles + 1), ncol = n_states, 
              dimnames = list(0:n_cycles, v_names_states))
# Store the initial state vector in the first row of the cohort trace
m_M_SoC[1, ] <- v_m_init

### Initialize cohort trace for strategies A, B, and AB 
# Structure and initial states are the same as for SoC
m_M_strA  <- m_M_SoC # Strategy A
m_M_strB  <- m_M_SoC # Strategy B
m_M_strAB <- m_M_SoC # Strategy AB
```

## 04.3 Create transition probability matrices

```{r}
## Create transition probability arrays for strategy SoC 
### Initialize transition probability array for strategy SoC 
# All transitions to a non-death state are assumed to be conditional on survival
a_P_SoC <- array(0,
                 dim  = c(n_states, n_states, n_cycles),
                 dimnames = list(v_names_states, 
                                 v_names_states, 
                                 0:(n_cycles - 1)))
### Fill in array
## From H
a_P_SoC["H", "H", ]   <- (1 - v_p_HDage) * (1 - p_HS1)
a_P_SoC["H", "S1", ]  <- (1 - v_p_HDage) *      p_HS1
a_P_SoC["H", "D", ]   <-      v_p_HDage
## From S1
a_P_SoC["S1", "H", ]  <- (1 - v_p_S1Dage) *       p_S1H
a_P_SoC["S1", "S1", ] <- (1 - v_p_S1Dage) * (1 - (p_S1H + p_S1S2))
a_P_SoC["S1", "S2", ] <- (1 - v_p_S1Dage) *               p_S1S2
a_P_SoC["S1", "D", ]  <-      v_p_S1Dage
## From S2
a_P_SoC["S2", "S2", ] <- 1 - v_p_S2Dage
a_P_SoC["S2", "D", ]  <-     v_p_S2Dage
## From D
a_P_SoC["D", "D", ]   <- 1

### Initialize transition probability array for strategy A as a copy of SoC's 
a_P_strA <- a_P_SoC

### Initialize transition probability array for strategy B 
a_P_strB <- a_P_SoC
# Update only transition probabilities from S1 involving p_S1S2
a_P_strB["S1", "S1", ] <- (1 - v_p_S1Dage) * (1 - (p_S1H + p_S1S2_trtB))
a_P_strB["S1", "S2", ] <- (1 - v_p_S1Dage) *               p_S1S2_trtB

### Initialize transition probability array for strategy AB as a copy of B's 
a_P_strAB <- a_P_strB

## Check if transition probability arrays are valid 
### Check that transition probabilities are [0, 1] 
check_transition_probability(a_P_SoC,   verbose = TRUE)
check_transition_probability(a_P_strA,  verbose = TRUE)
check_transition_probability(a_P_strB,  verbose = TRUE)
check_transition_probability(a_P_strAB, verbose = TRUE)
### Check that all rows for each slice of the array sum to 1 
check_sum_of_transition_array(a_P_SoC,   n_states = n_states, n_cycles = n_cycles, verbose = TRUE)
check_sum_of_transition_array(a_P_strA,  n_states = n_states, n_cycles = n_cycles, verbose = TRUE)
check_sum_of_transition_array(a_P_strB,  n_states = n_states, n_cycles = n_cycles, verbose = TRUE)
check_sum_of_transition_array(a_P_strAB, n_states = n_states, n_cycles = n_cycles, verbose = TRUE)
```

## 04.4 Create transition dynamics arrays

```{r}
# These arrays will capture transitions from each state to another over time 
### Initialize transition dynamics array for strategy SoC 
a_A_SoC <- array(0,
             dim      = c(n_states, n_states, n_cycles + 1),
             dimnames = list(v_names_states, v_names_states, 0:n_cycles))
# Set first slice of a_A_SoC with the initial state vector in its diagonal
diag(a_A_SoC[, , 1]) <- v_m_init
### Initialize transition-dynamics array for strategies A, B, and AB 
# Structure and initial states are the same as for SoC
a_A_strA  <- a_A_SoC
a_A_strB  <- a_A_SoC
a_A_strAB <- a_A_SoC
```

# 05 Run Markov model

```{r}
# Iterative solution of age-dependent cSTM
for(t in 1:n_cycles){
  ## Fill in cohort trace
  # For SoC
  m_M_SoC[t + 1, ]   <- m_M_SoC[t, ]   %*% a_P_SoC[, , t]
  # For strategy A
  m_M_strA[t + 1, ]  <- m_M_strA[t, ]  %*% a_P_strA[, , t]
  # For strategy B 
  m_M_strB[t + 1, ]  <- m_M_strB[t, ]  %*% a_P_strB[, , t]
  # For strategy ZB 
  m_M_strAB[t + 1, ] <- m_M_strAB[t, ] %*% a_P_strAB[, , t]
  
  ## Fill in transition-dynamics array
  # For SoC
  a_A_SoC[, , t + 1]   <- diag(m_M_SoC[t, ])   %*% a_P_SoC[, , t]
  # For strategy A
  a_A_strA[, , t + 1]  <- diag(m_M_strA[t, ])  %*% a_P_strA[, , t]
  # For strategy B
  a_A_strB[, , t + 1]  <- diag(m_M_strB[t, ])  %*% a_P_strB[, , t]
  # For strategy AB
  a_A_strAB[, , t + 1] <- diag(m_M_strAB[t, ]) %*% a_P_strAB[, , t]
}

## Store the cohort traces in a list 
l_m_M <- list(SoC =  m_M_SoC,
              A   =  m_M_strA,
              B   =  m_M_strB,
              AB  =  m_M_strAB)
names(l_m_M) <- v_names_str

## Store the transition dynamics array for each strategy in a list 
l_a_A <- list(SoC =  a_A_SoC,
              A   =  a_A_strA,
              B   =  a_A_strB,
              AB  =  a_A_strAB)
names(l_a_A) <- v_names_str
```

# 06 Plot Outputs

## 06.1 Plot the cohort trace for strategy SoC and all strategies

```{r}
## Plot the cohort trace for strategy SoC 
plot_trace(m_M_SoC)
```

```{r}
## Plot the cohort trace for all strategies 
plot_trace_strategy(l_m_M)
```

## 06.2 Plot the epidemiology outcomes

```{r}
### Survival 
survival_plot <- plot_surv(l_m_M, v_names_death_states = "D") +
  theme(legend.position = "bottom")
survival_plot
```

```{r}
### Prevalence 
prevalence_S1_plot   <- plot_prevalence(l_m_M, 
                                        v_names_sick_states = c("S1"), 
                                        v_names_dead_states = "D")  +
                        theme(legend.position = "")
prevalence_S1_plot
prevalence_S2_plot   <- plot_prevalence(l_m_M, 
                                        v_names_sick_states = c("S2"), 
                                        v_names_dead_states = "D")  +
                        theme(legend.position = "")
prevalence_S2_plot
prevalence_S1S2_plot <- plot_prevalence(l_m_M, 
                                        v_names_sick_states = c("S1", "S2"), 
                                        v_names_dead_states = "D") +
                        theme(legend.position = "")
prevalence_S1S2_plot
```

```{r}
# Proportion in sicker state
prop_sicker_plot     <- plot_proportion_sicker(l_m_M, 
                                               v_names_sick_states = c("S1", "S2"), 
                                               v_names_sicker_states = c("S2")) +
                        theme(legend.position = "bottom")
prop_sicker_plot
```

```{r}
## Combine plots 
gridExtra::grid.arrange(survival_plot, 
                        prevalence_S1_plot, 
                        prevalence_S2_plot, 
                        prevalence_S1S2_plot, 
                        prop_sicker_plot, 
                        ncol = 1, heights = c(0.75, 0.75, 0.75, 0.75, 1))
```

# 07 State Rewards 

```{r}
## Scale by the cycle length 
# Vector of state utilities under strategy SoC
v_u_SoC    <- c(H  = u_H, 
                S1 = u_S1, 
                S2 = u_S2, 
                D  = u_D) * cycle_length
# Vector of state costs under strategy SoC
v_c_SoC    <- c(H  = c_H, 
                S1 = c_S1,
                S2 = c_S2, 
                D  = c_D) * cycle_length
# Vector of state utilities under strategy A
v_u_strA   <- c(H  = u_H, 
                S1 = u_trtA, 
                S2 = u_S2, 
                D  = u_D) * cycle_length
# Vector of state costs under strategy A
v_c_strA   <- c(H  = c_H, 
                S1 = c_S1 + c_trtA,
                S2 = c_S2 + c_trtA, 
                D  = c_D) * cycle_length
# Vector of state utilities under strategy B
v_u_strB   <- c(H  = u_H, 
                S1 = u_S1, 
                S2 = u_S2, 
                D  = u_D) * cycle_length
# Vector of state costs under strategy B
v_c_strB   <- c(H  = c_H, 
                S1 = c_S1 + c_trtB, 
                S2 = c_S2 + c_trtB, 
                D  = c_D) * cycle_length
# Vector of state utilities under strategy AB
v_u_strAB  <- c(H  = u_H, 
                S1 = u_trtA, 
                S2 = u_S2, 
                D  = u_D) * cycle_length
# Vector of state costs under strategy AB
v_c_strAB  <- c(H  = c_H, 
                S1 = c_S1 + (c_trtA + c_trtB), 
                S2 = c_S2 + (c_trtA + c_trtB), 
                D  = c_D) * cycle_length

## Store state rewards 
# Store the vectors of state utilities for each strategy in a list 
l_u <- list(SoC = v_u_SoC,
            A   = v_u_strA,
            B   = v_u_strB,
            AB  = v_u_strAB)
# Store the vectors of state cost for each strategy in a list 
l_c <- list(SoC =  v_c_SoC,
            A   =  v_c_strA,
            B   =  v_c_strB,
            AB  =  v_c_strAB)

# assign strategy names to matching items in the lists
names(l_u) <- names(l_c) <- v_names_str
```

# 08 Compute expected outcomes 

```{r}
# Create empty vectors to store total utilities and costs 
v_tot_qaly <- v_tot_cost <- vector(mode = "numeric", length = n_str)
names(v_tot_qaly) <- names(v_tot_cost) <- v_names_str

## Loop through each strategy and calculate total utilities and costs 
for (i in 1:n_str) { # i <- 1
  v_u_str <- l_u[[i]]   # select the vector of state utilities for the i-th strategy
  v_c_str <- l_c[[i]]   # select the vector of state costs for the i-th strategy
  a_A_str <- l_a_A[[i]] # select the transition array for the i-th strategy
  
  ## Array of state rewards 
  # Create transition matrices of state utilities and state costs for the i-th strategy 
  m_u_str   <- matrix(v_u_str, nrow = n_states, ncol = n_states, byrow = T)
  m_c_str   <- matrix(v_c_str, nrow = n_states, ncol = n_states, byrow = T)
  # Expand the transition matrix of state utilities across cycles to form a transition array of state utilities
  a_R_u_str <- array(m_u_str, 
                     dim      = c(n_states, n_states, n_cycles + 1),
                     dimnames = list(v_names_states, v_names_states, 0:n_cycles))
  # Expand the transition matrix of state costs across cycles to form a transition array of state costs
  a_R_c_str <- array(m_c_str, 
                     dim      = c(n_states, n_states, n_cycles + 1),
                     dimnames = list(v_names_states, v_names_states, 0:n_cycles))
  
  ## Apply transition rewards
  # Apply disutility due to transition from H to S1
  a_R_u_str["H", "S1", ]      <- a_R_u_str["H", "S1", ]       - du_HS1
  # Add transition cost per cycle due to transition from H to S1
  a_R_c_str["H", "S1", ]      <- a_R_c_str["H", "S1", ]       + ic_HS1
  # Add transition cost  per cycle of dying from all non-dead states
  a_R_c_str[-n_states, "D", ] <- a_R_c_str[-n_states, "D", ] + ic_D
  
  ### Expected QALYs and costs for all transitions per cycle
  # QALYs = life years x QoL
  # Note: all parameters are annual in our example. In case your own case example is different make sure you correctly apply.
  a_Y_c_str <- a_A_str * a_R_c_str
  a_Y_u_str <- a_A_str * a_R_u_str 
  
  ### Expected QALYs and costs per cycle
  ## Vector of QALYs and costs
  v_qaly_str <- apply(a_Y_u_str, 3, sum) # sum the proportion of the cohort across transitions 
  v_cost_str <- apply(a_Y_c_str, 3, sum) # sum the proportion of the cohort across transitions
  
  ## Discounted total expected QALYs and Costs per strategy and apply within-cycle correction if applicable
  # QALYs
  v_tot_qaly[i] <- t(v_qaly_str) %*% (v_dwe * v_wcc)
  # Costs
  v_tot_cost[i] <- t(v_cost_str) %*% (v_dwc * v_wcc)
}
```

# 09 Cost-effectiveness analysis (CEA) 

```{r}
## Incremental cost-effectiveness ratios (ICERs) 
df_cea <- calculate_icers(cost       = v_tot_cost, 
                          effect     = v_tot_qaly,
                          strategies = v_names_str)
df_cea
```

```{r}
## CEA table in proper format 
table_cea <- format_table_cea(df_cea)
table_cea
```

```{r}
## CEA frontier -
plot(df_cea, label = "all", txtsize = 16) +
  expand_limits(x = max(table_cea$QALYs) + 0.1) +
  theme(legend.position = c(0.8, 0.2))
```
>>>>>>> da36a6af5b5df0f53272afa0e73089f8abde74fa
